---
title: "Paper-plots"
author: "Miriam Kretschmer & Vincent Fischer"
date: "28/02/2025"
output:
  bookdown::html_document2:
    code_folding: hide
bibliography: citations.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("/mnt/gapp/work/vincent/paper/Kretschmer2025/GitHub_submission"))
```

```{r}
suppressPackageStartupMessages({
library(AnnotationHub)
library(rtracklayer)
library(universalmotif)
library(MotifDb)
library(Rsamtools)
library(memes)
library(rGREAT)
library(MotifDb)
library(ggplot2)
library(ggrepel)
library(SummarizedExperiment)
library(SEtools)
library(sechm)
library(fgsea)
library(msigdbr)
library(RColorBrewer)
library(ggvenn)
library(viridis)
library(cowplot)
library(epiwraps)
  library(chromVAR)
  library(edgeR)
  library(limma)
  })
set.seed(123)
ah <- AnnotationHub()
```


```{r, eval = T}
# function to get non redundant motifs from MotifDB
getNonRedundantMotifs <- function(){
  motifs <- MotifDb::query(MotifDb::MotifDb, c("Mmusculus","HOCOMOCO"))
  modf <- data.frame(row.names=names(motifs),
                     TF=gsub("^Mmusculus-HOCOMOCOv10-|_MOUSE.+","",names(motifs)),
                     grade=gsub(".+\\.","",names(motifs)))
  modf <- modf[order(modf$TF,modf$grade),]
  modf <- modf[!duplicated(modf$TF),]
  setNames(universalmotif::convert_motifs(motifs[row.names(modf)]), modf$TF)
}
```


# Fig 1A - Motif enrichment sperm and oocyte

Sperm peaks
```{r}
d.sperm <- readRDS("objects/motif_enrichment_sperm_consensus_narrow.rds")
d.sperm <- readRDS("objects/motif_enrichment_sperm_consensus_broad.rds")

head(d.sperm)
length(d.sperm$motif_id)#281
```

MII oocyte peaks
```{r}
d.oo <- readRDS("objects/motif_enrichment_oocyte_broad.rds")

head(d.oo)
length(d.oo$motif_id)
d.oo$motif_id
```

Comparing the enrichments directly, e.g. by merging two data.frames and plotting the log(p) in one vs the other highlighting top 5 for each 
```{r}
m <- merge(d.oo, d.sperm, by="motif_id", suffixes=c(".oocyte",".sperm"))

m.oo.sort <- m[order(m[,2],decreasing = FALSE),]
m.oo5 <- m.oo.sort[1:3,]
m.sperm.sort <- m[order(m[,18],decreasing = FALSE),]
m.sperm5 <- m.sperm.sort[1:3,]
m
```

```{r}
ggplot(m, aes(x=log2((1+tp.sperm)/(1+fp.sperm)), y= log2((1+tp.oocyte)/(1+fp.oocyte)), size=(tp_percent.sperm))) + 
  geom_point() + 
  geom_text_repel(box.padding=0.5, min.segment.length = 0, max.overlaps = 10, hjust = "left", aes(label=motif_id), data=m.oo5, size=4) + 
  geom_text_repel(box.padding= 0.5, min.segment.length = 0, max.overlaps = 10, hjust = "right", aes(label=motif_id), data=m.sperm5)+
  geom_point(color="orange", data=m.oo5) +
  geom_point(color="blue", data=m.sperm5) +
  labs(color="Top enriched motifs")+
  labs(x="log2(fold-enrichment in spermatozoa)", y="log2(fold-enrichment in MII oocytes)", size="% of set") +
  #ylab("-log10 (adjusted p-value MII oocyte)")+
  #xlab("-log10 (adjusted p-value spermatozoa)")+
  theme_classic() +
  #theme(legend.position = "none")+
  ggtitle("Top ranked motifs in peaks of spermatozoa vs. MII oocytes")

  ggsave("./plots_svg/motif_enrichment_sperm_vs_oocyte_peaks.svg")
ggsave("./plots_pdf/motif_enrichment_sperm_vs_oocyte_peaks.pdf")#, width = 6, height = 4)
ggsave("./plots_png/motif_enrichment_sperm_vs_oocyte_peaks.png")
```


# Fig 1B - TF motif activity sperm vs oocyte

```{r}
library(TFBSTools)
sc <- readRDS("./objects/sc_sperm_oo.rds")
fa <- Rsamtools::FaFile("objects/Mus_musculus.GRCm39.dna_sm.primary_assembly.fa")

sc_gc <- addGCBias(sc, genome=fa)
motifs <- do.call(what = "PFMatrixList", lapply(getNonRedundantMotifs(), class="TFBSTools-PFMatrix", FUN= convert_motifs))
#motifs <- getNonRedundantMotifs()
motif_ix <- motifmatchr::matchMotifs(motifs, sc_gc, genome=fa)
motif_ix
dev <- computeDeviations(object=sc_gc, annotations=motif_ix)
dev$groups <- c("sperm","sperm","MIIoocyte")

fit <- lmFit(assays(dev)$z, model.matrix(~dev$groups),background_peaks = getBackgroundPeaks(sc_gc,rowData(sc_gc)$bias,niterations = 5000))

head(mores <- topTable(eBayes(fit),number=Inf),395)
sechm::sechm(dev, head(row.names(mores)), assayName = "z", top_annotation = "groups",do.scale=TRUE)
sechm::sechm(dev, head(row.names(mores),395), assayName = "z", top_annotation = "groups",do.scale = TRUE)
```

```{r}
ggplot(mores, aes(logFC, -log10(adj.P.Val))) + 
  geom_point() + 
  geom_hline(yintercept=-log10(0.05), linetype="dashed") +
  ggtitle("Transcription factor motif activity spermatozoa vs MIIoocyte") +
  theme_classic() +
  labs(y = "-log10(adjusted p-value)") +
  geom_text_repel(data = subset(mores, adj.P.Val < 0.01), 
                  aes(label = rownames(subset(mores, adj.P.Val < 0.01))), 
                  min.segment.length = 0,
                  size = 4)
#geom_point(aes(color = B)) +

ggsave("./plots_svg/TFmotif_activity_sperm_vs_oocyte_peaks.svg")
ggsave("./plots_pdf/TFmotif_activity_sperm_vs_oocyte_peaks.pdf")#, width = 6, height = 4)
ggsave("./plots_png/TFmotif_activity_sperm_vs_oocyte_peaks.png")
```




# Fig 1C - Human sperm 

```{r}
genome <- ah[["AH90808"]] #human 

GRE_motif <- MotifDb::MotifDb %>% 
  MotifDb::query("GRE") %>% 
  universalmotif::convert_motifs() %>% 
  .[[1]]
```

```{r GRE-peaks-hcp}
peaks <- import("extern_data/ATAC_Corces_1_peaks.narrowPeak")
peak_seqs <- get_sequence(peaks, genome)
fimo <- runFimo(peak_seqs, GRE_motif, meme_path = "/common/meme/bin") # you need to insert your indiviudual meme path here
GRE_peaks1 <- peaks[overlapsAny(peaks, fimo)]
```
```{r number-peaks-hcp, results='hide'}
length(peaks) #32027
length(fimo) #1338
sum(overlapsAny(peaks, fimo)) #716
length(GRE_peaks1) #716
```

```{r, eval=T}
tracks <- c(
  sperm1="extern_data/ATAC_Corces_1.full.bw",
  sperm2="extern_data/ATAC_Corces_2.full.bw"
)

m <- signal2Matrix(tracks, GRE_peaks1, w = 20, BPPARAM = 4)
m <- epiwraps:::.ese2ml(m)

fused_sperm_mat <- mergeSignalMatrices(m[1:2])
fused_sperm_mat

fused_m <- c(list("Human sperm"=fused_sperm_mat), m[-1:-2])


pdf("plots_pdf/humansperm.pdf", width=7.5, height=10)
h <- plotEnrichedHeatmaps(fused_m, row_title="Peaks from spermatozoa with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_trim = TRUE, mean_color = "#800080", heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL), use_raster = TRUE)
print(h)
dev.off()

png("plots_png/humansperm.png")
h <- plotEnrichedHeatmaps(fused_m, row_title="Peaks from spermatozoa with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080", heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL), use_raster = TRUE)
print(h)
dev.off()

svg("plots_svg/humansperm.svg", width=7.5, height=10)
h <- plotEnrichedHeatmaps(fused_m, row_title="Peaks from spermatozoa with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080",heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL), use_raster = TRUE)
print(h)
dev.off()
```


# Fig 1D - Sperm GRE peaks across spermatogenesis

```{r}
tracks <- c(
  uds1="extern_data/ATACse_PRJNA399533_Thy1posSpermatogonia_GSM2751121.bw",
  uds2="extern_data/ATACse_PRJNA399533_Thy1posSpermatogonia_GSM2751122.bw",
  ds1="extern_data/ATACse_PRJNA399533_cKitposSpermatogonia_GSM2751125.bw",
  ds2="extern_data/ATACse_PRJNA399533_cKitposSpermatogonia_GSM2751126.bw",
  ps1="extern_data/ATACse_PRJNA399533_PachytenSpermatocytes_GSM2751129.bw",
  ps2="extern_data/ATACse_PRJNA399533_PachytenSpermatocytes_GSM2751130.bw",
  rs1="extern_data/ATACse_PRJNA399533_RoundSpermatides_GSM2751133.bw",
  rs2="extern_data/ATACse_PRJNA399533_RoundSpermatides_GSM2751134.bw",
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data/ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  GVoocyte1="extern_data/ATAC_PRJNA480448_GVOocyte_GSM3262861.bw",
  GVoocyte2="extern_data/ATAC_PRJNA480448_GVOocyte_GSM3262862.bw",
  "MII oocyte"="extern_data/ATAC_PRJNA480448_MIIOocyte_GSM3262863.bw"
)
```

# create symbolic links

```{r}
peaks <- readRDS("objects/narrow_peaks_with_GRE_sperm.rds")
m <- signal2Matrix(tracks, peaks, w = 20, BPPARAM = 2)
m <- epiwraps:::.ese2ml(m)

fused_uds_mat <- mergeSignalMatrices(m[1:2])
fused_ds_mat <- mergeSignalMatrices(m[3:4])
fused_ps_mat <- mergeSignalMatrices(m[5:6])
fused_rs_mat <- mergeSignalMatrices(m[7:8])
fused_sperm_mat <- mergeSignalMatrices(m[9:12])
fused_gv_mat <- mergeSignalMatrices(m[13:14])

fused_m <- c(list(fused_uds=fused_uds_mat), list(fused_ds=fused_ds_mat), list(fused_ps=fused_ps_mat), list(fused_rs=fused_rs_mat), list(fused_sperm=fused_sperm_mat), list(fused_GVoocytes=fused_gv_mat), m[15], m[-1:-15])
fused_n <- renormalizeBorders(lapply(fused_m,log1p))

#fused_m <- readRDS("rds_saves/paper/spermatogenesis_fused_m.rds")
#fused_n <- readRDS("rds_saves/paper/spermatogenesis_fused_n.rds")
```

Combined in one plot
```{r}
pdf("plots_pdf/GREpeaks_from_sperm_across_spermatogenesis_normalized.pdf", width=17.5, height=10)
h <- plotEnrichedHeatmaps(fused_n, row_title="Peaks from spermatozoa with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()

png("plots_png/GREpeaks_from_sperm_across_spermatogenesis_normalized.png", height=768)
h <- plotEnrichedHeatmaps(fused_n, row_title="Peaks from spermatozoa with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()
```

Single panels of heatmaps 
```{r}
# Create a list of heatmaps
for (i in 1:length(fused_n)) {
  h <- plotEnrichedHeatmaps(fused_n[i], raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080",
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))
  
  # Save each heatmap to a PDF
  pdf(paste0("plots_pdf/GREpeaks_from_sperm_across_spermatogenesis_normalized_", i, ".pdf"), width=2.5, height=10)
  draw(h, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
}
```

```{r}
# Create a list of heatmaps
for (i in 1:length(fused_n)) {
  h <- plotEnrichedHeatmaps(fused_n[i], raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080",
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4, "npc"), title=NULL))

  # Save each heatmap to a PNG
  png(paste0("plots_png/GREpeaks_from_sperm_across_spermatogenesis_normalized_", i, ".png"), width=250, height=768)
  draw(h, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
}
```



# Fig 1E - Promotors with GRE; ATAC, H3k4m3 and 5mC


```{r}
tracks <- c(
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data/ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  sperm1_H3K4me3="extern_data/SRR3309386.bw",
  sperm2_H3K4me3="extern_data/SRR3309387.bw",
  sperm3_H3K4me3="extern_data/SRR3703039.bw",
  sperm4_H3K4me3="extern_data/SRR5049005.bw",
  sperm5_H3K4me3="extern_data/SRR3225849.bw",
  sperm1_H3K27ac="extern_data/SRR3225845.bw",
  sperm2_H3K27ac="extern_data/SRR5049001.bw",
  sperm1_H3K9ac="extern_data/SRR3225847.bw",
  sperm2_H3K9ac="extern_data/SRR5049003.bw",
  sperm1_H3K4me1="extern_data/SRR3225848.bw",
  sperm2_H3K4me1="extern_data/SRR5049004.bw",
  sperm1_H3.3="extern_data/ChIP_PRJNA480447_Sperm_GSM3262880.bw",  
  sperm2_H3.3="extern_data/ChIP_PRJNA480447_Sperm_GSM3262881.bw",
  sperm1_H2A.Z="extern_data/ChIP_PRJNA480447_Sperm_GSM3262882.bw",
  sperm2_H2A.Z="extern_data/ChIP_PRJNA480447_Sperm_GSM3262880.bw"
)

```






```{r}
methyl_bw <- list.files(path = "intern_data/Nanopore/",pattern = "39.bw",full.names = T)
names(methyl_bw) <- c("GR_Ctrl1","GR_Ctrl2","GR_KO1","GR_KO2")

methyl_bw_ctrl <- methyl_bw[1:2]

```
```{r}
methyl_bed <- sapply(names(methyl_bw),function(names){
  import.bw(methyl_bw[names])
  
  })
```


promoters regions with GRE 
```{r}
proms <- readRDS("objects/GRE_promoters_center.rds")

```

```{r}
sm2 <- signal2Matrix(w = 20L, filepaths = tracks[c(1:4,5:9)],regions = proms)
#plotEnrichedHeatmaps(sm2)
```


```{r}
sm_5mC <- signal2Matrix(w = 20L , smooth = T, limit = c(0,100), filepaths = methyl_bed, regions = proms)
sm2 <- sm2[rownames(sm2) %in% rownames(sm_5mC),]
```




```{r}
sm_merge <- mergeSignalMatrices(epiwraps:::.ese2ml(sm2)[1:4])
sm_merge_H3K4me3 <- mergeSignalMatrices(epiwraps:::.ese2ml(sm2)[5:9])
sm_merge_5mC <- mergeSignalMatrices(epiwraps:::.ese2ml(sm_5mC)[1:2])
```

```{r}
p <- plotEnrichedHeatmaps(ml = list(ATAC = sm_merge,H3K4me3 = sm_merge_H3K4me3),multiScale = F,row_title = "Promoters with GRE",color = rocket(2000),scale_title = "density",axis_name = c("-2kb","TSS","+2kb"), use_raster = T,mean_scale_side = "none",  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))+
                       plotEnrichedHeatmaps(ml = list("5mC" = sm_merge_5mC),row_title = "Promoters with GRE",use_raster =TRUE, scale_title = "5mc[%]",axis_name = c("-2kb","TSS","+2kb"),mean_scale_side = "none",  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
# p
 #draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
 
 #draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")

  png(paste0("plots_png/Fig1E.png"), width=750, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/Fig1E.pdf"), width=7.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig1E.svg"), width=7.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```
each heatmap on its own
```{r}
#ATAC
p <- plotEnrichedHeatmaps(ml = list(ATAC = sm_merge),multiScale = F,row_title = "Promoters with GRE",color = rocket(2000),scale_title = "density",axis_name = c("-2kb","TSS","+2kb"), use_raster = T,  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
png(paste0("plots_png//Fig1EATAC.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf//Fig1EATAC.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg//Fig1EATAC.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
# H3K4m3 
  p <- plotEnrichedHeatmaps(ml = list(H3K4me3 = sm_merge_H3K4me3),multiScale = F,row_title = "Promoters with GRE",color = rocket(2000),scale_title = "density",axis_name = c("-2kb","TSS","+2kb"), use_raster = T,  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
png(paste0("plots_png//Fig1EH3K4me3.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf//Fig1EH3K4me3.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig1EH3K4me3.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
# 5mC
  
p <-   plotEnrichedHeatmaps(ml = list("5mC" = sm_merge_5mC),row_title = "Promoters with GRE",use_raster =TRUE, scale_title = "5mc[%]",axis_name = c("-2kb","TSS","+2kb"),  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
png(paste0("plots_png/Fig1EH3K4me3.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/Fig1EH3K4me3.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig1EH3K4me3.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
```


# Fig 1F GR-ChIP
```{r}
bw_ChIP <- list.files("intern_data/ChIP/",pattern = "*subtract",full.names = T)
names(bw_ChIP)<- gsub(x = basename(bw_ChIP),pattern = ".subtract.bw",replacement = "")
bw_GR <- bw_ChIP[1]
names(bw_GR) <- "GR"
GR_ap_broad <- import("intern_data/ChIP/VF_sp_GR_ap_EKDL220011603-1A_H5J3HDSX5_L2_peaks.broadPeak",format = "broadPeak")
```

```{r}
bed_GR <- import.bw(bw_GR) 
sm_GR <-  signal2Matrix(filepaths = list(GR = bed_GR),regions = GR_ap_broad,w = 10L,smooth = T, background= 0) 
```


```{r,fig.width=7,fig.height=9}

p_GR1 <-  plotEnrichedHeatmaps(sm_GR,row_title = "GR-ChIP sites",scale_title = "GR\ndesnity",colors = mako(2000), raster_resize_mat = TRUE, use_raster = TRUE,axis_name = c("-2kb","0","+2kb"), trim = c(0.01,0.99),width = unit(0.8,"npc"), heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))


#p_GR1

#p_GR2 <-  plotEnrichedHeatmaps(sn_GR,row_title = "GR-ChIP sites",scale_title = "GR\ndesnity",colors = c("black","steelblue1"),mean_scale_side = "none", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"),trim = c(0.005,0.995), heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))

#p_GR2
  draw(p_GR1, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")

  png(paste0("plots_png/Fig1F.png"), width=250, height=768)
  draw(p_GR1, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/Fig1F.pdf"), width=2.5, height=10)
  draw(p_GR1, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig1F.svg"), width=2.5, height=10)
  draw(p_GR1, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```


# Fig 1G Promoter with GRE GR,ATAC,H3K4m3,H3k27ac
```{r}
tracks <- c(
  
  sperm1_H3K4me3="extern_data/SRR3309386.bw",
  sperm2_H3K4me3="extern_data/SRR3309387.bw",
  sperm3_H3K4me3="extern_data/SRR3703039.bw",
  sperm4_H3K4me3="extern_data/SRR5049005.bw",
  sperm5_H3K4me3="extern_data/SRR3225849.bw",
  sperm1_H3K27ac="extern_data/SRR3225845.bw",
  sperm2_H3K27ac="extern_data/SRR5049001.bw",
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data/ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  sperm1_GR = "intern_data/ChIP/VF_sp_GR_ap.subtract.bw"
)


proms <- readRDS("objects/GRE_promoters_center.rds")
set.seed(1234)
#proms <- sample(proms,10000)

m <- signal2Matrix(tracks, proms, w = 20, BPPARAM = 8)
m <- epiwraps:::.ese2ml(m)



fused_H3K4me3 <- mergeSignalMatrices(m[1:5])
fused_H3K27ac <- mergeSignalMatrices(m[6:7])
fused_ATAC <- mergeSignalMatrices(m[8:11])


fused_m <- c(list(GR =m[[12]]),list(H3K27ac=fused_H3K27ac), list(H3K4me3=fused_H3K4me3),list(ATAC=fused_ATAC))
fused_n <- renormalizeBorders(lapply(fused_m,log1p))


bg_nf <- getNormFactors(tracks,method = "background")


bg <- epiwraps::renormalizeSignalMatrices(ml = m ,scaleFactors = bg_nf)



fused_H3K4me3_bg <- mergeSignalMatrices(bg[1:5])
fused_H3K27ac_bg <- mergeSignalMatrices(bg[6:7])
fused_ATAC_bg <- mergeSignalMatrices(bg[8:11])
fused_bg <- c(list(GR =bg[[12]]),list(H3K27ac=fused_H3K27ac_bg), list(H3K4me3=fused_H3K4me3_bg), list(ATAC = fused_ATAC_bg))
 

```

```{r}
sperm1_GR <- "intern_data/ChIP/VF_sp_GR_ap.subtract.bw"
```

```{r}
sperm1_GR_bed <- import.bw(sperm1_GR)
sm_gr_smooth <- signal2Matrix(filepaths = list(GR = sperm1_GR_bed),regions = proms,w = 20L,smooth  = T,background = 0)
```


```{r}
plotEnrichedHeatmaps(sm_gr_smooth,row_title = "Promoter with GRE",scale_title = "GR\ndesnity",colors = mako(2000),mean_scale_side = "none", trim = c(0.02,0.98),
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))+
plotEnrichedHeatmaps(fused_bg[2:4],row_title = "Promoter with GRE",scale_title = "HPTM\ndensity",colors = rocket(2000),mean_scale_side = "none",
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL),scale_rows = FALSE)

```

```{r}
p <- plotEnrichedHeatmaps(sm_gr_smooth,row_title = "Promoter with GRE",scale_title = "GR\ndesnity",colors = mako(2000),mean_scale_side = "none", trim = c(0.02,0.98),axis_name = c("-2kb","TSS","+2kb"),
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))+
plotEnrichedHeatmaps(fused_bg[c(4,3,2)],row_title = "Promoter with GRE",scale_title = "HPTM\ndensity",colors = rocket(2000),mean_scale_side = "none",axis_name = c("-2kb","TSS","+2kb"),
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL),scale_rows = FALSE)

ro <- p@ht_list[[1]]@row_order
# p
 #draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
 
 #draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")

  png(paste0("plots_png//Fig1G.png"), width=1000, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf//Fig1G.pdf"), width=10, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg//Fig1G.svg"), width=10, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```
each heatmap on its own
```{r}
#GR
p <- plotEnrichedHeatmaps(sm_gr_smooth,row_title = "Promoter with GRE",scale_title = "GR\ndesnity",colors = mako(2000), trim = c(0.02,0.98),axis_name = c("-2kb","TSS","+2kb"),row_order = ro,
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))
png(paste0("plots_png//Fig1G_GR.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf//Fig1G_GR.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg//Fig1G_GR.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
# ATAC 
  p <-plotEnrichedHeatmaps(fused_bg[c(4)],row_title = "Promoter with GRE",scale_title = "HPTM\ndensity",colors = rocket(2000),axis_name = c("-2kb","TSS","+2kb"),row_order = ro,
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL),scale_rows = FALSE)
png(paste0("plots_png//Fig1G_ATAC.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf//Fig1G_ATAC.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg//Fig1G_ATAC.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
# H3K4me3
  
p <-  plotEnrichedHeatmaps(fused_bg[c(3)],row_title = "Promoter with GRE",scale_title = "HPTM\ndensity",colors = rocket(2000),axis_name = c("-2kb","TSS","+2kb"),row_order = ro,
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL),scale_rows = FALSE)
png(paste0("plots_png//Fig1G_H3K4me3.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf//Fig1G_H3K4me3.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg//Fig1G_H3K4me3.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  
# H3K27ac
  
p <-  plotEnrichedHeatmaps(fused_bg[c(2)],row_title = "Promoter with GRE",scale_title = "HPTM\ndensity",colors = rocket(2000),axis_name = c("-2kb","TSS","+2kb"),row_order = ro,
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL),scale_rows = FALSE)
png(paste0("plots_png/Fig1G_H3K27ac.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/Fig1G_H3K27ac.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig1G_H3K27ac.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")  
  dev.off()
  
```



#Fig S1A - Comparison sperm data of published data and of Fischer and Kretschmer 2023

```{r}
peaks <- readRDS("/mnt/gapp/work/miriam/EXP14_accessibility_sperm_embryo/rds_saves/peaks/consensus_peaks/consensus.peaks_sperm.rds")

tracks <- c(
  "Jung 1"="extern_data//ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  "Jung 2"="extern_data//ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  "Jung 3"="extern_data//ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  "Jung 4"="extern_data//ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  "Fischer 1"="extern_data/NoMx_N704_KG.bw",
  "Fischer 2"="extern_data/NoMx_N705_KG.bw"
)

m <- signal2Matrix(tracks, peaks, w = 20, BPPARAM = 4)
m <- epiwraps:::.ese2ml(m)
n <- renormalizeBorders(lapply(m,log1p))

pdf("plots_pdf/sperm_comparisons.pdf", width=17.5, height=10)
h <- plotEnrichedHeatmaps(n, row_title="Peaks from spermatozoa", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_trim = TRUE)
print(h)
dev.off()

png("plots_png/sperm_comparisons.png", height=768)
h <- plotEnrichedHeatmaps(n, row_title="Peaks from spermatozoa", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()
```



# Fig S1B A549 atac motif enrichment

sperm vs A549 enrichment

```{r}
d.sperm <- readRDS("objects/motif_enrichment_sperm_consensus_broad.rds")
d.a549 <- readRDS("objects/motif_enrichment_sperm_A549.rds")
m <- merge(d.a549, d.sperm, by="motif_id", suffixes=c(".a549",".sperm"))

m.a549.sort <- m[order(m[,2],decreasing = FALSE),]
m.a5495 <- m.a549.sort[1:3,]
m.sperm.sort <- m[order(m[,18],decreasing = FALSE),]
m.sperm5 <- m.sperm.sort[1:3,]
m

gr <- m.sperm.sort[2, ]
stat1 <- m.a549.sort[1, ]
```
```{r}
ggplot(m, aes(x=log2((1+tp.sperm)/(1+fp.sperm)), y= log2((1+tp.a549)/(1+fp.a549)), size=(tp_percent.sperm))) + 
  geom_point() + 
  #geom_point(aes(color = ifelse(abs(log2((1+tp.sperm)/(1+fp.sperm)) - log2((1+tp.a549)/(1+fp.a549))) <= 1, "grey", "black"))) + 
  #geom_text_repel(box.padding=0.5, min.segment.length = 0, max.overlaps = 10, hjust = "left", aes(label=motif_id), data=m.a5495, size=4) + 
  #geom_text_repel(box.padding= 0.5, min.segment.length = 0, max.overlaps = 10, hjust = "right", aes(label=motif_id), data=m.sperm5)+
  geom_point(data=gr, color="orange")+
  geom_text_repel(box.padding=0.5, min.segment.length = 0, max.overlaps = 10,  nudge_x = 0.3, direction = "x", hjust = "right", aes(label=motif_id ), data= gr)+
  geom_point(data=stat1, color="blue")+
  geom_text_repel(box.padding=0.5, min.segment.length = 0, max.overlaps = 10,  nudge_y = 0.3, direction = "y", hjust = "right", aes(label=motif_id ), data= stat1)+
  #geom_point(color="orange", data=m.a5495) +
  #geom_point(color="blue", data=m.sperm5) +
  labs(color="Top enriched motifs")+
  labs(x="log2(fold-enrichment in spermatozoa)", y="log2(fold-enrichment in A549)", size="% of set") +
  #ylab("-log10 (adjusted p-value A549)")+
  #xlab("-log10 (adjusted p-value spermatozoa)")+
  theme_classic() +
  #theme(legend.position = "none")+
  ggtitle("Top ranked motifs in peaks of spermatozoa vs. A549")

ggsave("./plots_svg/motif_enrichment_sperm_vs_A549_peaks.svg")
ggsave("./plots_pdf/motif_enrichment_sperm_vs_A549_peaks.pdf")#, width = 6, height = 4)
ggsave("./plots_png/motif_enrichment_sperm_vs_A549_peaks.png")

```

# Fig S1C - Sperm GRE peaks not accessible in immature germ cells across spermatogenesis 

```{r}
tracks <- c(
  uds1="extern_data/ATACse_PRJNA399533_Thy1posSpermatogonia_GSM2751121.bw",
  uds2="extern_data/ATACse_PRJNA399533_Thy1posSpermatogonia_GSM2751122.bw",
  ds1="extern_data/ATACse_PRJNA399533_cKitposSpermatogonia_GSM2751125.bw",
  ds2="extern_data/ATACse_PRJNA399533_cKitposSpermatogonia_GSM2751126.bw",
  ps1="extern_data/ATACse_PRJNA399533_PachytenSpermatocytes_GSM2751129.bw",
  ps2="extern_data/ATACse_PRJNA399533_PachytenSpermatocytes_GSM2751130.bw",
  rs1="extern_data/ATACse_PRJNA399533_RoundSpermatides_GSM2751133.bw",
  rs2="extern_data/ATACse_PRJNA399533_RoundSpermatides_GSM2751134.bw",
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data/ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  GVoocyte1="extern_data/ATAC_PRJNA480448_GVOocyte_GSM3262861.bw",
  GVoocyte2="extern_data/ATAC_PRJNA480448_GVOocyte_GSM3262862.bw",
  "MII oocyte"="extern_data/ATAC_PRJNA480448_MIIOocyte_GSM3262863.bw"
)
```

```{r}
peaks <- readRDS("./objects/narrow_overlaps_spermGRE_not_immatureGermCellsGRE.rds")
m <- signal2Matrix(tracks, peaks, w = 20, BPPARAM = 2)
m <- epiwraps:::.ese2ml(m)

fused_uds_mat <- mergeSignalMatrices(m[1:2])
fused_ds_mat <- mergeSignalMatrices(m[3:4])
fused_ps_mat <- mergeSignalMatrices(m[5:6])
fused_rs_mat <- mergeSignalMatrices(m[7:8])
fused_sperm_mat <- mergeSignalMatrices(m[9:12])
fused_gv_mat <- mergeSignalMatrices(m[13:14])

fused_m <- c(list(fused_uds=fused_uds_mat), list(fused_ds=fused_ds_mat), list(fused_ps=fused_ps_mat), list(fused_rs=fused_rs_mat), list(fused_sperm=fused_sperm_mat), list(fused_GVoocytes=fused_gv_mat), m[15], m[-1:-15])
fused_n <- renormalizeBorders(lapply(fused_m,log1p))

#fused_m <- readRDS("rds_saves/paper/spermatogenesis_notimmature_fused_m.rds")
#fused_n <- readRDS("rds_saves/paper/spermatogenesis_notimmature_fused_n.rds")
```

```{r}
pdf("plots_pdf/GREpeaks_from_sperm_notimmature_across_spermatogenesis.pdf", width=17.5, height=10)
h <- plotEnrichedHeatmaps(fused_m, row_title="Peaks from spermatozoa with GRE \n not accessible in immature germ cells", raster_resize_mat = FALSE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()
```

```{r}
png("plots_png/GREpeaks_from_sperm_notimmature_across_spermatogenesis.png", height=768)
h <- plotEnrichedHeatmaps(fused_m, row_title="Peaks from spermatozoa with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()
```
# Fig S1D - Sperm GRE peaks across spermatogenesis hPTMs

Combined plot normalized
```{r activePTM-input, results='hide', fig.height=5, fig.width=4.5, fig.cap="Accessibility and ChIP signal of male germ cells at sperm peak regions"}
tracks <- c(
  H3K4me3="extern_data//SRR5049005.bw",#sperm4
  H3K27ac="extern_data/SRR3225845.bw",#sperm1
  H3K9ac="extern_data/SRR3225847.bw" #sperm1
)

peaks <- readRDS("objects/narrow_overlaps_spermGRE_not_immatureGermCellsGRE.rds")
m <- signal2Matrix(tracks, peaks, w=20, BPPARAM = 4)
m <- epiwraps:::.ese2ml(m)

#norm_factor <- bwNormFactors(x = tracks)
#m <- rescaleSignalMatrices(m,scaleFactors = norm_factor)
n <- renormalizeBorders(lapply(m,log1p))

#m <- readRDS("rds_saves/paper/hPTMs_spermNotImmature_m.rds")
#n <- readRDS("rds_saves/paper/hPTMs_spermNotImmature_n.rds")
```



Single panels of heatmaps
```{r}
# Create a list of heatmaps
for (i in 1:length(n)) {
  h <- plotEnrichedHeatmaps(n[i], raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = magma(2000), mean_color = "#800080",
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))
  
  # Save each heatmap to a PDF
    pdf(paste0("plots_pdf/GREpeaks_from_sperm_notImmature_hPTMs_normalized_", i, ".pdf"), width=2.5, height=10)
  draw(h, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
}
```

```{r}
# Create a list of heatmaps
for (i in 1:length(n)) {
  h <- plotEnrichedHeatmaps(n[i], raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = magma(2000), mean_color = "#800080",
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4, "npc"), title=NULL))

  # Save each heatmap to a PNG
  png(paste0("plots_png/GREpeaks_from_sperm_notImmature_hPTMs_normalized_", i, ".png"), width=250, height=768)
  draw(h, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
}
```



# Fig  S1E - Promoters with GRE across sperm ATAC and hPTMs

```{r}

tracks <- c(
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data/ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  sperm1_H3K4me3="extern_data/SRR3309386.bw",
  sperm2_H3K4me3="extern_data/SRR3309387.bw",
  sperm3_H3K4me3="extern_data/SRR3703039.bw",
  sperm4_H3K4me3="extern_data/SRR5049005.bw",
  sperm5_H3K4me3="extern_data/SRR3225849.bw",
  sperm1_H3K27ac="extern_data/SRR3225845.bw",
  sperm2_H3K27ac="extern_data/SRR5049001.bw",
  sperm1_H3K9ac="extern_data/SRR3225847.bw",
  sperm2_H3K9ac="extern_data/SRR5049003.bw",
  sperm1_H3K4me1="extern_data/SRR3225848.bw",
  sperm2_H3K4me1="extern_data/SRR5049004.bw",
  sperm1_H3.3="extern_data/ChIP_PRJNA480447_Sperm_GSM3262880.bw",  
  sperm2_H3.3="extern_data/ChIP_PRJNA480447_Sperm_GSM3262881.bw",
  sperm1_H2A.Z="extern_data/ChIP_PRJNA480447_Sperm_GSM3262882.bw",
  sperm2_H2A.Z="extern_data/ChIP_PRJNA480447_Sperm_GSM3262880.bw"
)



proms <- readRDS("objects/GRE_promoters_center.rds")

m <- signal2Matrix(tracks, proms, w = 20, BPPARAM = 8)


fused_sperm <- mergeSignalMatrices(m[1:4])
fused_H3K4me3 <- mergeSignalMatrices(m[5:9])
fused_H3K27ac <- mergeSignalMatrices(m[10:11])
fused_H3K9ac <- mergeSignalMatrices(m[12:13])
fused_H3K4me1 <- mergeSignalMatrices(m[14:15])
fused_H3.3 <- mergeSignalMatrices(m[16:17])
fused_H2A.Z <- mergeSignalMatrices(m[18:19])

fused_m <- c(list(fused_sperm=fused_sperm), list(fused_H3K4me3=fused_H3K4me3), list(fused_H3K27ac=fused_H3K27ac), list(fused_H3K9ac=fused_H3K9ac), list(fused_H3K4me1=fused_H3K4me1), list(fused_H3.3=fused_H3.3), list(fused_H2A.Z=fused_H2A.Z), m[-1:-19])
fused_n <- renormalizeBorders(lapply(fused_m,log1p))

fused_m1 <- fused_m[1:3]
```


# create symbloic links
```{r}
# List of files
tracks <- gsub(x = tracks,pattern = "../../../",replacement = "/mnt/")

# Target directory for symbolic links
target_dir <- "/mnt/gapp/work/vincent/paper/Kretschmer2025/GitHub_submission/extern_data/"

# Create symbolic links
sapply(tracks, function(file) {
  target <- file.path(target_dir, basename(file))
  file.symlink(file, target)
})
```

```{r}
pdf("plots_pdf/GREpromoters.pdf", width=7, height=10)
h <- plotEnrichedHeatmaps(fused_m1, row_title="Promoter regions with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()

png("plots_png/GREpromoters.png", height=768)
h <- plotEnrichedHeatmaps(fused_m1, row_title="Promoter regions with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()
```


# Fig S1F - Heatmap for ICR in mm10
```{r}
ICR <- read.csv("objects/ICR_Weber_mm10.csv",header = T)
ICR$seqnames <- ICR$X.chrom
ICR <- GRanges(ICR)

bw_mm10 <- list.files(path = "intern_data/Nanopore/",pattern = "_5mC_4reads|_5mC_(.*_noSex)",full.names = T)
names(bw_mm10) <- c("Ctrl1","Ctrl2","KO1","KO2")
```

```{r}
Met_bed <- sapply(names(bw_mm10),function(names){
  import.bw(bw_mm10[names])
  
  })
```




```{r}
sm <- signal2Matrix(filepaths = Met_bed,regions = ICR,w =10,type = "scaled",smooth = T,limit = c(0,100))
plotEnrichedHeatmaps(sm)
as.data.frame(ICR)
set.seed(123)
cl <- epiwraps::clusterSignalMatrices(sm,k = 2)
cl2 <- as.character(cl)

cl2[cl2 == 1] <- ("Maternal")

cl2[cl2 == 2] <- ("Paternal")


cl2 <- as.factor(cl2)

```

```{r}
sm_merge_Ctrl <- mergeSignalMatrices(sm[,1:2])


```

```{r}
mycolors2 <- c("Maternal" = "firebrick","Paternal"= "steelblue")
p <- plotEnrichedHeatmaps(list("ICR" = sm_merge_Ctrl),row_split = cl2,scale_title = "5mC[%]",mean_color  = mycolors2, use_raster = T,heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28,"npc"), title = NULL))
```

```{r}
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")

  png(paste0("plots_png/FigS1F.png"), width=250, height=500)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/FigS1F.pdf"), width=3.5, height=7.5)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/FigS1F.svg"), width=3.5, height=7.5)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```

# Fig S1G - Display of methlyation score distribution
```{r}
set.seed(1234)
Met_bed_Ctrl <- Met_bed[1:2]
df <- rbind(as.data.frame(sample(Met_bed_Ctrl$Ctrl1$score,size = 100000)),y = as.data.frame(sample(Met_bed_Ctrl$Ctrl1$score,size = 100000))) 
df$score <- df$`sample(Met_bed_Ctrl$Ctrl1$score, size = 1e+05)`
df$Sample <- rep("S1",nrow(df))
df$Sample[((nrow(df)/2)+1):nrow(df)] <- "S2" 
```
```{r}
ggplot(data = df,aes(x = score,group = Sample)) +
  #geom_histogram(aes(y = ..density..), bins = 30, fill = NULL, alpha = 0.5) +
  geom_density(color = "orange2", size = 1,fill = "orange2",alpha = 0.5) +
  labs(title = "5mC Score Distribution in mature sperm", x = "5mC score [%]", y = "Density") +
  theme_bw()+theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
ggsave("plots_svg/FigS1G_5mC_score_distribution.svg")
ggsave("plots_pdf/FigS1G_5mC_score_distribution.pdf")
ggsave("plots_pdf/FigS1G_5mC_score_distribution.png")
```

# Fig S1H - Promoters with GRE at accesible site of ATAC H3k4me3 5mC

```{r}
proms <- readRDS("objects/GRE_promoters_center.rds")
```


```{r}
tracks <- c(
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data/ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  sperm1_H3K4me3="extern_data/SRR3309386.bw",
  sperm2_H3K4me3="extern_data/SRR3309387.bw",
  sperm3_H3K4me3="extern_data/SRR3703039.bw",
  sperm4_H3K4me3="extern_data/SRR5049005.bw",
  sperm5_H3K4me3="extern_data/SRR3225849.bw",
  sperm1_H3K27ac="extern_data/SRR3225845.bw",
  sperm2_H3K27ac="extern_data/SRR5049001.bw",
  sperm1_H3K9ac="extern_data/SRR3225847.bw",
  sperm2_H3K9ac="extern_data/SRR5049003.bw",
  sperm1_H3K4me1="extern_data/SRR3225848.bw",
  sperm2_H3K4me1="extern_data/SRR5049004.bw",
  sperm1_H3.3="extern_data/ChIP_PRJNA480447_Sperm_GSM3262880.bw",  
  sperm2_H3.3="extern_data/ChIP_PRJNA480447_Sperm_GSM3262881.bw",
  sperm1_H2A.Z="extern_data/ChIP_PRJNA480447_Sperm_GSM3262882.bw",
  sperm2_H2A.Z="extern_data/ChIP_PRJNA480447_Sperm_GSM3262880.bw"
)



```

```{r}
peaks <- readRDS("objects/narrow_peaks_with_GRE_sperm.rds")
```

```{r}
peaks2 <- peaks[overlapsAny(peaks,proms)]
seqnames(peaks2)
peaks2 <- keepStandardChromosomes(peaks2,pruning.mode = "coarse") 
peaks2 <- dropSeqlevels(peaks2,value = c("MT","Y","X"),pruning.mode = "coarse") 
```


```{r}
sm_5mC <- signal2Matrix(w = 20L , smooth = T, limit = c(0,100), filepaths = methyl_bed, regions = peaks2)
```

```{r}
plotEnrichedHeatmaps(sm_5mC)
```
```{r}
sm2 <- signal2Matrix(w = 20L, filepaths = tracks[c(1:4,5:9)],regions = peaks2)
plotEnrichedHeatmaps(sm2)
```
```{r}
plotEnrichedHeatmaps(c(epiwraps:::.ese2ml(sm2),epiwraps:::.ese2ml(sm_5mC)),multiScale = T,show_heatmap_legend = F)
```
```{r}
sm_merge <- mergeSignalMatrices(epiwraps:::.ese2ml(sm2)[1:4])
sm_merge_H3K4me3 <- mergeSignalMatrices(epiwraps:::.ese2ml(sm2)[5:9])
sm_merge_5mC <- mergeSignalMatrices(epiwraps:::.ese2ml(sm_5mC)[1:2])
```

```{r,fig.height=10}
plotEnrichedHeatmaps(ml = list(ATAC = sm_merge,H3K4me3 = sm_merge_H3K4me3,"5mC" = sm_merge_5mC),multiScale = T,row_title = "Accessible promoters with GRE")
```
```{r}
p <- plotEnrichedHeatmaps(ml = list(ATAC = sm_merge,H3K4me3 = sm_merge_H3K4me3),multiScale = T,row_title = "Accessible promoters with GRE",color = rocket(2000),scale_title = "density",axis_name = c("-2kb","TSS","+2kb"), use_raster = T,mean_scale_side = "none",  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))+
                       plotEnrichedHeatmaps(ml = list("5mC" = sm_merge_5mC),row_title = "Accessible promoters with GRE",use_raster =TRUE, scale_title = "5mc[%]",axis_name = c("-2kb","TSS","+2kb"),mean_scale_side = "none",  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL));p
# p
 #draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
 
 #draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")

  png(paste0("plots_png/FigS1H.png"), width=750, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/FigS1H.pdf"), width=7.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/FigS1H.svg"), width=7.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```
each heatmap on its own
```{r}
#ATAC
p <- plotEnrichedHeatmaps(ml = list(ATAC = sm_merge),multiScale = F,row_title = "Accessible promoters with GRE",color = rocket(2000),scale_title = "density",axis_name = c("-2kb","TSS","+2kb"), use_raster = T,  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
png(paste0("plots_png/FigS1HATAC.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/FigS1HATAC.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/FigS1HATAC.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
# H*K4m3 
  p <- plotEnrichedHeatmaps(ml = list(H3K4me3 = sm_merge_H3K4me3),multiScale = F,row_title = "Accessible promoters with GRE",color = rocket(2000),scale_title = "density",axis_name = c("-2kb","TSS","+2kb"), use_raster = T,  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
png(paste0("plots_png/FigS1HH3K4me3.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/FigS1HH3K4me3.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/FigS1HH3K4me3.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
# 5mC
  
p <-   plotEnrichedHeatmaps(ml = list("5mC" = sm_merge_5mC),row_title = "Accessible promoters with GRE",use_raster =TRUE, scale_title = "5mc[%]",axis_name = c("-2kb","TSS","+2kb"),  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
png(paste0("plots_png/FigS1H5mc.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/FigS1H5mc.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/FigS1H5mc.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
```

# Fig 2B - Validation of GR-KO

```{r}
methyl_bw2 <- list.files(path = "intern_data/Nanopore/",pattern = "_..bw$",full.names = T)
```



```{r}
ensdb_mm10 <- ah[["AH89211"]]
```


```{r}
plotSignalTracks(files = list( "PGC-GR-KO" = methyl_bw2[3:4], Control = methyl_bw2[1:2]),overlay.alpha = 75,region = "18:39420652-39434571",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb_mm10,extend = c(0,0), fontface.main = 1,cex.main = 1,cex.id=0.9,cex.title = 0.7, ylim = c(0,0.1),) # , rotation.title = 0
grid.text(paste0("Nr3c1"), x = 0.5, y = 0.13, gp = gpar(fontsize = 10, fontface = "bold")) 
```

```{r}
png(paste0("plots_png/Fig2B.png"), width=750, height=512)
plotSignalTracks(files = list( "PGC-GR-KO" = methyl_bw2[3:4], Control = methyl_bw2[1:2]),overlay.alpha = 75,region = "18:39420652-39434571",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb_mm10,extend = c(0,0), fontface.main = 1.4,cex.main = 1,cex.id=0.9,cex.title = 0.9, ylim = c(0,0.1),) # , rotation.title = 0
grid.text(paste0("Nr3c1"), x = 0.5, y = 0.15, gp = gpar(fontsize = 15, fontface = "bold")) 
dev.off()

pdf(paste0("plots_pdf/Fig2B.pdf"), width=8, height=7,colormodel = "cymk")
plotSignalTracks(files = list( "PGC-GR-KO" = methyl_bw2[3:4], Control = methyl_bw2[1:2]),overlay.alpha = 75,region = "18:39420652-39434571",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb_mm10,extend = c(0,0), fontface.main = 1.4,cex.main = 1,cex.id=0.9,cex.title = 0.9, ylim = c(0,0.1),) # , rotation.title = 0
grid.text(paste0("Nr3c1"), x = 0.5, y = 0.15, gp = gpar(fontsize = 15, fontface = "bold")) 
dev.off()


svg(paste0("plots_svg/Fig2B.svg"), width=8, height=7)
plotSignalTracks(files = list( "PGC-GR-KO" = methyl_bw2[3:4], Control = methyl_bw2[1:2]),overlay.alpha = 75,region = "18:39420652-39434571",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb_mm10,extend = c(0,0), fontface.main = 1.4,cex.main = 1,cex.id=0.9,cex.title = 0.9, ylim = c(0,0.1),) # , rotation.title = 0
grid.text(paste0("Nr3c1"), x = 0.5, y = 0.15, gp = gpar(fontsize = 15, fontface = "bold")) 
dev.off()

```


# Fig 2C DMRs PGC-GR-KO sperm heatmap
regions of interest
```{r}
HyperDMR <- readRDS("objects/hyper_mm39.rds") 
HypoDMR <- readRDS("objects/hypo_mm39.rds")

DMR <- unlist(GRangesList(HyperDMR,HypoDMR))


```




```{r}
i <- integer(length(DMR))
i[1:(length(HyperDMR))] <- "HyperDMR"
i[((length(HyperDMR))+1):length(DMR)] <- "HypoDMR"

```

tracks
```{r}
#NanoporeTracks
methyl_bw <- list.files(path = "intern_data/Nanopore//",pattern = "39.bw",full.names = T)
names(methyl_bw) <- c("Control 1","Control 2","PGC-GR-KO 1","PGC-GR-KO 2")
```



```{r}
methyl_bed <- sapply(names(methyl_bw),function(names){
  import.bw(methyl_bw[names])
  
  })
```




initial heatmap of DMRs smooth
```{r}
sm1 <- signal2Matrix(filepaths = methyl_bed,w = 20,regions = DMR,smooth = T,limit = c(0,100))
```


```{r,fig.width=10}
mycolors <- c("HyperDMR"="#ffa500ff", "HypoDMR"= "#c0c0ffff")
p <- plotEnrichedHeatmaps((sm1),row_split = i,scale_rows = F,mean_color = mycolors,mean_scale_side = "left", top_annotation = TRUE,scale_title = "5mC [%]",axis_name = c("-2kb","0","+2kb"),row_title = "DMRs")

p
```

```{r}
mycolors <- c("HyperDMR"="#ffa500ff", "HypoDMR"= "#c0c0ffff")
p <- plotEnrichedHeatmaps((sm1),row_split = i,scale_rows = F,mean_color = mycolors,mean_scale_side = "left", top_annotation = TRUE,scale_title = "5mC [%]",axis_name = c("-2kb","0","+2kb"),row_title = "DMRs", heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))

p
```


```{r}


  png(paste0("plots_png/Fig2C.png"), width=1000, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/Fig2C.pdf"), width=10, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig2C.svg"), width=10, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```



# Fig 2D DMRs overlapping with GRE in promoter heatmap

```{r}
ensdb <- ah[["AH89457"]] # GRCm39

promoters <- promoters(ensdb)
```
we use fimo to get GRE in all promoters
```{r}
genome <- Rsamtools::FaFile("objects/Mus_musculus.GRCm39.dna_sm.primary_assembly.fa")
#promoters <- keepStandardChromosomes(promoters,pruning.mode = "coarse")
#promoters <- dropSeqlevels(x = promoters,value = "MT",pruning.mode = "coarse")
#promoter_seq <- get_sequence(regions = keepStandardChromosomes(promoters,pruning.mode = "coarse"),genome = genome)
#GRE_promoters <- memes::runFimo(sequences = promoter_seq,motifs = motifs$GCR,meme_path = "/common/meme/bin/")
#GRE_promoters <- reduce(GRE_promoters)
#saveRDS(GRE_promoters,"reduced_GRE_promoters_mm39.rds")
GRE_promoters <- readRDS("objects/reduced_GRE_promoters_mm39.rds")
```



```{r}
DMR_GRE_prom <- DMR[overlapsAny(DMR,GRE_promoters)]

DMR_GRE_up <- HyperDMR[overlapsAny(HyperDMR,GRE_promoters)]

DMR_GRE_down <- HypoDMR[overlapsAny(HypoDMR,GRE_promoters)]


DMR_GRE_both <- unlist(GRangesList(DMR_GRE_up,DMR_GRE_down))
```



initial heatmap of DMRs smooth
```{r}

sm1 <- signal2Matrix(filepaths = methyl_bed,w = 20,regions = DMR_GRE_both,smooth = T,limit = c(0,100))
```



```{r}
i <- integer(length(DMR_GRE_both))
i[1:(length(DMR_GRE_up))] <- "HyperDMR"
i[((length(DMR_GRE_up))+1):length(DMR_GRE_both)] <- "HypoDMR"

```





```{r,fig.width=10}

mycolors <- c("HyperDMR"="#ffa500ff", "HypoDMR"= "#c0c0ffff")
p <- plotEnrichedHeatmaps((sm1),row_split = i,mean_color = mycolors,scale_rows = F, mean_scale_side = "left", top_annotation = TRUE,scale_title = "5mC [%]",row_title = "DMRs overlapping GRE-sites in Promoters",axis_name = c("-2kb","GRE","+2kb"))

p
```

```{r}
mycolors <- c("HyperDMR"="#ffa500ff", "HypoDMR"= "#c0c0ffff")
p <- plotEnrichedHeatmaps((sm1),row_split = i,mean_color = mycolors,scale_rows = F, mean_scale_side = "left", top_annotation = TRUE,scale_title = "5mC [%]",row_title = "DMRs overlapping GRE-sites in Promoters", axis_name = c("-2kb","GRE","+2kb") ,heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL),use_raster = T,trim = c(0.0001,0.9999))

p
```


```{r}


  png(paste0("plots_png/Fig2D.png"), width=1000, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/Fig2D.pdf"), width=10, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig2D.svg"), width=10, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```


# Fig 2E - DEG heatmap PGC-GR-KO embryos

```{r}
# Load the data
se <- readRDS("objects/SEs_for_SEV/VF_2cell_embyro_early_paternal_GR_KO(Alpl).SE.rds")

# Extract assays and row data
assays(se)
rowData(se)
dea <- rowData(se)[[2]]
dea

# Subset differentially expressed genes with FDR < 0.05
deaFDR <- subset(dea, dea$FDR < 0.05)
deaFDR

# Define treatments and colors
treatments <- c("ctrl", "Early_GR_KO") # Specify the order explicitly
colors <- c("#DEDFE4", "darkgrey") # Adjust colors as needed

# Ensure the number of colors matches the number of treatments
if (length(colors) != length(treatments)) {
  stop("Number of colors must match number of treatments")
}

# Create a named vector for color mapping
color_mapping <- setNames(colors, treatments)

# Reorder columns in the se object
se <- se[, order(factor(colData(se)$condition, levels = treatments))]

# Add color_mapping to the heatmap annotation
ha = HeatmapAnnotation(df = data.frame(condition = factor(colData(se)$condition, levels = treatments)),
                       col = list(condition = color_mapping))

# Define color palette for the heatmap
color_palette <- colorRampPalette(c("blue", "black", "orange"))(100)

pdf("plots_pdf/DEGs_VF-Alpl-KO_heatmap.pdf", height=4) #, width=14.5, height=10)
x <- sechm(se, rownames(deaFDR), assayName = "log2FC", do.scale = FALSE, hmcols = color_palette, top_annotation = ha, row_title = "Differentially expressed genes", gaps_at="condition", use_raster = TRUE)
print(x)
dev.off()

png("plots_png/DEGs_VF-Alpl-KO_PROTAC_heatmap.png", height=768)
x <- sechm(se, rownames(deaFDR), assayName = "log2FC", do.scale = FALSE, hmcols = color_palette, top_annotation = ha, row_title = "Differentially expressed genes", gaps_at="condition", use_raster = TRUE)
print(x)
dev.off()

svg("plots_svg/DEGs_VF-Alpl-KO_heatmap.svg") #, width=14.5, height=10)
x <- sechm(se, rownames(deaFDR), assayName = "log2FC", do.scale = FALSE, hmcols = color_palette, top_annotation = ha, row_title = "Differentially expressed genes", gaps_at="condition", use_raster = TRUE)
print(x)
dev.off()

sechm(se, rownames(deaFDR), assayName = "log2FC", do.scale = FALSE, hmcols = color_palette, top_annotation = ha, row_title = "Differentially expressed genes", gaps_at="condition", use_raster = TRUE) 
```


#Fig 2F - volcano DEGs PGC-GR-KO


```{r}
library(scales)
se <- readRDS("objects/VF_2cell_embyro_early_paternal_GR_KO(Alpl).SE.rds")

# Extract assays and row data
assays(se)
rowData(se)
dea <- rowData(se)[[2]]
dea$Gene <- dea$ID

# Subset differentially expressed genes with FDR < 0.05
deaFDR <- subset(dea, dea$FDR < 0.005)
deaFDR

xlim <- c(-7, 7)
ylim <- c(0,6)

# Function to generate volcano plots
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$FDR < 0.01
  df$extreme[df$logFC > xlim[2] | df$logFC < xlim[1] | -log10(df$FDR) > ylim[2] | -log10(df$FDR) < ylim[1]] <- "extreme"
  df$highlight[df$extreme == "extreme"] <- "extreme"
  top_genes <- df[order(df$FDR), ][1:25, ]
  ggplot(df, aes(x=logFC, y=-log10(FDR))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black", "extreme" = "firebrick3")) +
    geom_text_repel(data = subset(top_genes, highlight == TRUE | highlight == "extreme"), aes(label=Gene), size=3.5, min.segment.length = 0, max.overlaps = Inf, nudge_y = 0.1) +
    geom_hline(yintercept=-log10(0.05), linetype="dashed")+
    xlab("logFC") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    scale_x_continuous(limits = xlim, oob = squish) +
    scale_y_continuous(limits = ylim, oob = squish) +
    ggtitle(title)
}

# Apply the function to each dataframe in the list
generate_volcano_plot(dea, "Differentially expressed genes in PGC-GR-KO 2-cell embryos")
ggsave("plots_svg/Fig2F_DEGs_VF-KO_volcano.svg")
ggsave("plots_pdf/Fig2F_DEGs_VF-KO_volcano.pdf")
ggsave("plots_png/Fig2F_DEGs_VF-KO_volcano.png")
```



# Fig 2G 5mC at Mtus1

```{r}
ensdb <- ah[["AH89457"]]
```


```{r}
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "8: 41494253-41495035",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1,cex.main = 1,cex.id=0.9,cex.title = 0.7) # , rotation.title = 0
grid.text(paste0("DMR: Mtus1/Gm16192"), x = 0.5, y = 0.95, gp = gpar(fontsize = 10, fontface = "bold"))
```

```{r}
png(paste0("plots_png/Fig2G.png"), width=750, height=512)
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "8: 41494253-41495035",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Mtus1/Gm16192"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold")) 
dev.off()

pdf(paste0("plots_pdf/Fig2G.pdf"), width=8, height=7,colormodel = "cymk")
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "8: 41494253-41495035",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Mtus1/Gm16192"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold"))
dev.off()


svg(paste0("plots_svg/Fig2G.svg"), width=8, height=7)
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "8: 41494253-41495035",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Mtus1/Gm16192"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold"))
dev.off()

```



#Fig 2H 5mC at Promoter of embryo DEA

```{r}
dea <- readRDS("objects/DEA_e2c.rds")
annGR <- readRDS("objects/annGR.rds")
```



```{r}
set.seed(1234)
annGR <- readRDS("objects/annGR.rds")
annGR2 <-keepStandardChromosomes(annGR[annGR$gene_name %in% dea$ID,],pruning.mode = "coarse"); annGR <- dropSeqlevels(annGR,value = c("chrMT","X","Y"),pruning.mode = "coarse") 
DEA_up <- annGR[annGR$gene_name %in% dea$ID[(dea$FDR < 0.05) & (dea$logFC < 0)] ,]
DEA_up <- keepStandardChromosomes(DEA_up,pruning.mode = "coarse"); DEA_up <- dropSeqlevels(DEA_up,value = c("chrMT","X","Y"),pruning.mode = "coarse")
DEA_up <- reduce(DEA_up)



DEA_down <- annGR[annGR$gene_name %in% dea$ID[(dea$FDR < 0.05) & (dea$logFC > 0)] ,]
DEA_down <- keepStandardChromosomes(DEA_down,pruning.mode = "coarse"); DEA_down <- dropSeqlevels(DEA_down,value = c("chrMT","X","Y"),pruning.mode = "coarse")  
DEA_down <- reduce(DEA_down)
DEA_down <- sample(DEA_down,size = length(DEA_up))


DEA <- unlist(GRangesList(DEA_up,DEA_down))



# we remove random promoters which have now overlap with any methylData
annGR2 <- annGR2[overlapsAny(annGR2,unlist(GRangesList(methyl_bed)))]

randDEA <- sample(reduce(annGR2[!overlapsAny(annGR2,DEA),]),length(DEA_up))

DEA2 <- unlist(GRangesList(DEA_up,DEA_down,randDEA))

DEA <- resize(DEA,width = 2,fix = "start")
DEA2 <- resize(DEA2,width = 2,fix = "start")


# sanity check
#proms <- promoters(ensdb,upstream = 5,downstream = 5)

#overlapsAny(DEA2,proms)
```

```{r}
k <- integer(length(DEA2))
k[1:(length(DEA_up))] <- "PGC-GR-KO\nearlyEmbryo UP"
k[((length(DEA_up))+1):(length(DEA_up)+length(DEA_down))] <- "PGC-GR-KO\nearlyEmbryo DOWN"
k[(((length(DEA_up)+length(DEA_down)))+1):length(DEA2)] <- "randomProms"
```


```{r}
i <- integer(length(DEA))
i[1:(length(DEA_up))] <- "PGC-GR-KO\nearlyEmbryo UP"
i[((length(DEA_up))+1):length(DEA)] <- "PGC-GR-KO\nearlyEmbryo DOWN"

```

```{r}
sm1 <- signal2Matrix(filepaths = methyl_bed,w = 1,regions = DEA2,smooth = T,limit = c(0,100),extend = c(500,1500))

plotEnrichedHeatmaps(sm1)
```
```{r}
melt <- meltSignals(sm1,splitBy = k)
melt$Group <- melt$split
```
```{r}
theme_set(theme_bw())
ggplot(melt, aes(position, mean, colour=Group, group = Group, fill = Group)) +
  geom_vline(xintercept=0, linetype="dashed") +
  labs(x="TSS", y="mean 5mC [%]")+
  stat_summary(geom="ribbon", alpha=0.3, colour = NA ) + stat_summary(fun=median, geom="line", size = 0.6)+
  ggtitle("5mC levels of sperm at\npromoter regions of DEAs in the PGC-GR-KO embryo")+   
  scale_color_manual(values = c("#c0c0ffff","#ffa500ff","forestgreen")) +
  scale_fill_manual(values = c("#c0c0ffff","#ffa500ff","forestgreen")) +
  theme(text = element_text(size = 10), axis.title.x = element_text(hjust = 0.26))




#ggsave("/mnt/gapp/work/vincent/paper/Kretschmer2025/plots/paper/Fig2H_5mC_levels_at_embryo_prom.svg")
#ggsave("/mnt/gapp/work/vincent/paper/Kretschmer2025/plots/paper/Fig2H_5mC_levels_at_embryo_prom.pdf")
#ggsave("/mnt/gapp/work/vincent/paper/Kretschmer2025/plots/paper/Fig2H_5mC_levels_at_embryo_prom.png")
```
```{r}
theme_set(theme_bw())
ggplot(melt, aes(position, mean, colour=Group, group = Group, fill = Group)) +
  geom_vline(xintercept=0, linetype="dashed") +
  labs(x="TSS", y="mean 5mC [%]")+
  stat_summary(geom="ribbon", alpha=0.3, colour = NA ) + stat_summary_bin(fun=mean, geom="line", size = 0.6)+
  ggtitle("5mC levels of sperm at\npromoter regions of DEAs in the PGC-GR-KO embryo")+   
  scale_color_manual(values = c("#c0c0ffff","#ffa500ff","forestgreen")) +
  scale_fill_manual(values = c("#c0c0ffff","#ffa500ff","forestgreen")) +
  theme(text = element_text(size = 10), axis.title.x = element_text(hjust = 0.26))


```



```{r}
melt[melt$split == "randomProms",]
```

## AUC
here we use all genes e2c up and down


```{r}
set.seed(1234)
annGR <- readRDS("objects/annGR.rds")
annGR2 <-keepStandardChromosomes(annGR[annGR$gene_name %in% dea$ID,],pruning.mode = "coarse"); annGR <- dropSeqlevels(annGR,value = c("chrMT","X","Y"),pruning.mode = "coarse") 
DEA_up <- annGR[annGR$gene_name %in% dea$ID[(dea$FDR < 0.05) & (dea$logFC < 0)] ,]
DEA_up <- keepStandardChromosomes(DEA_up,pruning.mode = "coarse"); DEA_up <- dropSeqlevels(DEA_up,value = c("chrMT","X","Y"),pruning.mode = "coarse")
DEA_up <- reduce(DEA_up)



DEA_down <- annGR[annGR$gene_name %in% dea$ID[(dea$FDR < 0.05) & (dea$logFC > 0)] ,]
DEA_down <- keepStandardChromosomes(DEA_down,pruning.mode = "coarse"); DEA_down <- dropSeqlevels(DEA_down,value = c("chrMT","X","Y"),pruning.mode = "coarse")  
DEA_down <- reduce(DEA_down)
#DEA_down <- sample(DEA_down,size = length(DEA_up))


DEA <- unlist(GRangesList(DEA_up,DEA_down))



# we remove random promoters which have now overlap with any methylData
annGR2 <- annGR2[overlapsAny(annGR2,unlist(GRangesList(methyl_bed)))]

randDEA <- sample(reduce(annGR2[!overlapsAny(annGR2,DEA),]),length(DEA_up))

DEA2 <- unlist(GRangesList(DEA_up,DEA_down,randDEA))

DEA <- resize(DEA,width = 2,fix = "start")
DEA2 <- resize(DEA2,width = 2,fix = "start")


# sanity check
#proms <- promoters(ensdb,upstream = 5,downstream = 5)

#overlapsAny(DEA2,proms)
```

```{r}
k <- integer(length(DEA2))
k[1:(length(DEA_up))] <- "PGC-GR-KO\nearlyEmbryo UP"
k[((length(DEA_up))+1):(length(DEA_up)+length(DEA_down))] <- "PGC-GR-KO\nearlyEmbryo DOWN"
k[(((length(DEA_up)+length(DEA_down)))+1):length(DEA2)] <- "randomProms"
```


```{r}
i <- integer(length(DEA))
i[1:(length(DEA_up))] <- "PGC-GR-KO\nearlyEmbryo UP"
i[((length(DEA_up))+1):length(DEA)] <- "PGC-GR-KO\nearlyEmbryo DOWN"

```

```{r}
sm1 <- signal2Matrix(filepaths = methyl_bed,w = 1,regions = DEA,smooth = T,limit = c(0,100),extend = c(500,1500))
# for proper calculations we ensure to remove regions were smoothing failed
sm1[is.na(as.data.frame(assay(sm1)))]
rows_with_na <- apply(as.data.frame(assay(sm1)), 1, function(row) any(is.na(row)))
sm1_clean <- sm1[!rows_with_na,]
i_clean <- i[!rows_with_na]
```


```{r}
plotEnrichedHeatmaps(sm1_clean)
assay(sm1_clean)
sm1_clean_merged <- mergeSignalMatrices(sm1_clean)  
df <- as.data.frame(sm1_clean_merged)
df <- df[,501:ncol(df)]
```
```{r}
AUC_per_region <- sapply(1:nrow(df),function(row){
  sm_auc(x =1:length(unlist(c(df[row,]))) , y =unlist(c(df[row,])))
})


mean(AUC_per_region[i_clean == "PGC-GR-KO\nearlyEmbryo DOWN"])/mean(AUC_per_region[i_clean == "PGC-GR-KO\nearlyEmbryo UP"])
median(AUC_per_region[i_clean == "PGC-GR-KO\nearlyEmbryo DOWN"])/median(AUC_per_region[i_clean == "PGC-GR-KO\nearlyEmbryo UP"])
```
```{r}
library(coin)
coin_wilcox <- coin::wilcox_test(AUC_per_region ~as.factor(i_clean),distribution=approximate(nresample = 10000))
  pvalue_wilcox <- pvalue(coin_wilcox)
  sigma_wilcox <- statistic(coin_wilcox)  
```


```{r}
melt <- meltSignals(sm1_clean,splitBy = i_clean)
melt$Group <- melt$split

melt_UP <- melt[melt$split == "PGC-GR-KO\nearlyEmbryo UP",]
melt_DOWN <- melt[melt$split == "PGC-GR-KO\nearlyEmbryo DOWN",]
```

```{r}
mean_data_up <- melt_UP %>%
  group_by(position) %>%
  summarize(mean_y = mean(mean))


mean_data_down <- melt_DOWN %>%
  group_by(position) %>%
  summarize(mean_y = mean(mean))


x_min <- 0
x_max <- 1500



filtered_data_up <- mean_data_up %>%
  filter(position >= x_min & position <= x_max)
filtered_data_down <- mean_data_down %>%
  filter(position >= x_min & position <= x_max)

```

```{r}
library(smplot2)
up_auc <- sm_auc(filtered_data_up$position,filtered_data_up$mean_y)

down_auc <- sm_auc(filtered_data_down$position,filtered_data_down$mean_y)

down_auc/up_auc
```


# Fig S2C - DMRs with GR track
```{r}
HyperDMR <- readRDS("objects/hyper_mm39.rds") 
HypoDMR <- readRDS("objects/hypo_mm39.rds")

DMR <- unlist(GRangesList(HyperDMR,HypoDMR))


```
```{r}
bw_ChIP <- list.files("intern_data/ChIP/",pattern = "*subtract",full.names = T)
names(bw_ChIP)<- gsub(x = basename(bw_ChIP),pattern = ".subtract.bw",replacement = "")
bw_GR <- bw_ChIP[1]
names(bw_GR) <- "GR"
bed_GR <- import.bw(bw_GR) 
```



```{r}

  sm_GR_5fC <- signal2Matrix(regions = DMR,filepaths = list(GR = bed_GR),type = "scaled")  
```

```{r}
plotEnrichedHeatmaps(sm_GR_5fC,row_title = "DMR",scale_title = "GRdensity", color = mako(2000))
```

# Fig S2E - DEAs PGC-GR-KO early embryo RegionUpset
```{r}
df <- read.csv("objects/DEAFDR005-VF-KO.csv", sep = ";")
# Create a list of vectors, each vector represents a category
sperm_overlaps <- df$Gene[df$Overlap.sperm.ATAC == 1]
gre <- df$Gene[df$GRE.in.gene == 1]
motifs <- df$Gene[df$gene.body.motifs.connected.to.GR == 1]

gene_list <- list(DEGs = df$Gene, "DEGs accesible in Sperm" = sperm_overlaps, "DEGs with GRE motif" = gre, "DEGs with GR binding partner motif" = motifs) 
```

```{r}
png(paste0("plots_png/FigS2E.png"),width=8, height=7,res = 300,units = "in")
regionUpset2(gene_list)
dev.off()

pdf(paste0("plots_pdf/FigS2E.pdf"), width=8, height=7,colormodel = "cymk")
regionUpset2(gene_list)
dev.off()


svg(paste0("plots_svg/FigS2E.svg"), width=8, height=7)
regionUpset2(gene_list)
dev.off()

```

# Fig S2F - GSEA PGC-GR-KO

```{r}
deaGO <- function(dea, threshold=0.05, direction=NULL, ontology="BP"){
  isDE <- dea$adj.P.Val<threshold
  if(!is.null(direction)) isDE <- isDE & sign(dea$logFC)==sign(direction)
  isDE <- setNames(factor(as.integer(isDE)),row.names(dea))
  print(table(isDE))
  god <- new("topGOdata", ontology=ontology, allGenes=isDE, 
             nodeSize=10, annot=annFUN.org, 
             ID="symbol", mapping="org.Mm.eg.db")
  ea <- runTest(god, algorithm="classic", statistic="fisher")
  res <- GenTable(god, classicFisher=ea, topNodes=length(god@graph@nodes))
  res$Term2 <- plgINS::breakStrings(res$Term)
  res$log2Enrichment <- log2(res$Significant/res$Expected)
  res$FDR <- p.adjust(res$classicFisher)
  res
}

deaGSEA <- function(dea, gsets=NULL){
  if(is.null(gsets))
    gsets <- plgINS::getMsigSets("Mus musculus", collections=c("H","C5"))
  signature <- setNames(sign(dea$logFC)*-log10(dea$P.Value), row.names(dea))
  fg <- fgseaMultilevel(gsets, signature, minSize=10, nproc=3)
  fg$genes <- sapply(fg$leadingEdge, collapse=", ", FUN=paste)
  fg$leadingEdge <- NULL
  as.data.frame(fg[order(fg$pval),])
}
```

```{r}
se <- readRDS("objects/VF_2cell_embyro_early_paternal_GR_KO(Alpl).SE.rds")
deas <- getDEA(se)
dea <- deas$glmFit
dea$P.Value <- dea$PValue

dea2 <- dea %>% distinct(ID, .keep_all = TRUE)
rownames(dea2) <- dea2$ID

res2 <- deaGSEA(dea2)

ggplot(head(res2,10), aes(x = reorder(pathway, NES), y = NES, fill = -log10(padj))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "grey", high = "orange") +
  theme_minimal() +
  labs(
    title = "Enriched gene sets in \n PGC-GR-KO derived embryos",
    x = "Gene set",
    y = "Enrichment Score",
    fill = "-log10(FDR)"
  )

ggsave("./plots_svg/GSEA_KO.svg")
ggsave("./plots_pdf/GSEA_KO.pdf")
ggsave("./plots_png/GSEA_KO.png")
```

# Fig S2G - Gene expression of Nr3c1 in early embryo og PGC-GR-KO sperm

```{r}
se <- readRDS("objects/VF_2cell_embyro_early_paternal_GR_KO(Alpl).SE.rds")
se$condition <- as.character(se$condition)
se$condition[se$condition == "Early_GR_KO"] <- "PGC-GR-KO"
se$condition[se$condition == "Ctrl"] <- "Control"
se$condition <- as.factor(se$condition)
```

```{r}
df <- as.data.frame(cbind(t(assays(se)$logcpm),colData(se))) # we extract data from the SE object and combine it with the colData as a dataframe, this makes it more easy to be accessed by ggplot

colnames(df) <- gsub(x = colnames(df),pattern = "^EN(.*)\\.",replacement = "")

df <- df[,!duplicated(colnames(df))]
```

```{r}
p <- ggplot((df), aes(x = condition, y = Nr3c1, color = condition)) +
  geom_violin(trim = TRUE) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "Gene expression levels of Nr3c1",
    y = "logcpm",
    x =""
  )
ggsave("plots_svg/FigS2G_Violing_Nr3c1.svg")
ggsave("plots_pdf/FigS2G_Violing_Nr3c1.pdf")
ggsave("plots_png/FigS2G_Violing_Nr3c1.png")
```



# Fig S2E - Tracks for Rdx and Ptp4a2

```{r}
png(paste0("plots_png/Fig2G_Rdx.png"), width=750, height=512)
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "9:52011472-52012171",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Rdx"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold")) 
dev.off()

pdf(paste0("plots_pdf/Fig2G_Rdx.pdf"), width=8, height=7,colormodel = "cymk")
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "9:52011472-52012171",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Rdx"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold"))
dev.off()


svg(paste0("plots_svg/Fig2G_Rdx.svg"), width=8, height=7)
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "9:52011472-52012171",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Rdx"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold"))
dev.off()

```


```{r}
png(paste0("plots_png/Fig2G_Ptp4a2.png"), width=750, height=512)
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "4: 129711699-129711894",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Ptp4a2"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold")) 
dev.off()

pdf(paste0("plots_pdf/Fig2G_Ptp4a2.pdf"), width=8, height=7,colormodel = "cymk")
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "4: 129711699-129711894",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Ptp4a2"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold"))
dev.off()


svg(paste0("plots_svg/Fig2G_Ptp4a2.svg"), width=8, height=7)
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "4: 129711699-129711894",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Ptp4a2"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold"))
dev.off()

```
# Fig 3A

```{r}
peak <- import("objects/NFpeaks/embryo_consensus.bed")
length(peak)

peak_seqs <- memes::get_sequence(peak, genome)
length(peak_seqs)

ame <- runAme(peak_seqs, database=convert_motifs(motifs), meme_path="/common/meme/bin") #https://meme-suite.org/meme/doc/ame-output-format.html #this step doesnt work??

ame[1:10]
```

```{r}
ame <- readRDS("objects/ame_embryo_consensus.rds")
ame[1:10]
gr <- ame[82, , drop=TRUE]
```


```{r}
ggplot(ame, aes(log2((1+tp)/(1+fp)), -log10(adj.pvalue))) + 
  geom_point(data=gr, aes(size=tp_percent), color="orange")+
  geom_text_repel(box.padding=0.5, min.segment.length = 0, max.overlaps = 10,  nudge_y = 0.3, direction = "y", hjust = "right", aes(label=motif_id ), data= gr)+
  geom_point(aes(size=tp_percent), alpha=0.3) + 
  geom_text_repel(box.padding=0.5, min.segment.length = 0, max.overlaps = 10,  nudge_y = 0.3, direction = "y", hjust = "right", aes(label=motif_id ), data= head(ame,10))+
  labs(x="log2(fold-enrichment)", size="% of set")+
  ylab("-log10 (adjusted p-value)")+
  xlab("Enrichment")+
  theme_classic() +
  ggtitle("Top enriched motifs in the early embryo")+
  guides(fill = guide_legend(override.aes = aes(color = NA)))

ggsave("./plots_svg/motif_enrichment_embryo_highconfidencepeaks.svg")
ggsave("./plots_pdf/motif_enrichment_embryo_highconfidencepeaks.pdf")
ggsave("./plots_png/motif_enrichment_embryo_highconfidencepeaks.png")
```




# Fig 3B - Clustering GRE peaks

```{r}
tracks <- c(
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data//ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data//ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  MIIoocyte1="extern_data/ATAC_PRJNA480448_MIIOocyte_GSM3262863.bw",
  embryoearly2cell="extern_data/ATAC_PRJNA277362_Early2cell_merged.bw",
  embryo2cell="extern_data/ATAC_PRJNA277362_2cell_merged.bw",
  embryo4cell="extern_data/ATAC_PRJNA277362_4cell_merged.bw",
  ESC="extern_data/ATAC_PRJNA277362_ESC_GSM1625858.bw"
  )
```


# create symbloic links
```{r}
# List of files
tracks <- gsub(x = tracks,pattern = "../../../",replacement = "/mnt/")

# Target directory for symbolic links
target_dir <- "/mnt/gapp/work/vincent/paper/Kretschmer2025/GitHub_submission/extern_data/"

# Create symbolic links
sapply(tracks, function(file) {
  target <- file.path(target_dir, basename(file))
  file.symlink(file, target)
})
```


```{r}
mycolors <- c("1"=brewer.pal(n=20,"Set1")[1], "2"=brewer.pal(n=20,"Set1")[2], "3"=brewer.pal(n=20,"Set1")[3], "4"=brewer.pal(n=20,"Set1")[4], "5"=brewer.pal(n=20,"Set1")[5], "6"=brewer.pal(n=20,"Set1")[6], "7"=brewer.pal(n=20,"Set1")[7], "8"=brewer.pal(n=20,"Set1")[8],
              "9"=brewer.pal(n=20,"Paired")[1], "10"=brewer.pal(n=20,"Paired")[2], "11"=brewer.pal(n=20,"Paired")[3], "12"=brewer.pal(n=20,"Paired")[4], "13"=brewer.pal(n=20,"Paired")[5], "14"=brewer.pal(n=20,"Paired")[6], "15"=brewer.pal(n=20,"Paired")[7], "16"=brewer.pal(n=20,"Paired")[8], "17"=brewer.pal(n=20,"Paired")[9], "18"=brewer.pal(n=20,"Paired")[10], "19"=brewer.pal(n=20,"Paired")[11])

set.seed(123)

peaks_emb <- readRDS("./objects/peaks_with_GRE_embryo_inAny.rds") 
peaks_sperm <- readRDS("./objects/narrow_peaks_with_GRE_sperm.rds")
peaks_oo <- readRDS("./objects/narrow_peaks_with_GRE_mIIoocytes.rds")

length(peaks_emb) #2084
length(peaks_sperm) #9311
length(peaks_oo) #792

peaks_emb <- reduce(peaks_emb, with.revmap=TRUE)
length(peaks_emb)
peaks_sperm <- reduce(peaks_sperm, with.revmap=TRUE)
length(peaks_sperm)
peaks_oo <- reduce(peaks_oo, with.revmap=TRUE)
length(peaks_oo)

peaklist <- unlist(GRangesList(peaks_emb, peaks_sperm, peaks_oo))
length(peaklist) #12117

m <- signal2Matrix(tracks, peaklist, w=5, type="center",  BPPARAM = 4)
m <- epiwraps:::.ese2ml(m)

fused_sperm_mat <- mergeSignalMatrices(m[1:4])
fused_m <- c(list(sperm=fused_sperm_mat), m[5:9], m[-1:-9])
fused_n <- renormalizeBorders(lapply(fused_m,log1p))

fused_m <- readRDS("objects/clustering_fused_m.rds")
fused_n <- readRDS("objects/clustering_fused_n.rds")

cl <- clusterSignalMatrices(fused_n, k=9) #% variance explained by clusters
ro <- order(rowMeans(fused_n$embryoearly2cell), decreasing = TRUE)
```

```{r}
pdf("plots_pdf/GREpeaks_clustering.pdf", width=14.5, height=10)
x <- plotEnrichedHeatmaps(fused_n, 
                     row_split = cl,
                     row_order = ro,
                     #trim = 0.99, 
                     raster_resize_mat = FALSE, 
                     mean_color = mycolors, 
                     scale_rows = FALSE,
                     axis_name = c("-2kb","0","+2kb")
                     )
print(x)
dev.off()



png("plots_pdf/GREpeaks_clustering.png", height=768)
x <- plotEnrichedHeatmaps(fused_n, 
                     row_split = cl,
                     row_order = ro,
                     #trim = 0.99, 
                     raster_resize_mat = FALSE, 
                     mean_color = mycolors, 
                     scale_rows = FALSE,
                     axis_name = c("-2kb","0","+2kb")
                     )
print(x)
dev.off()
```



# Fig 3C - 2-cell SmartSeq

```{r}
se <- readRDS("objects/MK_EXP20.1_ProtacEmbryo_merged.SE.rds")

# Generate the dataframes
dex <- rowData(se)[[4]]
protac <- rowData(se)[[3]]
pd <- rowData(se)[[5]]

protac$Gene <- rownames(protac)
dex$Gene <- rownames(dex)
pd$Gene <- rownames(pd)

# List of dataframes
df_list <- list(Dex = dex, "KH-103" = protac, "KH-103+Dex" = pd)

# Function to generate volcano plots
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$FDR < 0.1
  top_genes <- df[order(df$FDR), ][1:10, ]
  ggplot(df, aes(x=logFC, y=-log10(FDR))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black")) +
    geom_text_repel(data = subset(top_genes, highlight == TRUE), aes(label=Gene), size=3.5, min.segment.length = 0, max.overlaps = Inf, nudge_y = 0.1) +
    geom_hline(yintercept=-log10(0.05), linetype="dashed")+
    xlab("logFC") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    ggtitle(title)
}

# Apply the function to each dataframe in the list
plots <- lapply(names(df_list), function(name) {
  generate_volcano_plot(df_list[[name]], paste("Differentially expressed genes in 2-cell embryos upon ", name, " treatment"))
})

print(plots)

print(plots[1])
  ggsave("plots_svg//DEGs_EXP20-1_Dex_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-1_Dex_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-1_Dex_volcano.png")

print(plots[2])
ggsave("plots_svg/DEGs_EXP20-1_PROTAC_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-1_PROTAC_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-1_PROTAC_volcano.png")

print(plots[3])
ggsave("plots_svg/DEGs_EXP20-1_PD_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-1_PD_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-1_PD_volcano.png")
```

Heatmaps
```{r}
se <- readRDS("objects/MK_EXP20.1_ProtacEmbryo_merged.SE.rds")
se <- se[, se$treatment != "Dex"]
se <- se[, se$treatment != "PD"]

# Generate the dataframes
protac <- rowData(se)[[3]]
protac$Gene <- rownames(protac)
protacFDR <- subset(protac, protac$FDR <0.1)

treatments <- unique(se$treatment)

# Define own colors
colors <- c("grey", "orange" )#"800080", "grey", "orange", "black")  

# Make sure the number of colors matches the number of treatments
if (length(colors) != length(treatments)) {
  stop("Number of colors must match number of treatments")
}

# Create a named vector for color mapping
color_mapping <- setNames(colors, treatments)

# Add color_mapping to the heatmap annotation
ha = HeatmapAnnotation(df = data.frame(treatment = factor(colData(se)$treatment, levels = names(color_mapping))),
                       col = list(treatment = color_mapping))

sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes in 2-cell embryos", gaps_at="treatment")

pdf("plots_pdf/DEGs_EXP20-1_PROTAC_heatmap.pdf") #, width=14.5, height=10)
x <- sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes in 2-cell embryos", gaps_at="treatment")
print(x)
dev.off()

png("plots_png/DEGs_EXP20-1_PROTAC_heatmap.png", height=768)
x <- sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes in 2-cell embryos", gaps_at="treatment")
print(x)
dev.off()
#needs title: KH-103 effects on gene regulation in 2-cell embryos
```


# Fig 3D - 8-cell SmartSeq

```{r}
se <- readRDS("objects/MK_EXP20.2_ProtacEmbryo8cells.SE.rds")

# Generate the dataframes
dex <- rowData(se)[[2]]
protac <- rowData(se)[[3]]
pd <- rowData(se)[[4]]

protac$Gene <- rownames(protac)
dex$Gene <- rownames(dex)
pd$Gene <- rownames(pd)

# List of dataframes
df_list <- list(Dex = dex, "KH-103" = protac, "KH-103+Dex" = pd)

# Function for volcano plots
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$adj.P.Val < 0.1
  top_genes <- df[order(df$adj.P.Val), ][1:10, ]
  ggplot(df, aes(x=logFC, y=-log10(adj.P.Val))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black")) +
    geom_text_repel(data = subset(top_genes, highlight == TRUE), aes(label=Gene), size=3.5, min.segment.length = 0, max.overlaps = Inf, nudge_y = 1) +
    geom_hline(yintercept=-log10(0.05), linetype="dashed")+
    xlab("logFC") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    ggtitle(title)
}


# Apply the function to each dataframe in the list
plots <- lapply(names(df_list), function(name) {
  generate_volcano_plot(df_list[[name]], paste("Differentially expressed genes in 8-cell embryos upon ", name, " treatment"))
})
print(plots[1])
ggsave("plots_svg/DEGs_EXP20-2_Dex_volcano.svg")
  ggsave("plots_pdf/DEGs_EXP20-2_Dex_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-2_Dex_volcano.png")

print(plots[2])
ggsave("plots_svg/DEGs_EXP20-2_PROTAC_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-2_PROTAC_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-2_PROTAC_volcano.png")

print(plots[3])
ggsave("plots_svg/DEGs_EXP20-2_PD_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-2_PD_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-2_PD_volcano.png")
```

heatmaps
```{r}
se <- readRDS("objects/MK_EXP20.2_ProtacEmbryo8cells.SE.rds")
se <- se[, se$treatment != "DEXR"]
se <- se[, se$treatment != "DEX"]
se <- se[, se$treatment != "PDR"]
se <- se[, se$treatment != "PD"]

# Generate the dataframes
protac <- rowData(se)[[3]]
protac$Gene <- rownames(protac)
protacFDR <- subset(protac, protac$adj.P.Val <0.1)


treatments <- unique(se$treatment)

# Define own colors
colors <- c("#DEDFE4", "darkgrey") # "#919297", "#393A3F",)# , "#800080", "black", "yellow", "orange")#232429"  

# Make sure the number of colors matches the number of treatments
if (length(colors) != length(treatments)) {
  stop("Number of colors must match number of treatments")
}

# Create a named vector for color mapping
color_mapping <- setNames(colors, treatments)

# Add color_mapping to the heatmap annotation
ha = HeatmapAnnotation(df = data.frame(treatment = factor(colData(se)$treatment, levels = names(color_mapping))),
                       col = list(treatment = color_mapping))

color_palette <- colorRampPalette(c("blue", "black", "orange"))(100)

pdf("plots_pdf/DEGs_EXP20-2_PROTAC_heatmap.pdf") #, width=14.5, height=10)
x <- sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
print(x)
dev.off()

png("plots_png/DEGs_EXP20-2_PROTAC_heatmap.png", height=768)
x <- sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
print(x)
dev.off()
#needs title : KH-103 effects on gene regulation in 8-cell embryos

svg("plots_svg/DEGs_EXP20-2_PROTAC_heatmap.svg")
x <- sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
print(x)
dev.off()

sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
```


# Fig 3E - correlation of EXP20.2 target DEGs in 20.1 and Fig S3D - bland-altmann-plot correlation

```{r}
se <- readRDS("objects/MK_EXP20.1_ProtacEmbryo_merged.SE.rds")

# Generate the dataframes
dex <- rowData(se)[[4]]
protac <- rowData(se)[[3]]
pd <- rowData(se)[[5]]
dex$Gene <- rownames(dex)
protac$Gene <- rownames(protac)
pd$Gene <- rownames(pd)

# Generate dataframes with FDR<0.05 used in this analysis
dexFDR <- subset(dex, dex$FDR <0.05)
protacFDR <- subset(protac, protac$FDR <0.05)
pdFDR <- subset(pd, pd$FDR <0.05)
dexFDR$Gene <- rownames(dexFDR)
protacFDR$Gene <- rownames(protacFDR)
pdFDR$Gene <- rownames(pdFDR)



# Create a list of dataframes and add column Gene to it
df_list <- list(dexFDR = dexFDR, protacFDR = protacFDR, pdFDR = pdFDR)

rEXP20_2 <- readRDS("objects/df_list_FDR01.rds")
rEXP20_2 <- lapply(rEXP20_2, function(df) {
  df$Gene <- rownames(df)
  return(df)
})

# Function to automate the process
process_dataframes <- function(df_name, df_list, rEXP20_2) {
  # Get the dataframe from the list
  df <- df_list[[df_name]]
  
  # Construct the variable names
  FDR_var <- paste0(df_name, "FDR")
  
  # Filter the df dataframe to include only the genes that are in rEXP20_2[[FDR_var]]
  df_filtered <- df[df$Gene %in% rEXP20_2[[FDR_var]]$Gene, ]
  
  # Filter rEXP20_2[[FDR_var]] to include only the genes that are in df_filtered
  rEXP20_2_FDR_filtered <- rEXP20_2[[FDR_var]][rEXP20_2[[FDR_var]]$Gene %in% df_filtered$Gene, ]
  
  # Merge the two dataframes on the 'Gene' column
  merged <- merge(df_filtered, rEXP20_2_FDR_filtered, by = "Gene")
  
  # Calculate the correlation of the logFC between the two dataframes
  correlation <- cor.test(merged$logFC.x, merged$logFC.y)
  
  # Print the correlation
  print(correlation)
  
  correlation <- cor(merged$logFC.x, merged$logFC.y)
  
  # Create a scatter plot
  plot <- ggplot(merged, aes(x=logFC.x, y=logFC.y)) +
    geom_point() +
    geom_text_repel(aes(label=Gene)) +  # Add the gene names to the plot
    geom_smooth(method=lm, se=FALSE, color="orange") +
    annotate("text", x = Inf, y = -Inf, label = paste("Correlation:", round(correlation, 4)), vjust = -1.5, hjust = 1.2) +
    xlab(paste("logFC in 2-cell embryos")) +
    ylab(paste("logFC in 8-cell embryos")) +
    theme_classic() +
    ggtitle(paste("Correlation of gene expression changes between 2- and 8- cell embryos"))
  
  print(plot)
  ggsave("plots_svg/correlation_EXP20-2_EXP20-1_PROTAC.svg")
  ggsave("plots_pdf/correlation_EXP20-2_EXP20-1_PROTAC.pdf")
  ggsave("plots_png/correlation_EXP20-2_EXP20-1_PROTAC.png")
  
  
  # Calculate means and differences
  merged$mean <- rowMeans(merged[,c("logFC.x", "logFC.y")])
  merged$difference <- merged$logFC.x - merged$logFC.y
  
  # Create a Bland-Altman plot
  bland_altman_plot <- ggplot(merged, aes(x=mean, y=difference)) +
    geom_point() +
    geom_text_repel(aes(label=Gene)) +
    geom_smooth(method=lm, se=FALSE, color="orange") +
    xlab(paste("Mean of logFC in 2-cell and 8-cell embryos")) +
    ylab(paste("Difference of logFC in 2-cell and 8-cell embryos")) +
    theme_classic() +
    ggtitle("Correlation of gene expression changes between 2- and 8-cell embryos")
  
  print(bland_altman_plot)
  ggsave("plots_svg/correlationBAP_EXP20-2_EXP20-1_PROTAC.svg")
  ggsave("plots_pdf/correlationBAP_EXP20-2_EXP20-1_PROTAC.pdf")
  ggsave("plots_png/correlationBAP_EXP20-2_EXP20-1_PROTAC.png")
  
  return(merged)
}

# Can call this function with the dataframe names and the list of dataframes
df_names <- c("protac") #, "protac", "pd")
df_list <- list(protac = protac) #, protac = protac, pd = pd)

# Apply the function to each dataframe
lapply(df_names, process_dataframes, df_list = df_list, rEXP20_2 = rEXP20_2)
```

# Fig 3F - regionUpset DEGs in 2cell and 8cell


```{r}
library(UpSetR)

regionUpset2 <- function (x = NULL, genelist = NULL, returnList = FALSE, ...) 
{
    if (!is.null(genelist)) {
        x <- genelist
    }
    if (is.character(x)) 
        x <- as.list(x)
    stopifnot(length(x) > 1)
    if (is.list(x)) {
        if (is.null(names(x)) && all(unlist(lapply(x, is.character))) && 
            all(lengths(x) == 1)) 
            names(x) <- unlist(x)
        if (any(unlist(lapply(x, FUN = function(x) {
            if (is.character(x)) {
                if (length(x) != 0) return(TRUE)
                if (!file.exists(x)) return(TRUE)
            }
            FALSE
        })))) {
            stop("`x` should be a named list of character vectors.")
        }
    }
    x <- lapply(x, function(x) as.character(x))
    if (returnList) 
        return(x)
    UpSetR::upset(UpSetR::fromList(x), ...)
}
```

```{r}
df <- read.csv("objects/DEAFDR01-EXP20-1-2-PROTAC.csv", sep = ";")
# Create a list of vectors, each vector represents a category
sperm_overlaps <- df$Gene[df$Overlap.sperm.ATAC == 1]
gre <- df$Gene[df$GRE.in.gene == 1]
motifs <- df$Gene[df$gene.body.motifs.connected.to.GR == 1]

gene_list <- list(DEGs = df$Gene, "DEGs accesible in Sperm" = sperm_overlaps, "DEGs with GRE motif" = gre, "DEGs with GR binding partner motif" = motifs) 
```

```{r}
regionUpset2(gene_list)
```


```{r}
png(paste0("plots_png/Fig3F.png"), width=8, height=7,res = 300,units = "in")
regionUpset2(gene_list)
dev.off()

pdf(paste0("plots_pdf/Fig3F.pdf"), width=8, height=7,colormodel = "cymk")
regionUpset2(gene_list)
dev.off()


svg(paste0("plots_svg/Fig3F.svg"), width=8, height=7)
regionUpset2(gene_list)
dev.off()

```

# Fig S3A - GO term analysis embryo peaks with GRE

```{r}
embryo_consensus_peaks <- readRDS("./objects/peaks_with_GRE_embryo_consensus.rds")
seqlevelsStyle(embryo_consensus_peaks) <- "UCSC"
```

Liftover and pruning as GOterm analysis with submitGreatJob can only use mm10 and standard chromosomes

```{bash, eval = FALSE}
rsync -avzP rsync://hgdownload.soe.ucsc.edu/goldenPath/mm39/liftOver/mm39ToMm10.over.chain.gz -d ../../../tools/liftOver/
gzip -d ../../../tools/liftOver/mm39ToMm10.over.chain.gz
```

```{r}
mm39_mm10_chain <- import.chain("objects/mm39ToMm10.over.chain")

embryo_hc_peaks_mm10 <- liftOver(embryo_consensus_peaks, chain = mm39_mm10_chain)
embryo_hc_peaks_mm10 <- unlist(embryo_hc_peaks_mm10)
embryo_hc_peaks_mm10 <- keepStandardChromosomes(embryo_hc_peaks_mm10, pruning.mode = "coarse")
```

```{r}
length(embryo_consensus_peaks) #110
length(embryo_hc_peaks_mm10) #108
```

```{r, fig.width=10}
job <- submitGreatJob(embryo_hc_peaks_mm10, species="mm10") 
res <- getEnrichmentTables(job)

bp <- res$`GO Biological Process`
head(bp)
write.table(bp, file = "./objects//GO-analysis_biolprocess_hcembryoGRE.csv")

ggplot(head(bp,20), aes(Hyper_Fold_Enrichment, reorder(name, Hyper_Adjp_BH),  color=-log10(Hyper_Adjp_BH))) + geom_point() + scale_color_viridis_c() + ggtitle("Biological processes in embryo peaks with GRE")

ggsave("./plots_png//GO-term_enrichment_BiologicalProcess_embryo_GREhighconfidencepeaks.png")
```

# Fig S3B - sperm GRE peaks across embryo tracks

```{r, eval=FALSE}
tracks_embryos <- c(
  embryoearly2cell="extern_data/ATAC_PRJNA277362_Early2cell_merged.bw",
  embryo2cell="extern_data/tracks/ATAC_PRJNA277362_2cell_merged.bw",
  embryo4cell="extern_data/ATAC_PRJNA277362_4cell_merged.bw",
  ESC="extern_data/ATAC_PRJNA277362_ESC_GSM1625858.bw")

peaks <- readRDS("./rds_saves/peaks/GRE_peaks/peaks_with_GRE_sperm.rds")
m <- signal2Matrix(tracks_embryos, peaks, w=20, BPPARAM = 4)
m <- epiwraps:::.ese2ml(m)
n <- renormalizeBorders(lapply(m, log1p))


pdf("plots_pdf/GREpeaks-sperm-embryotracks.pdf")
x <- plotEnrichedHeatmaps(n, row_title="Peaks from sperm with GRE", trim = 0.99, raster_resize_mat = TRUE)
print(x)
dev.off()


png("plots_png/GREpeaks-sperm-embryotracks.png", width=1024, height=768)
x <- plotEnrichedHeatmaps(n, row_title="Peaks from sperm with GRE", trim = 0.99, raster_resize_mat = TRUE)
print(x)
dev.off()

print(x)
```


# create symbloic links
```{r}
# List of files
tracks_embryos <- gsub(x = tracks_embryos,pattern = "../../../",replacement = "/mnt/gapp/")

# Target directory for symbolic links
target_dir <- "/mnt/gapp/work/vincent/paper/Kretschmer2025/GitHub_submission/extern_data/"

# Create symbolic links
sapply(tracks_embryos, function(file) {
  target <- file.path(target_dir, basename(file))
  file.symlink(file, target)
})
```

# Fig S3C - GSEA of DEGs 

```{r}
deaGO <- function(dea, threshold=0.05, direction=NULL, ontology="BP"){
  isDE <- dea$adj.P.Val<threshold
  if(!is.null(direction)) isDE <- isDE & sign(dea$logFC)==sign(direction)
  isDE <- setNames(factor(as.integer(isDE)),row.names(dea))
  print(table(isDE))
  god <- new("topGOdata", ontology=ontology, allGenes=isDE, 
             nodeSize=10, annot=annFUN.org, 
             ID="symbol", mapping="org.Mm.eg.db")
  ea <- runTest(god, algorithm="classic", statistic="fisher")
  res <- GenTable(god, classicFisher=ea, topNodes=length(god@graph@nodes))
  res$Term2 <- plgINS::breakStrings(res$Term)
  res$log2Enrichment <- log2(res$Significant/res$Expected)
  res$FDR <- p.adjust(res$classicFisher)
  res
}

deaGSEA <- function(dea, gsets=NULL){
  if(is.null(gsets))
    gsets <- plgINS::getMsigSets("Mus musculus", collections=c("H","C5"))
  signature <- setNames(sign(dea$logFC)*-log10(dea$P.Value), row.names(dea))
  fg <- fgseaMultilevel(gsets, signature, minSize=10, nproc=3)
  fg$genes <- sapply(fg$leadingEdge, collapse=", ", FUN=paste)
  fg$leadingEdge <- NULL
  as.data.frame(fg[order(fg$pval),])
}
```

for 2-cell
```{r}
se <- readRDS("objects/MK_EXP20.1_ProtacEmbryo_merged.SE.rds")

deas <- getDEA(se)
dea <- deas$PROTAC
```
```{r, eval=FALSE}
res2 <- deaGSEA(dea)
head(res2)
#write.csv(res2, "./results/GSEA_PROTAC.csv")
colnames(res2)
res2
#saveRDS(res2, "GSEA_2cell.rds")
```
```{r}
res2 <- readRDS("rds_saves/paper/GSEA_2cell.rds")
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$padj < 0.05
  ggplot(df, aes(x=NES, y=-log10(padj))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black")) +
    geom_text_repel(data = subset(df, highlight == TRUE), aes(label=pathway), size=3.5, min.segment.length = 0) +
    xlab("logFC") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    ggtitle(title)
}
generate_volcano_plot(res2, "Enriched gene sets in 2-cell embryos upon KH-103 treatment")
ggsave("plots_svg/GSEA_EXP20-1_PROTAC.svg")
ggsave("plots_pdf/GSEA_EXP20-1_PROTAC.pdf")
ggsave("plots_png/GSEA_EXP20-1_PROTAC.png")
```

alternative plot

```{r}
ggplot(head(res2,10), aes(x = reorder(pathway, NES), y = NES, fill = -log10(padj))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "grey", high = "orange") +
  theme_minimal() +
  labs(
    title = "Gene Ontology Enrichment overall",
    x = "GO Term",
    y = "Enrichment Score",
    fill = "-log10(FDR)"
  )
```

for 8-cell
```{r}
se <- readRDS("objects/MK_EXP20.2_ProtacEmbryo8cells.SE.rds")

deas <- getDEA(se)
dea <- deas$PROTAC
```
```{r, eval = FALSE}
res4 <- deaGSEA(dea)
res4

#saveRDS(res4, "GSEA_8cell.rds")
```
```{r}
res4 <- readRDS("objects/GSEA_8cell.rds")
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$padj < 0.05
  ggplot(df, aes(x=NES, y=-log10(padj))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black")) +
    geom_text_repel(data = subset(df, highlight == TRUE), aes(label=pathway), size=3.5, min.segment.length = 0) +
    xlab("NES") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    ggtitle(title)
}
generate_volcano_plot(res4, "Enriched gene sets in 8-cell embryos upon KH-103 treatment")
ggsave("plots_svg/GSEA_EXP20-2_PROTAC.svg")
ggsave("plots_pdf/GSEA_EXP20-2_PROTAC.pdf")
ggsave("plots_png/GSEA_EXP20-2_PROTAC.png")
```



# Fig 4E - 2cell Smartseq after Dex treatment in sperm and Fig S4F - 2cell Smartseq after KH-103 treatment in sperm

```{r}
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$adj.P.Val < 0.1
  top_genes <- df[order(df$adj.P.Val), ][1:10, ]
  ggplot(df, aes(x=logFC, y=-log10(adj.P.Val))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black")) +
    geom_text_repel(data = subset(top_genes, highlight == TRUE), aes(label=Gene), size=3.5, min.segment.length = 0, max.overlaps = Inf, nudge_y = 0.1) +
    geom_hline(yintercept=-log10(0.05), linetype="dashed")+
    xlab("logFC") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    ggtitle(title)
}

se <- readRDS("objects/MK_EXP16.4_embryoSeq_spermProtac_merged.SE.rds")

# Generate the dataframes 
dex <- rowData(se)[[4]]
protac <- rowData(se)[[3]]
pd <- rowData(se)[[5]]

protac$Gene <- rownames(protac)
dex$Gene <- rownames(dex)
pd$Gene <- rownames(pd)

# List of dataframes
df_list <- list(Dex = dex, "KH-103" = protac, "KH-103+Dex" = pd)

# Apply the function to each dataframe in the list
plots <- lapply(names(df_list), function(name) {
  generate_volcano_plot(df_list[[name]], paste("Differentially expressed genes in 2-cell embryos upon ", name, " treatment in sperm"))
})
print(plots[1])
ggsave("plots_svg/DEGs_EXP16-4_Dex_volcano.svg")
ggsave("plots_pdf/DEGs_EXP16-4_Dex_volcano.pdf")
ggsave("plots_png/DEGs_EXP16-4_Dex_volcano.png")

print(plots[2])
ggsave("plots_svg/DEGs_EXP16-4_PROTAC_volcano.svg")
ggsave("plots_pdf/DEGs_EXP16-4_PROTAC_volcano.pdf")
ggsave("plots_png/DEGs_EXP16-4_PROTAC_volcano.png")

print(plots[3])
ggsave("plots_svg/DEGs_EXP16-4_PD_volcano.svg")
ggsave("plots_pdf/DEGs_EXP16-4_PD_volcano.pdf")
ggsave("plots_png/DEGs_EXP16-4_PD_volcano.png")
```


# Fig 4F - Reversal of Dex by KH-103

Can we see reversal for genes not passing significance?

```{r}
se <- readRDS("objects/MK_EXP16.4_embryoSeq_spermProtac_merged.SE.rds")
se <- se[, se$treatment != "MED"]
rescue <- rowData(se)[[6]]
rescue$Gene <- rownames(rescue)

rescue_sorted <- rescue[order(rescue$adj.P.Val), ]
rescue_sorted <- rescue_sorted[1:20, ]

treatments <- unique(se$treatment)

# Define your own colors
colors <- c(brewer.pal(9,"Set2")[3], "grey", "orange", brewer.pal(9,"Set2")[5] )#"#0000FF", "#800080", "grey", "black")  

# Make sure the number of colors matches the number of treatments
if (length(colors) != length(treatments)) {
  stop("Number of colors must match number of treatments")
}

colors <- c("#919297", "#DEDFE4", "darkgrey", "#393A3F")
# Create a named vector for color mapping
color_mapping <- setNames(colors, treatments)

# Add color_mapping to the heatmap annotation
ha = HeatmapAnnotation(df = data.frame(treatment = factor(colData(se)$treatment, levels = names(color_mapping))),
                       col = list(treatment = color_mapping))

#sechm(se, rescue_sorted, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="KH103-mediated\nrescue of DEX effect", gaps_at="treatment")
#needs title: Effects of KH-103 treatment in sperm on gene regulation in 2-cell embryos


color_palette <- colorRampPalette(c("blue", "black", "orange"))(10000)

pdf("plots_pdf/DEGs_EXP16-4_rescue_heatmap.pdf", width=6, height=4) #, width=14.5, height=10)
x <- sechm(se, rescue_sorted, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="KH103-mediated\nrescue of DEX effect", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
print(x)
dev.off()

png("plots_png/DEGs_EXP16-4_rescue_heatmap.png")#, height=768)
x <- sechm(se, rescue_sorted, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="KH103-mediated\nrescue of DEX effect", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
print(x)
dev.off()

svg("plots_svg/DEGs_EXP16-4_rescue_heatmap.svg")#, height=768)
x <- sechm(se, rescue_sorted, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="KH103-mediated\nrescue of DEX effect", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
print(x)
dev.off()
rescue_sorted

sechm(se, rescue_sorted, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
```


#Fig 4G - correlation with PGC-GR-KO

Calculate correaltion and visualize with scatterplot

```{r}
KO <- readRDS("objects/VF_2cell_embyro_early_paternal_GR_KO(Alpl).SE.rds")
KO <- rowData(KO)[[2]]
koFDR <- subset(KO, KO$FDR <0.05)
koFDR$Gene <- koFDR$ID


se <- readRDS("objects/SEs_for_SEV/MK_EXP16.4_embryoSeq_spermProtac_merged.SE.rds")
protac <- rowData(se)[[3]]
protac$Gene <- rownames(protac)


merged_data <- merge(koFDR, protac, by.x = "ID", by.y = "Gene")

correlation <- cor.test(merged_data$logFC.x, merged_data$logFC.y)
correlation

# Create a scatter plot to visualize the correlation
plot <- ggplot(merged_data, aes(x=logFC.x, y=logFC.y)) +
    geom_point() +
    geom_text_repel(aes(label=Gene)) +  # Add the gene names to the plot
    geom_smooth(method=lm, se=FALSE, color="orange") +
    #annotate("text", x = Inf, y = -Inf, label = paste("Correlation:", correlation)) +#, vjust = -1.5, hjust = 1.2) +
    xlab(paste("logFC in PGC-GR KO embryos")) +
    ylab(paste("logFC in embryos from KH-103 treated sperm")) +
    ggtitle(paste("Correlation of gene expression changes \n between PGC-GR KO embryos and embryos from KH-103 treated sperm"))+
    theme_classic()

print(plot)
ggsave("plots_svg/correlation_VF-KO_EXP16-4_PROTAC.svg")
ggsave("plots_pdf/orrelation_VF-KO_EXP16-4_PROTAC.pdf")
ggsave("plots_png/correlation_VF-KO_EXP16-4_PROTAC.png")
```


# Fig S4G - tracks accesscible in sperm

```{r}
gr1 <- txs[txs$gene_name == "1810019D21Rik"]
gr2 <- txs[txs$gene_name == "Magea14"]
gr3 <- txs[txs$gene_name == "Gm9895"]
gr4 <- txs[txs$gene_name == "Zfp472"]
gr5 <- txs[txs$gene_name == "Cyp4f17"]
gr5 <- txs[txs$gene_name == "Adamts4"]
gr6 <- txs[txs$gene_name == "Tmem68"]
gr7 <- txs[txs$gene_name == "Ptpmt1"]
gr8 <- txs[txs$gene_name == "Zfp873"] #
gr9 <- txs[txs$gene_name == "Trib3"] #

ensdb <- ah[["AH89457"]]

txs <- transcripts(ensdb, columns=c("tx_id","tx_biotype","tx_name","gene_id","gene_name"))

tracks <- c(
  Spermatozoa="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw"
)


plotSignalTracks(tracks, region=gr4[1], ensdb=ensdb, transcripts="full", genomeAxis = TRUE)

plotSignalTracks(tracks, region=gr6[1], ensdb=ensdb, transcripts="full", genomeAxis = TRUE)

plotSignalTracks(tracks, region=gr8[1], ensdb=ensdb, transcripts="full", genomeAxis = TRUE)

gene_symbols <- mapIds(org.Mm.eg.db, keys=tracks$gene_id, column="SYMBOL", keytype="ENSEMBL", multiVals="first")
tracks$gene_symbol <- gene_symbols

```

```{r}
pdf("plots_pdf/rescue_tracks.pdf" , width=10, height=5)
x <- plotSignalTracks(tracks, region=gr8[1], ensdb=ensdb, transcripts="full", genomeAxis = TRUE)
print(x)
dev.off()

png("plots_png/rescue_tracks.png", height=768)
x <- plotSignalTracks(tracks, region=gr8[1], ensdb=ensdb, transcripts="full", genomeAxis = TRUE)
print(x)
dev.off()
```

# Fig S4H - RegionUpset
```{r}
df <- read.csv("objects/DEA_rescueDEGs_Fig4_venn_categories.csv", sep = ";")
# Create a list of vectors, each vector represents a category
sperm_overlaps <- df$Gene[df$Overlap.sperm.ATAC == 1]
gre <- df$Gene[df$GRE.in.gene == 1]
motifs <- df$Gene[df$gene.body.motifs.connected.to.GR == 1]

gene_list <- list(DEGs = df$Gene, "DEGs accesible in Sperm" = sperm_overlaps, "DEGs with GRE motif" = gre, "DEGs with GR binding partner motif" = motifs) 
```





```{r}
library(UpSetR)

regionUpset2 <- function (x = NULL, genelist = NULL, returnList = FALSE, ...) 
{
    if (!is.null(genelist)) {
        x <- genelist
    }
    if (is.character(x)) 
        x <- as.list(x)
    stopifnot(length(x) > 1)
    if (is.list(x)) {
        if (is.null(names(x)) && all(unlist(lapply(x, is.character))) && 
            all(lengths(x) == 1)) 
            names(x) <- unlist(x)
        if (any(unlist(lapply(x, FUN = function(x) {
            if (is.character(x)) {
                if (length(x) != 0) return(TRUE)
                if (!file.exists(x)) return(TRUE)
            }
            FALSE
        })))) {
            stop("`x` should be a named list of character vectors.")
        }
    }
    x <- lapply(x, function(x) as.character(x))
    if (returnList) 
        return(x)
    UpSetR::upset(UpSetR::fromList(x), ...)
}
```


```{r}
png(paste0("plots_png//Fig4rescue.png"),width=8, height=7,res = 300,units = "in")
regionUpset2(gene_list)
dev.off()

pdf(paste0("plots_pdf/Fig4rescue.pdf"), width=8, height=7,colormodel = "cymk")
regionUpset2(gene_list)
dev.off()


svg(paste0("plots_svg/Fig4rescue.svg"), width=8, height=7)
regionUpset2(gene_list)
dev.off()

```

# Fig S4H - 2cell volcano


```{r}
se <- readRDS("objects/MK_EXP20.1_ProtacEmbryo_merged.SE.rds")

# Generate the dataframes
dex <- rowData(se)[[4]]
protac <- rowData(se)[[3]]
pd <- rowData(se)[[5]]

protac$Gene <- rownames(protac)
dex$Gene <- rownames(dex)
pd$Gene <- rownames(pd)

# List of dataframes
df_list <- list(Dex = dex, "KH-103" = protac, "KH-103+Dex" = pd)

# Function to generate volcano plots
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$FDR < 0.1
  top_genes <- df[order(df$FDR), ][1:10, ]
  ggplot(df, aes(x=logFC, y=-log10(FDR))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black")) +
    geom_text_repel(data = subset(top_genes, highlight == TRUE), aes(label=Gene), size=3.5, min.segment.length = 0, max.overlaps = Inf, nudge_y = 0.1) +
    geom_hline(yintercept=-log10(0.05), linetype="dashed")+
    xlab("logFC") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    ggtitle(title)
}

# Apply the function to each dataframe in the list
plots <- lapply(names(df_list), function(name) {
  generate_volcano_plot(df_list[[name]], paste("Differentially expressed genes in 2-cell embryos upon ", name, " treatment"))
})

print(plots)

print(plots[1])
ggsave("plots_svg/DEGs_EXP20-1_Dex_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-1_Dex_volcano.pdf")
ggsave("plots_pdf/DEGs_EXP20-1_Dex_volcano.png")

print(plots[2])
ggsave("plots_svg/DEGs_EXP20-1_PROTAC_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-1_PROTAC_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-1_PROTAC_volcano.png")

print(plots[3])
ggsave("plots_svg/DEGs_EXP20-1_PD_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-1_PD_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-1_PD_volcano.png")
```
# Fig 5C - DAA heatmap of Stress vs Control

et the differential peaks from IFS_ATAC

```{r}
se <- readRDS(file = "objects/Counts_broadPeak_noDup_bam.SE")
se$groups <- c("CTRL","IFS","CTRL","IFS")
se$groups <-  relevel(factor(se$groups), "CTRL")
rownames(se) <- granges(se)
```

```{r}

mm <- model.matrix(~groups, data=as.data.frame(colData(se)))

dds <- edgeR::DGEList(assay(se), group=se$groups)
dds <- edgeR::calcNormFactors(dds)

dds <- dds[filterByExpr(dds, mm, min.count=10,min.total.count=5),]
 dds <- estimateDisp(dds, mm)
 fit <- glmFit(dds, mm)
 res <- as.data.frame(topTags(glmLRT(fit),n = Inf))
```

```{r}
res_FDR <- res[res$FDR <= 0.05,] 

res_FDR$ranges <- rownames(res_FDR)
```



 with all significant genes not only the once with a gene

```{r}
se4 <- se[names(se)%in% rownames(res_FDR[res_FDR$logFC > 0,])] 
se5 <- se[names(se)%in% rownames(res_FDR[res_FDR$logFC < 0,])] 


significant_GR_ATAC_IFS <- unlist( GRangesList(granges(se4),granges(se5)))
```



```{r}
norm_IFS_bw <- list.files(path = "intern_data/ATAC/",pattern = "*full.norm13.bw$",full.names = T)
names(norm_IFS_bw) <- gsub(x = basename(norm_IFS_bw),pattern = ".full.norm13.bw",replacement = "")
# reorder
norm_IFS_bw <- norm_IFS_bw[c(1,3,4,2)]
norm_IFS_bw_2 <-norm_IFS_bw[c(1,2)]
```


```{r}
significant_GR_ATAC_IFS <- resize(significant_GR_ATAC_IFS,width = 660,fix = "center")

sm1 <- signal2Matrix(filepaths = norm_IFS_bw,regions = significant_GR_ATAC_IFS,w = 10L,extend = 2000)# extend doesnt work ?

```


list of heatmap
```{r}
i <- rep("Increased Accessibility", sum(res_FDR$logFC < 0))
i[57:134] <- "Reduced Accessibility"
```


```{r,fig.width= 12}
my_colors <- (c("#c0c0ffff", "black", "#ffa500ff"))
#mean_color <- (c("#c0c0ffff", "#ffa500ff"))
mean_color <- c("Increased Accessibility"="#ffa500ff","Reduced Accessibility"="#c0c0ffff")
x <- as.factor(mean_color)
x <- relevel(x,"#ffa500ff")
colnames(sm1) <- c("CTRL1","CTRL2","IFS1","IFS2")
p1 <- plotEnrichedHeatmaps(sm1,scale_title = "log2FC",scale_rows = F,colors = my_colors,top_annotation = F,trim = c(0.001,0.999),mean_scale_side = "left",row_split = i,axis_name = c("-2kb","0","+2kb"),mean_color = x,use_raster = T,heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL));p1




```

```{r}


  png(paste0("plots_png/Fig5C.png"), width=1000, height=768)
  draw(p1, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/Fig5C.pdf"), width=10, height=10)
  draw(p1, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig5C.svg"), width=10, height=10)
  draw(p1, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```


# Fig 5D - Ubr2 Stress and GR tracks


```{r}
# list files
bw_full <- list.files(path = "intern_data/ATAC/",pattern = "full.bw$",full.names = T)
# rename
names(bw_full)<- gsub(x = basename(bw_full),pattern = ".full.bw",replacement = "" )
bw_full <- bw_full[c(1,3,2,4)]
```

```{r}
bam2bw_list <- list.files(path = "intern_data/ChIP/",pattern = "*subtract.bw$",full.names = T)
names(bam2bw_list)<- basename(bam2bw_list)
```

```{r}
png(paste0("plots_png/Fig5D.png"), width=750, height=512)

plotSignalTracks(files = list(IFS=bw_full[c(3,4)],CTRL =bw_full[c(1,2)],GR = bam2bw_list[1]),overlay.alpha = 85,region = "17:47261741-47262114",aggregation = "overlay",colors = c("orange","darkgrey","steelblue"),genomeAxis = T,ensdb = ensdb,extend = c(10000,10000),ylim= c(0,10), fontface.main = 1,cex.main = 1,cex.id=0.9,cex.title = 0.7) # , rotation.title = 0
grid.text(paste0("Peak: 17:47261563-47262222","; ","Ubr2"), x = 0.28, y = 0.95, gp = gpar(fontsize = 14, fontface = "bold"))

dev.off()

pdf(paste0("plots_pdf/Fig5D.pdf"), width=8, height=7,colormodel = "cymk")
plotSignalTracks(files = list(IFS=bw_full[c(3,4)],CTRL =bw_full[c(1,2)],GR = bam2bw_list[1]),overlay.alpha = 85,region = "17:47261741-47262114",aggregation = "overlay",colors = c("orange","darkgrey","steelblue"),genomeAxis = T,ensdb = ensdb,extend = c(10000,10000),ylim= c(0,10), fontface.main = 1,cex.main = 1,cex.id=0.9,cex.title = 0.7) # , rotation.title = 0
grid.text(paste0("Peak: 17:47261563-47262222","; ","Ubr2"), x = 0.28, y = 0.95, gp = gpar(fontsize = 14, fontface = "bold"))
dev.off()


svg(paste0("plots_svg/Fig5D.svg"), width=8, height=7)
plotSignalTracks(files = list(IFS=bw_full[c(3,4)],CTRL =bw_full[c(1,2)],GR = bam2bw_list[1]),overlay.alpha = 85,region = "17:47261741-47262114",aggregation = "overlay",colors = c("orange","darkgrey","steelblue"),genomeAxis = T,ensdb = ensdb,extend = c(10000,10000),ylim= c(0,10), fontface.main = 1,cex.main = 1,cex.id=0.9,cex.title = 0.7) # , rotation.title = 0
grid.text(paste0("Peak: 17:47261563-47262222","; ","Ubr2"), x = 0.28, y = 0.95, gp = gpar(fontsize = 14, fontface = "bold"))
dev.off()

```


# Fig S5B
here we show the heatmaps between the 4 IFS samples (broadpeaks)

```{r}

peak_list <- list.files(path = "intern_data/ATAC/",pattern = ".broadPeak$",full.names = T)
names(peak_list) <- basename(peak_list)

peak_name_list <- c()
for (name in 1:length(basename(peak_list))){
  peak_name_list[name] <- gsub("_MKD(.*).broadPeak$",replacement = "",x = basename(peak_list)[name]) # give clean name from sequencing data from encode
}
names(peak_list) <- peak_name_list
names(peak_list) <-  peak_name_list# we have 2 IFS_9 files, we keep them spearate for now
names(peak_list) <- gsub(x = names(peak_list),pattern = "FS_",replacement = "FS_noDup_")

```

```{r}
peaks_noDup <- list()
for (f in 1:length(peak_list)){
  peaks_noDup[f] <- import(con = peak_list[f],format = "broadPeak")
}

names(peaks_noDup) <- names(peak_list)

```
```{r}
#peaks
total_consensus_peaks <- reduce(unlist(GRangesList(peaks_noDup)), with.revmap=TRUE)
y <- keepStandardChromosomes(total_consensus_peaks, pruning.mode = "coarse")
total_consensus_peaks <- granges(y[lengths(y$revmap)>1])
blacklist <- import("objects/blacklist.liftedTo.mm39.bed")
seqlevelsStyle(blacklist) <- "ensembl"
total_consensus_peaks <- total_consensus_peaks[!overlapsAny(total_consensus_peaks, blacklist)]# thats how you get rid of blacklisted peaks!!!!
total_consensus_peaks
broadPeak_1 <- total_consensus_peaks
hist(width(broadPeak_1),xlim = c(0,2000),breaks = 5000)
median(width(broadPeak_1))
```

load bw
```{r}
# list files
bw_full <- list.files(path = "intern_data/ATAC/",pattern = "full.bw$",full.names = T)
# rename

bw_full <- bw_full[c(1,3,4,2)]


names(bw_full) <- c("Ctrl 1", "Ctrl 2", "IFS 1", "IFS 2")
```



generate signalmatrix
```{r}

sm <- signal2Matrix(filepaths = c(bw_full),regions = broadPeak_1,w = 10)
sm <- epiwraps:::.ese2ml(sm)
sn <-  renormalizeBorders(lapply(sm,log1p))
```

```{r}

p_IFS1 <-  plotEnrichedHeatmaps(sm,row_title = "IFS accessible sites",scale_title = "desnity",colors = rocket(2000),mean_scale_side = "none", raster_resize_mat = TRUE, ,use_raster = TRUE, axis_name = c("-2kb","0","+2kb"), trim = c(0.01,0.99), heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))

p_IFS1



p_IFS2 <-  plotEnrichedHeatmaps(sn,row_title = "IFS accessible sites",scale_title = "desnity",colors = rocket(2000), raster_resize_mat = TRUE, use_raster = TRUE,axis_name = c("-2kb","0","+2kb"),trim = c(0.005,0.995), heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))

p_IFS2

```
```{r}
  # Save each heatmap to a PDF
  pdf("plots_pdf/FigS5B_IFS_heatmap.pdf", width=10, height=10)
  draw(p_IFS2, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  

```


```{r}
  svg("plots_svg/FigS5B_IFS_heatmap.svg", width=10, height=10)
  draw(p_IFS2, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  
  png("plots_png/FigS5B_IFS_heatmap.png", width=1000, height=1000)
  draw(p_IFS2, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
```


# Fig S5C - volcano DAA

ChIpeakAnno
```{r}
?ChIPpeakAnno::annoPeaks
ah <- AnnotationHub()
ensdb <- ah[["AH89457"]]
annGR <- ChIPpeakAnno::toGRanges(ensdb)
seqlevelsStyle(annGR)<- "Ensembl"

#ChIPpeakAnno_gr <- ChIPpeakAnno::annoPeaks(peaks_GR,annoData = annGR)


```


```{r}
ChIPpeakAnno_IFS_long <- ChIPpeakAnno::annoPeaks(granges(se),annoData = annGR,bindingType = "fullRange",select = "bestOne",bindingRegion = c(-10000,10000))
se2 <- se[names(se) %in% ChIPpeakAnno_IFS_long$peak]
res2 <- res[rownames(res) %in% ChIPpeakAnno_IFS_long$peak,]
ChIPpeakAnno_IFS_long <- ChIPpeakAnno_IFS_long[order(ChIPpeakAnno_IFS_long$peak)]
res2 <- res2[order(rownames(res2)),]; table(rownames(res2)==ChIPpeakAnno_IFS_long$peak)
res2$Gene <- ChIPpeakAnno_IFS_long$gene_name
```

```{r,fig.width=12}

res2 <- res2[order(res2$PValue),]
w <- res2$FDR < 0.05
ggplot(res2, aes(logFC, -log10(FDR))) + 
  geom_text_repel(data=head(res2[w,],50), aes(label=Gene),max.overlaps = 20,nudge_y= 0.05)+
  geom_hline(yintercept=-log10(0.05), linetype="dashed")+
  annotate(geom = "text",x = -7,y =-log10(0.05)-0.5,label = "FDR = 0.05")+
  geom_point()+
  ggtitle("Differentially accessible loci associated with genes")


  ggsave("./FigS5C_plots_svg/PGC_GR_KO_DAA_volcano.svg")
ggsave("./FigS5C_plots_pdf/PGC_GR_KO_DAA_volcano.pdf")#, width = 6, height = 4)
ggsave("./FigS5C_plots_png/PGC_GR_KO_DAA_volcano.png")
```

# Fig S5D - GO term of DAA genes


```{r}
 library(AnnotationHub)
ah <- AnnotationHub()   
ensdb <- ah[["AH89457"]] # GRCm39
annGR <- ChIPpeakAnno::toGRanges(ensdb)

seqlevelsStyle(annGR)<- "Ensembl"

```

```{r}
DDA_anno <-  ChIPpeakAnno::annoPeaks(granges(se),annoData = annGR,bindingType = "fullRange",select = "bestOne",bindingRegion = c(-1000,1000))

res2 <- res[rownames(res) %in% DDA_anno$peak,]
res2$peak <- rownames(res2)
res3 <- merge(res2, as.data.frame(DDA_anno), by = "peak")
res3 <- res3[order(res3$PValue),]
cleaned_res3 <- res3[!duplicated(res3$gene_name), ]
```

```{r}
gene_list <- cleaned_res3[cleaned_res3$FDR <= 0.05,]$gene_name # this get us all significantly differential expressed genes
ego <- enrichGO(gene          = gene_list,
                OrgDb         = org.Mm.eg.db,
                keyType       = "SYMBOL",
                ont           = "BP",  # BP: Biological Process, MF: Molecular Function, CC: Cellular Component
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.5,
                qvalueCutoff  = 0.5,
                universe = cleaned_res3$gene_name # the background is really important, always try to think about what background would be the best, in our case for instance we can take all the gene passing the threshold of counting 
                    )

ego@result  





```
```{r,fig.width=10,fig.height=9}
# Bar plot of enriched GO terms
barplot(ego, showCategory=20)

# Dot plot of enriched GO terms
dotplot(ego, showCategory=20)

```


```{r}

png(paste0("plots_png/FigS5D.png"), width=750, height=512)
dotplot(ego, showCategory=20)
dev.off()

pdf(paste0("plots_pdf/FigS5D.pdf"), width=8, height=7,colormodel = "cymk")
dotplot(ego, showCategory=20)
dev.off()


svg(paste0("plots_svg/FigS5D.svg"), width=8, height=7)
dotplot(ego, showCategory=20) 
dev.off()

```


