---
title: "Paper-plots"
author: "Miriam Kretschmer & Vincent Fischer"
date: "28/02/2025"
output:
  bookdown::html_document2:
    code_folding: hide
bibliography: citations.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("/mnt/gapp/work/vincent/paper/Kretschmer2025/GitHub_submission2"))
```

```{r}
suppressPackageStartupMessages({
library(AnnotationHub)
library(rtracklayer)
library(universalmotif)
library(MotifDb)
library(Rsamtools)
library(memes)
library(rGREAT)
library(MotifDb)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(SummarizedExperiment)
library(SEtools)
library(sechm)
library(fgsea)
library(msigdbr)
library(RColorBrewer)
library(ggvenn)
library(viridis)
library(cowplot)
library(epiwraps)
  library(chromVAR)
  library(edgeR)
  library(limma)
  library(org.Hs.eg.db)
  library(org.Mm.eg.db)
  library(clusterProfiler)
  })
set.seed(123)
ah <- AnnotationHub()
source("objects/BFA_BFF.R")
source("objects/DLCAnalyzer.R")
```


```{r, eval = T}
# function to get non redundant motifs from MotifDB
getNonRedundantMotifs <- function(){
  motifs <- MotifDb::query(MotifDb::MotifDb, c("Mmusculus","HOCOMOCO"))
  modf <- data.frame(row.names=names(motifs),
                     TF=gsub("^Mmusculus-HOCOMOCOv10-|_MOUSE.+","",names(motifs)),
                     grade=gsub(".+\\.","",names(motifs)))
  modf <- modf[order(modf$TF,modf$grade),]
  modf <- modf[!duplicated(modf$TF),]
  setNames(universalmotif::convert_motifs(motifs[row.names(modf)]), modf$TF)
}
motifs <- getNonRedundantMotifs()
motifs$GCR
```


# Fig 1A - Motif enrichment sperm and oocyte

Sperm peaks
```{r}
d.sperm <- readRDS("objects/motif_enrichment_sperm_consensus_narrow.rds")
d.sperm <- readRDS("objects/motif_enrichment_sperm_consensus_broad.rds")

head(d.sperm)
length(d.sperm$motif_id)#281
```

MII oocyte peaks
```{r}
d.oo <- readRDS("objects/motif_enrichment_oocyte_broad.rds")

head(d.oo)
length(d.oo$motif_id)
d.oo$motif_id
```

Comparing the enrichments directly, e.g. by merging two data.frames and plotting the log(p) in one vs the other highlighting top 5 for each 
```{r}
m <- merge(d.oo, d.sperm, by="motif_id", suffixes=c(".oocyte",".sperm"))

m.oo.sort <- m[order(m[,2],decreasing = FALSE),]
m.oo5 <- m.oo.sort[1:3,]
m.sperm.sort <- m[order(m[,18],decreasing = FALSE),]
m.sperm5 <- m.sperm.sort[1:3,]
m
```

```{r}
ggplot(m, aes(x=log2((1+tp.sperm)/(1+fp.sperm)), y= log2((1+tp.oocyte)/(1+fp.oocyte)), size=(tp_percent.sperm))) + 
  geom_point() + 
  geom_text_repel(box.padding=0.5, min.segment.length = 0, max.overlaps = 10, hjust = "left", aes(label=motif_id), data=m.oo5, size=4) + 
  geom_text_repel(box.padding= 0.5, min.segment.length = 0, max.overlaps = 10, hjust = "right", aes(label=motif_id), data=m.sperm5)+
  geom_point(color="orange", data=m.oo5) +
  geom_point(color="blue", data=m.sperm5) +
  labs(color="Top enriched motifs")+
  labs(x="log2(fold-enrichment in spermatozoa)", y="log2(fold-enrichment in MII oocytes)", size="% of set") +
  #ylab("-log10 (adjusted p-value MII oocyte)")+
  #xlab("-log10 (adjusted p-value spermatozoa)")+
  theme_classic() +
  #theme(legend.position = "none")+
  ggtitle("Top ranked motifs in peaks of spermatozoa vs. MII oocytes")

  ggsave("./plots_svg/motif_enrichment_sperm_vs_oocyte_peaks.svg")
ggsave("./plots_pdf/motif_enrichment_sperm_vs_oocyte_peaks.pdf")#, width = 6, height = 4)
ggsave("./plots_png/motif_enrichment_sperm_vs_oocyte_peaks.png")
```


# Fig 1B - TF motif activity sperm vs oocyte

```{r}
library(TFBSTools)
sc <- readRDS("./objects/sc_sperm_oo.rds")
fa <- Rsamtools::FaFile("objects/Mus_musculus.GRCm39.dna_sm.primary_assembly.fa")

sc_gc <- addGCBias(sc, genome=fa)
motifs <- do.call(what = "PFMatrixList", lapply(getNonRedundantMotifs(), class="TFBSTools-PFMatrix", FUN= convert_motifs))
#motifs <- getNonRedundantMotifs()
motif_ix <- motifmatchr::matchMotifs(motifs, sc_gc, genome=fa)
motif_ix
dev <- computeDeviations(object=sc_gc, annotations=motif_ix)
dev$groups <- c("sperm","sperm","MIIoocyte")

fit <- lmFit(assays(dev)$z, model.matrix(~dev$groups),background_peaks = getBackgroundPeaks(sc_gc,rowData(sc_gc)$bias,niterations = 5000))

head(mores <- topTable(eBayes(fit),number=Inf),395)
sechm::sechm(dev, head(row.names(mores)), assayName = "z", top_annotation = "groups",do.scale=TRUE)
sechm::sechm(dev, head(row.names(mores),395), assayName = "z", top_annotation = "groups",do.scale = TRUE)
```

```{r}
ggplot(mores, aes(logFC, -log10(adj.P.Val))) + 
  geom_point() + 
  geom_hline(yintercept=-log10(0.05), linetype="dashed") +
  ggtitle("Transcription factor motif activity spermatozoa vs MIIoocyte") +
  theme_classic() +
  labs(y = "-log10(adjusted p-value)") +
  geom_text_repel(data = subset(mores, adj.P.Val < 0.01), 
                  aes(label = rownames(subset(mores, adj.P.Val < 0.01))), 
                  min.segment.length = 0,
                  size = 4)
#geom_point(aes(color = B)) +

ggsave("./plots_svg/TFmotif_activity_sperm_vs_oocyte_peaks.svg")
ggsave("./plots_pdf/TFmotif_activity_sperm_vs_oocyte_peaks.pdf")#, width = 6, height = 4)
ggsave("./plots_png/TFmotif_activity_sperm_vs_oocyte_peaks.png")
```




# Fig 1C - Human sperm 

```{r}
genome <- ah[["AH90808"]] #human 

GRE_motif <- MotifDb::MotifDb %>% 
  MotifDb::query("GRE") %>% 
  universalmotif::convert_motifs() %>% 
  .[[1]]
```

```{r GRE-peaks-hcp}
peaks <- import("extern_data/ATAC_Corces_1_peaks.narrowPeak")
peak_seqs <- get_sequence(peaks, genome)
fimo <- runFimo(peak_seqs, GRE_motif, meme_path = "/common/meme/bin") # you need to insert your indiviudual meme path here
GRE_peaks1 <- peaks[overlapsAny(peaks, fimo)]
```
```{r number-peaks-hcp, results='hide'}
length(peaks) #32027
length(fimo) #1338
sum(overlapsAny(peaks, fimo)) #716
length(GRE_peaks1) #716
```

```{r, eval=T}
tracks <- c(
  sperm1="extern_data/ATAC_Corces_1.full.bw",
  sperm2="extern_data/ATAC_Corces_2.full.bw"
)

m <- signal2Matrix(tracks, GRE_peaks1, w = 20, BPPARAM = 4)
m <- epiwraps:::.ese2ml(m)

fused_sperm_mat <- mergeSignalMatrices(m[1:2])
fused_sperm_mat

fused_m <- c(list("Human sperm"=fused_sperm_mat), m[-1:-2])


pdf("plots_pdf/humansperm.pdf", width=7.5, height=10)
h <- plotEnrichedHeatmaps(fused_m, row_title="Peaks from spermatozoa with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_trim = TRUE, mean_color = "#800080", heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL), use_raster = TRUE)
print(h)
dev.off()

png("plots_png/humansperm.png")
h <- plotEnrichedHeatmaps(fused_m, row_title="Peaks from spermatozoa with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080", heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL), use_raster = TRUE)
print(h)
dev.off()

svg("plots_svg/humansperm.svg", width=7.5, height=10)
h <- plotEnrichedHeatmaps(fused_m, row_title="Peaks from spermatozoa with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080",heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL), use_raster = TRUE)
print(h)
dev.off()
```


# Fig 1D - Sperm GRE peaks across spermatogenesis

```{r}
tracks <- c(
  uds1="extern_data/ATACse_PRJNA399533_Thy1posSpermatogonia_GSM2751121.bw",
  uds2="extern_data/ATACse_PRJNA399533_Thy1posSpermatogonia_GSM2751122.bw",
  ds1="extern_data/ATACse_PRJNA399533_cKitposSpermatogonia_GSM2751125.bw",
  ds2="extern_data/ATACse_PRJNA399533_cKitposSpermatogonia_GSM2751126.bw",
  ps1="extern_data/ATACse_PRJNA399533_PachytenSpermatocytes_GSM2751129.bw",
  ps2="extern_data/ATACse_PRJNA399533_PachytenSpermatocytes_GSM2751130.bw",
  rs1="extern_data/ATACse_PRJNA399533_RoundSpermatides_GSM2751133.bw",
  rs2="extern_data/ATACse_PRJNA399533_RoundSpermatides_GSM2751134.bw",
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data/ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  GVoocyte1="extern_data/ATAC_PRJNA480448_GVOocyte_GSM3262861.bw",
  GVoocyte2="extern_data/ATAC_PRJNA480448_GVOocyte_GSM3262862.bw",
  "MII oocyte"="extern_data/ATAC_PRJNA480448_MIIOocyte_GSM3262863.bw"
)
```

# create symbolic links

```{r}
peaks <- readRDS("objects/narrow_peaks_with_GRE_sperm.rds")
m <- signal2Matrix(tracks, peaks, w = 20, BPPARAM = 2)
m <- epiwraps:::.ese2ml(m)

fused_uds_mat <- mergeSignalMatrices(m[1:2])
fused_ds_mat <- mergeSignalMatrices(m[3:4])
fused_ps_mat <- mergeSignalMatrices(m[5:6])
fused_rs_mat <- mergeSignalMatrices(m[7:8])
fused_sperm_mat <- mergeSignalMatrices(m[9:12])
fused_gv_mat <- mergeSignalMatrices(m[13:14])

fused_m <- c(list(fused_uds=fused_uds_mat), list(fused_ds=fused_ds_mat), list(fused_ps=fused_ps_mat), list(fused_rs=fused_rs_mat), list(fused_sperm=fused_sperm_mat), list(fused_GVoocytes=fused_gv_mat), m[15], m[-1:-15])
fused_n <- renormalizeBorders(lapply(fused_m,log1p))

#fused_m <- readRDS("rds_saves/paper/spermatogenesis_fused_m.rds")
#fused_n <- readRDS("rds_saves/paper/spermatogenesis_fused_n.rds")
```

Combined in one plot
```{r}
pdf("plots_pdf/GREpeaks_from_sperm_across_spermatogenesis_normalized.pdf", width=17.5, height=10)
h <- plotEnrichedHeatmaps(fused_n, row_title="Peaks from spermatozoa with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()

png("plots_png/GREpeaks_from_sperm_across_spermatogenesis_normalized.png", height=768)
h <- plotEnrichedHeatmaps(fused_n, row_title="Peaks from spermatozoa with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()
```

Single panels of heatmaps 
```{r}
# Create a list of heatmaps
for (i in 1:length(fused_n)) {
  h <- plotEnrichedHeatmaps(fused_n[i], raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080",
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))
  
  # Save each heatmap to a PDF
  pdf(paste0("plots_pdf/GREpeaks_from_sperm_across_spermatogenesis_normalized_", i, ".pdf"), width=2.5, height=10)
  draw(h, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
}
```

```{r}
# Create a list of heatmaps
for (i in 1:length(fused_n)) {
  h <- plotEnrichedHeatmaps(fused_n[i], raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080",
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4, "npc"), title=NULL))

  # Save each heatmap to a PNG
  png(paste0("plots_png/GREpeaks_from_sperm_across_spermatogenesis_normalized_", i, ".png"), width=250, height=768)
  draw(h, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
}
```



# Fig 1E - Promotors with GRE; ATAC, H3k4m3 and 5mC


```{r}
tracks <- c(
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data/ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  sperm1_H3K4me3="extern_data/SRR3309386.bw",
  sperm2_H3K4me3="extern_data/SRR3309387.bw",
  sperm3_H3K4me3="extern_data/SRR3703039.bw",
  sperm4_H3K4me3="extern_data/SRR5049005.bw",
  sperm5_H3K4me3="extern_data/SRR3225849.bw",
  sperm1_H3K27ac="extern_data/SRR3225845.bw",
  sperm2_H3K27ac="extern_data/SRR5049001.bw",
  sperm1_H3K9ac="extern_data/SRR3225847.bw",
  sperm2_H3K9ac="extern_data/SRR5049003.bw",
  sperm1_H3K4me1="extern_data/SRR3225848.bw",
  sperm2_H3K4me1="extern_data/SRR5049004.bw",
  sperm1_H3.3="extern_data/ChIP_PRJNA480447_Sperm_GSM3262880.bw",  
  sperm2_H3.3="extern_data/ChIP_PRJNA480447_Sperm_GSM3262881.bw",
  sperm1_H2A.Z="extern_data/ChIP_PRJNA480447_Sperm_GSM3262882.bw",
  sperm2_H2A.Z="extern_data/ChIP_PRJNA480447_Sperm_GSM3262880.bw"
)

```






```{r}
methyl_bw <- list.files(path = "intern_data/Nanopore/",pattern = "39.bw",full.names = T)
names(methyl_bw) <- c("GR_Ctrl1","GR_Ctrl2","GR_KO1","GR_KO2")

methyl_bw_ctrl <- methyl_bw[1:2]

```
```{r}
methyl_bed <- sapply(names(methyl_bw),function(names){
  import.bw(methyl_bw[names])
  
  })
```


promoters regions with GRE 
```{r}
proms <- readRDS("objects/GRE_promoters_center.rds")

```

```{r}
sm2 <- signal2Matrix(w = 20L, filepaths = tracks[c(1:4,5:9)],regions = proms)
#plotEnrichedHeatmaps(sm2)
```


```{r}
sm_5mC <- signal2Matrix(w = 20L , smooth = T, limit = c(0,100), filepaths = methyl_bed, regions = proms)
sm2 <- sm2[rownames(sm2) %in% rownames(sm_5mC),]
```




```{r}
sm_merge <- mergeSignalMatrices(epiwraps:::.ese2ml(sm2)[1:4])
sm_merge_H3K4me3 <- mergeSignalMatrices(epiwraps:::.ese2ml(sm2)[5:9])
sm_merge_5mC <- mergeSignalMatrices(epiwraps:::.ese2ml(sm_5mC)[1:2])
```

```{r}
p <- plotEnrichedHeatmaps(ml = list(ATAC = sm_merge,H3K4me3 = sm_merge_H3K4me3),multiScale = F,row_title = "Promoters with GRE",color = rocket(2000),scale_title = "density",axis_name = c("-2kb","TSS","+2kb"), use_raster = T,mean_scale_side = "none",  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))+
                       plotEnrichedHeatmaps(ml = list("5mC" = sm_merge_5mC),row_title = "Promoters with GRE",use_raster =TRUE, scale_title = "5mc[%]",axis_name = c("-2kb","TSS","+2kb"),mean_scale_side = "none",  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
# p
 #draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
 
 #draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")

  png(paste0("plots_png/Fig1E.png"), width=750, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/Fig1E.pdf"), width=7.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig1E.svg"), width=7.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```
each heatmap on its own
```{r}
#ATAC
p <- plotEnrichedHeatmaps(ml = list(ATAC = sm_merge),multiScale = F,row_title = "Promoters with GRE",color = rocket(2000),scale_title = "density",axis_name = c("-2kb","TSS","+2kb"), use_raster = T,  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
png(paste0("plots_png//Fig1EATAC.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf//Fig1EATAC.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg//Fig1EATAC.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
# H3K4m3 
  p <- plotEnrichedHeatmaps(ml = list(H3K4me3 = sm_merge_H3K4me3),multiScale = F,row_title = "Promoters with GRE",color = rocket(2000),scale_title = "density",axis_name = c("-2kb","TSS","+2kb"), use_raster = T,  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
png(paste0("plots_png//Fig1EH3K4me3.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf//Fig1EH3K4me3.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig1EH3K4me3.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
# 5mC
  
p <-   plotEnrichedHeatmaps(ml = list("5mC" = sm_merge_5mC),row_title = "Promoters with GRE",use_raster =TRUE, scale_title = "5mc[%]",axis_name = c("-2kb","TSS","+2kb"),  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
png(paste0("plots_png/Fig1EH3K4me3.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/Fig1EH3K4me3.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig1EH3K4me3.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
```


# Fig 1F GR-ChIP
```{r}
bw_ChIP <- list.files("intern_data/ChIP/",pattern = "*subtract",full.names = T)
names(bw_ChIP)<- gsub(x = basename(bw_ChIP),pattern = ".subtract.bw",replacement = "")
bw_GR <- bw_ChIP[1]
names(bw_GR) <- "GR"
GR_ap_broad <- import("intern_data/ChIP/VF_sp_GR_ap_EKDL220011603-1A_H5J3HDSX5_L2_peaks.broadPeak",format = "broadPeak")
```

```{r}
bed_GR <- import.bw(bw_GR) 
sm_GR <-  signal2Matrix(filepaths = list(GR = bed_GR),regions = GR_ap_broad,w = 10L,smooth = T, background= 0) 
```


```{r,fig.width=7,fig.height=9}

p_GR1 <-  plotEnrichedHeatmaps(sm_GR,row_title = "GR-ChIP sites",scale_title = "GR\ndesnity",colors = mako(2000), raster_resize_mat = TRUE, use_raster = TRUE,axis_name = c("-2kb","0","+2kb"), trim = c(0.01,0.99),width = unit(0.8,"npc"), heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))


#p_GR1

#p_GR2 <-  plotEnrichedHeatmaps(sn_GR,row_title = "GR-ChIP sites",scale_title = "GR\ndesnity",colors = c("black","steelblue1"),mean_scale_side = "none", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"),trim = c(0.005,0.995), heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))

#p_GR2
  draw(p_GR1, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")

  png(paste0("plots_png/Fig1F.png"), width=250, height=768)
  draw(p_GR1, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/Fig1F.pdf"), width=2.5, height=10)
  draw(p_GR1, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig1F.svg"), width=2.5, height=10)
  draw(p_GR1, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```


# Fig 1G Promoter with GRE GR,ATAC,H3K4m3,H3k27ac
```{r}
tracks <- c(
  
  sperm1_H3K4me3="extern_data/SRR3309386.bw",
  sperm2_H3K4me3="extern_data/SRR3309387.bw",
  sperm3_H3K4me3="extern_data/SRR3703039.bw",
  sperm4_H3K4me3="extern_data/SRR5049005.bw",
  sperm5_H3K4me3="extern_data/SRR3225849.bw",
  sperm1_H3K27ac="extern_data/SRR3225845.bw",
  sperm2_H3K27ac="extern_data/SRR5049001.bw",
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data/ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  sperm1_GR = "intern_data/ChIP/VF_sp_GR_ap.subtract.bw"
)


proms <- readRDS("objects/GRE_promoters_center.rds")
set.seed(1234)
#proms <- sample(proms,10000)

m <- signal2Matrix(tracks, proms, w = 20, BPPARAM = 8)
m <- epiwraps:::.ese2ml(m)



fused_H3K4me3 <- mergeSignalMatrices(m[1:5])
fused_H3K27ac <- mergeSignalMatrices(m[6:7])
fused_ATAC <- mergeSignalMatrices(m[8:11])


fused_m <- c(list(GR =m[[12]]),list(H3K27ac=fused_H3K27ac), list(H3K4me3=fused_H3K4me3),list(ATAC=fused_ATAC))
fused_n <- renormalizeBorders(lapply(fused_m,log1p))


bg_nf <- getNormFactors(tracks,method = "background")


bg <- epiwraps::renormalizeSignalMatrices(ml = m ,scaleFactors = bg_nf)



fused_H3K4me3_bg <- mergeSignalMatrices(bg[1:5])
fused_H3K27ac_bg <- mergeSignalMatrices(bg[6:7])
fused_ATAC_bg <- mergeSignalMatrices(bg[8:11])
fused_bg <- c(list(GR =bg[[12]]),list(H3K27ac=fused_H3K27ac_bg), list(H3K4me3=fused_H3K4me3_bg), list(ATAC = fused_ATAC_bg))
 

```

```{r}
sperm1_GR <- "intern_data/ChIP/VF_sp_GR_ap.subtract.bw"
```

```{r}
sperm1_GR_bed <- import.bw(sperm1_GR)
sm_gr_smooth <- signal2Matrix(filepaths = list(GR = sperm1_GR_bed),regions = proms,w = 20L,smooth  = T,background = 0)
```


```{r}
plotEnrichedHeatmaps(sm_gr_smooth,row_title = "Promoter with GRE",scale_title = "GR\ndesnity",colors = mako(2000),mean_scale_side = "none", trim = c(0.02,0.98),
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))+
plotEnrichedHeatmaps(fused_bg[2:4],row_title = "Promoter with GRE",scale_title = "HPTM\ndensity",colors = rocket(2000),mean_scale_side = "none",
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL),scale_rows = FALSE)

```

```{r}
p <- plotEnrichedHeatmaps(sm_gr_smooth,row_title = "Promoter with GRE",scale_title = "GR\ndesnity",colors = mako(2000),mean_scale_side = "none", trim = c(0.02,0.98),axis_name = c("-2kb","TSS","+2kb"),
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))+
plotEnrichedHeatmaps(fused_bg[c(4,3,2)],row_title = "Promoter with GRE",scale_title = "HPTM\ndensity",colors = rocket(2000),mean_scale_side = "none",axis_name = c("-2kb","TSS","+2kb"),
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL),scale_rows = FALSE)

ro <- p@ht_list[[1]]@row_order
# p
 #draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
 
 #draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")

  png(paste0("plots_png//Fig1G.png"), width=1000, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf//Fig1G.pdf"), width=10, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg//Fig1G.svg"), width=10, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```
each heatmap on its own
```{r}
#GR
p <- plotEnrichedHeatmaps(sm_gr_smooth,row_title = "Promoter with GRE",scale_title = "GR\ndesnity",colors = mako(2000), trim = c(0.02,0.98),axis_name = c("-2kb","TSS","+2kb"),row_order = ro,
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))
png(paste0("plots_png//Fig1G_GR.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf//Fig1G_GR.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg//Fig1G_GR.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
# ATAC 
  p <-plotEnrichedHeatmaps(fused_bg[c(4)],row_title = "Promoter with GRE",scale_title = "HPTM\ndensity",colors = rocket(2000),axis_name = c("-2kb","TSS","+2kb"),row_order = ro,
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL),scale_rows = FALSE)
png(paste0("plots_png//Fig1G_ATAC.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf//Fig1G_ATAC.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg//Fig1G_ATAC.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
# H3K4me3
  
p <-  plotEnrichedHeatmaps(fused_bg[c(3)],row_title = "Promoter with GRE",scale_title = "HPTM\ndensity",colors = rocket(2000),axis_name = c("-2kb","TSS","+2kb"),row_order = ro,
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL),scale_rows = FALSE)
png(paste0("plots_png//Fig1G_H3K4me3.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf//Fig1G_H3K4me3.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg//Fig1G_H3K4me3.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  
# H3K27ac
  
p <-  plotEnrichedHeatmaps(fused_bg[c(2)],row_title = "Promoter with GRE",scale_title = "HPTM\ndensity",colors = rocket(2000),axis_name = c("-2kb","TSS","+2kb"),row_order = ro,
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL),scale_rows = FALSE)
png(paste0("plots_png/Fig1G_H3K27ac.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/Fig1G_H3K27ac.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig1G_H3K27ac.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")  
  dev.off()
  
```



#Fig S1A - Comparison sperm data of published data and of Fischer and Kretschmer 2023

```{r}
peaks <- readRDS("objects/consensus.peaks_sperm.rds")
tracks <- c(
  "Jung 1"="extern_data//ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  "Jung 2"="extern_data//ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  "Jung 3"="extern_data//ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  "Jung 4"="extern_data//ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  "Fischer 1"="extern_data/NoMx_N704_KG.bw",
  "Fischer 2"="extern_data/NoMx_N705_KG.bw"
)

m <- signal2Matrix(tracks, peaks, w = 20, BPPARAM = 4)
m <- epiwraps:::.ese2ml(m)
n <- renormalizeBorders(lapply(m,log1p))

pdf("plots_pdf/sperm_comparisons.pdf", width=17.5, height=10)
h <- plotEnrichedHeatmaps(n, row_title="Peaks from spermatozoa", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_trim = TRUE)
print(h)
dev.off()

png("plots_png/sperm_comparisons.png", height=768)
h <- plotEnrichedHeatmaps(n, row_title="Peaks from spermatozoa", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()
```



# Fig S1B A549 atac motif enrichment

sperm vs A549 enrichment

```{r}
d.sperm <- readRDS("objects/motif_enrichment_sperm_consensus_broad.rds")
d.a549 <- readRDS("objects/motif_enrichment_sperm_A549.rds")
m <- merge(d.a549, d.sperm, by="motif_id", suffixes=c(".a549",".sperm"))

m.a549.sort <- m[order(m[,2],decreasing = FALSE),]
m.a5495 <- m.a549.sort[1:3,]
m.sperm.sort <- m[order(m[,18],decreasing = FALSE),]
m.sperm5 <- m.sperm.sort[1:3,]
m

gr <- m.sperm.sort[2, ]
stat1 <- m.a549.sort[1, ]
```
```{r}
ggplot(m, aes(x=log2((1+tp.sperm)/(1+fp.sperm)), y= log2((1+tp.a549)/(1+fp.a549)), size=(tp_percent.sperm))) + 
  geom_point() + 
  #geom_point(aes(color = ifelse(abs(log2((1+tp.sperm)/(1+fp.sperm)) - log2((1+tp.a549)/(1+fp.a549))) <= 1, "grey", "black"))) + 
  #geom_text_repel(box.padding=0.5, min.segment.length = 0, max.overlaps = 10, hjust = "left", aes(label=motif_id), data=m.a5495, size=4) + 
  #geom_text_repel(box.padding= 0.5, min.segment.length = 0, max.overlaps = 10, hjust = "right", aes(label=motif_id), data=m.sperm5)+
  geom_point(data=gr, color="orange")+
  geom_text_repel(box.padding=0.5, min.segment.length = 0, max.overlaps = 10,  nudge_x = 0.3, direction = "x", hjust = "right", aes(label=motif_id ), data= gr)+
  geom_point(data=stat1, color="blue")+
  geom_text_repel(box.padding=0.5, min.segment.length = 0, max.overlaps = 10,  nudge_y = 0.3, direction = "y", hjust = "right", aes(label=motif_id ), data= stat1)+
  #geom_point(color="orange", data=m.a5495) +
  #geom_point(color="blue", data=m.sperm5) +
  labs(color="Top enriched motifs")+
  labs(x="log2(fold-enrichment in spermatozoa)", y="log2(fold-enrichment in A549)", size="% of set") +
  #ylab("-log10 (adjusted p-value A549)")+
  #xlab("-log10 (adjusted p-value spermatozoa)")+
  theme_classic() +
  #theme(legend.position = "none")+
  ggtitle("Top ranked motifs in peaks of spermatozoa vs. A549")

ggsave("./plots_svg/motif_enrichment_sperm_vs_A549_peaks.svg")
ggsave("./plots_pdf/motif_enrichment_sperm_vs_A549_peaks.pdf")#, width = 6, height = 4)
ggsave("./plots_png/motif_enrichment_sperm_vs_A549_peaks.png")

```

# Fig S1C - Sperm GRE peaks not accessible in immature germ cells across spermatogenesis 

```{r}
tracks <- c(
  uds1="extern_data/ATACse_PRJNA399533_Thy1posSpermatogonia_GSM2751121.bw",
  uds2="extern_data/ATACse_PRJNA399533_Thy1posSpermatogonia_GSM2751122.bw",
  ds1="extern_data/ATACse_PRJNA399533_cKitposSpermatogonia_GSM2751125.bw",
  ds2="extern_data/ATACse_PRJNA399533_cKitposSpermatogonia_GSM2751126.bw",
  ps1="extern_data/ATACse_PRJNA399533_PachytenSpermatocytes_GSM2751129.bw",
  ps2="extern_data/ATACse_PRJNA399533_PachytenSpermatocytes_GSM2751130.bw",
  rs1="extern_data/ATACse_PRJNA399533_RoundSpermatides_GSM2751133.bw",
  rs2="extern_data/ATACse_PRJNA399533_RoundSpermatides_GSM2751134.bw",
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data/ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  GVoocyte1="extern_data/ATAC_PRJNA480448_GVOocyte_GSM3262861.bw",
  GVoocyte2="extern_data/ATAC_PRJNA480448_GVOocyte_GSM3262862.bw",
  "MII oocyte"="extern_data/ATAC_PRJNA480448_MIIOocyte_GSM3262863.bw"
)
```

```{r}
peaks <- readRDS("./objects/narrow_overlaps_spermGRE_not_immatureGermCellsGRE.rds")
m <- signal2Matrix(tracks, peaks, w = 20, BPPARAM = 2)
m <- epiwraps:::.ese2ml(m)

fused_uds_mat <- mergeSignalMatrices(m[1:2])
fused_ds_mat <- mergeSignalMatrices(m[3:4])
fused_ps_mat <- mergeSignalMatrices(m[5:6])
fused_rs_mat <- mergeSignalMatrices(m[7:8])
fused_sperm_mat <- mergeSignalMatrices(m[9:12])
fused_gv_mat <- mergeSignalMatrices(m[13:14])

fused_m <- c(list(fused_uds=fused_uds_mat), list(fused_ds=fused_ds_mat), list(fused_ps=fused_ps_mat), list(fused_rs=fused_rs_mat), list(fused_sperm=fused_sperm_mat), list(fused_GVoocytes=fused_gv_mat), m[15], m[-1:-15])
fused_n <- renormalizeBorders(lapply(fused_m,log1p))

#fused_m <- readRDS("rds_saves/paper/spermatogenesis_notimmature_fused_m.rds")
#fused_n <- readRDS("rds_saves/paper/spermatogenesis_notimmature_fused_n.rds")
```

```{r}
pdf("plots_pdf/GREpeaks_from_sperm_notimmature_across_spermatogenesis.pdf", width=17.5, height=10)
h <- plotEnrichedHeatmaps(fused_m, row_title="Peaks from spermatozoa with GRE \n not accessible in immature germ cells", raster_resize_mat = FALSE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()
```

```{r}
png("plots_png/GREpeaks_from_sperm_notimmature_across_spermatogenesis.png", height=768)
h <- plotEnrichedHeatmaps(fused_m, row_title="Peaks from spermatozoa with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()
```
# Fig S1D - Sperm GRE peaks across spermatogenesis hPTMs

Combined plot normalized
```{r activePTM-input, results='hide', fig.height=5, fig.width=4.5, fig.cap="Accessibility and ChIP signal of male germ cells at sperm peak regions"}
tracks <- c(
  H3K4me3="extern_data//SRR5049005.bw",#sperm4
  H3K27ac="extern_data/SRR3225845.bw",#sperm1
  H3K9ac="extern_data/SRR3225847.bw" #sperm1
)

peaks <- readRDS("objects/narrow_overlaps_spermGRE_not_immatureGermCellsGRE.rds")
m <- signal2Matrix(tracks, peaks, w=20, BPPARAM = 4)
m <- epiwraps:::.ese2ml(m)

#norm_factor <- bwNormFactors(x = tracks)
#m <- rescaleSignalMatrices(m,scaleFactors = norm_factor)
n <- renormalizeBorders(lapply(m,log1p))

#m <- readRDS("rds_saves/paper/hPTMs_spermNotImmature_m.rds")
#n <- readRDS("rds_saves/paper/hPTMs_spermNotImmature_n.rds")
```



Single panels of heatmaps
```{r}
# Create a list of heatmaps
for (i in 1:length(n)) {
  h <- plotEnrichedHeatmaps(n[i], raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = magma(2000), mean_color = "#800080",
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))
  
  # Save each heatmap to a PDF
    pdf(paste0("plots_pdf/GREpeaks_from_sperm_notImmature_hPTMs_normalized_", i, ".pdf"), width=2.5, height=10)
  draw(h, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
}
```

```{r}
# Create a list of heatmaps
for (i in 1:length(n)) {
  h <- plotEnrichedHeatmaps(n[i], raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = magma(2000), mean_color = "#800080",
                            heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4, "npc"), title=NULL))

  # Save each heatmap to a PNG
  png(paste0("plots_png/GREpeaks_from_sperm_notImmature_hPTMs_normalized_", i, ".png"), width=250, height=768)
  draw(h, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
}
```



# Fig  S1E - Promoters with GRE across sperm ATAC and hPTMs

```{r}

tracks <- c(
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data/ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  sperm1_H3K4me3="extern_data/SRR3309386.bw",
  sperm2_H3K4me3="extern_data/SRR3309387.bw",
  sperm3_H3K4me3="extern_data/SRR3703039.bw",
  sperm4_H3K4me3="extern_data/SRR5049005.bw",
  sperm5_H3K4me3="extern_data/SRR3225849.bw",
  sperm1_H3K27ac="extern_data/SRR3225845.bw",
  sperm2_H3K27ac="extern_data/SRR5049001.bw",
  sperm1_H3K9ac="extern_data/SRR3225847.bw",
  sperm2_H3K9ac="extern_data/SRR5049003.bw",
  sperm1_H3K4me1="extern_data/SRR3225848.bw",
  sperm2_H3K4me1="extern_data/SRR5049004.bw",
  sperm1_H3.3="extern_data/ChIP_PRJNA480447_Sperm_GSM3262880.bw",  
  sperm2_H3.3="extern_data/ChIP_PRJNA480447_Sperm_GSM3262881.bw",
  sperm1_H2A.Z="extern_data/ChIP_PRJNA480447_Sperm_GSM3262882.bw",
  sperm2_H2A.Z="extern_data/ChIP_PRJNA480447_Sperm_GSM3262880.bw"
)



proms <- readRDS("objects/GRE_promoters_center.rds")

m <- signal2Matrix(tracks, proms, w = 20, BPPARAM = 8)


fused_sperm <- mergeSignalMatrices(m[1:4])
fused_H3K4me3 <- mergeSignalMatrices(m[5:9])
fused_H3K27ac <- mergeSignalMatrices(m[10:11])
fused_H3K9ac <- mergeSignalMatrices(m[12:13])
fused_H3K4me1 <- mergeSignalMatrices(m[14:15])
fused_H3.3 <- mergeSignalMatrices(m[16:17])
fused_H2A.Z <- mergeSignalMatrices(m[18:19])

fused_m <- c(list(fused_sperm=fused_sperm), list(fused_H3K4me3=fused_H3K4me3), list(fused_H3K27ac=fused_H3K27ac), list(fused_H3K9ac=fused_H3K9ac), list(fused_H3K4me1=fused_H3K4me1), list(fused_H3.3=fused_H3.3), list(fused_H2A.Z=fused_H2A.Z), m[-1:-19])
fused_n <- renormalizeBorders(lapply(fused_m,log1p))

fused_m1 <- fused_m[1:3]
```




```{r}
pdf("plots_pdf/GREpromoters.pdf", width=7, height=10)
h <- plotEnrichedHeatmaps(fused_m1, row_title="Promoter regions with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()

png("plots_png/GREpromoters.png", height=768)
h <- plotEnrichedHeatmaps(fused_m1, row_title="Promoter regions with GRE", raster_resize_mat = TRUE, axis_name = c("-2kb","0","+2kb"), colors = rocket(2000), mean_color = "#800080")
print(h)
dev.off()
```


# Fig S1F - Heatmap for ICR in mm10
```{r}
ICR <- read.csv("objects/ICR_Weber_mm10.csv",header = T)
ICR$seqnames <- ICR$X.chrom
ICR <- GRanges(ICR)

bw_mm10 <- list.files(path = "intern_data/Nanopore/",pattern = "_5mC_4reads|_5mC_(.*_noSex)",full.names = T)
names(bw_mm10) <- c("Ctrl1","Ctrl2","KO1","KO2")
```

```{r}
Met_bed <- sapply(names(bw_mm10),function(names){
  import.bw(bw_mm10[names])
  
  })
```




```{r}
sm <- signal2Matrix(filepaths = Met_bed,regions = ICR,w =10,type = "scaled",smooth = T,limit = c(0,100))
plotEnrichedHeatmaps(sm)
as.data.frame(ICR)
set.seed(123)
cl <- epiwraps::clusterSignalMatrices(sm,k = 2)
cl2 <- as.character(cl)

cl2[cl2 == 1] <- ("Maternal")

cl2[cl2 == 2] <- ("Paternal")


cl2 <- as.factor(cl2)

```

```{r}
sm_merge_Ctrl <- mergeSignalMatrices(sm[,1:2])


```

```{r}
mycolors2 <- c("Maternal" = "firebrick","Paternal"= "steelblue")
p <- plotEnrichedHeatmaps(list("ICR" = sm_merge_Ctrl),row_split = cl2,scale_title = "5mC[%]",mean_color  = mycolors2, use_raster = T,heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28,"npc"), title = NULL))
```

```{r}
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")

  png(paste0("plots_png/FigS1F.png"), width=250, height=500)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/FigS1F.pdf"), width=3.5, height=7.5)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/FigS1F.svg"), width=3.5, height=7.5)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```

# Fig S1G - Display of methlyation score distribution
```{r}
set.seed(1234)
Met_bed_Ctrl <- Met_bed[1:2]
df <- rbind(as.data.frame(sample(Met_bed_Ctrl$Ctrl1$score,size = 100000)),y = as.data.frame(sample(Met_bed_Ctrl$Ctrl1$score,size = 100000))) 
df$score <- df$`sample(Met_bed_Ctrl$Ctrl1$score, size = 1e+05)`
df$Sample <- rep("S1",nrow(df))
df$Sample[((nrow(df)/2)+1):nrow(df)] <- "S2" 
```
```{r}
ggplot(data = df,aes(x = score,group = Sample)) +
  #geom_histogram(aes(y = ..density..), bins = 30, fill = NULL, alpha = 0.5) +
  geom_density(color = "orange2", size = 1,fill = "orange2",alpha = 0.5) +
  labs(title = "5mC Score Distribution in mature sperm", x = "5mC score [%]", y = "Density") +
  theme_bw()+theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
ggsave("plots_svg/FigS1G_5mC_score_distribution.svg")
ggsave("plots_pdf/FigS1G_5mC_score_distribution.pdf")
ggsave("plots_png/FigS1G_5mC_score_distribution.png")
```

# Fig S1H - Promoters with GRE at accesible site of ATAC H3k4me3 5mC

```{r}
proms <- readRDS("objects/GRE_promoters_center.rds")
```


```{r}
tracks <- c(
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data/ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  sperm1_H3K4me3="extern_data/SRR3309386.bw",
  sperm2_H3K4me3="extern_data/SRR3309387.bw",
  sperm3_H3K4me3="extern_data/SRR3703039.bw",
  sperm4_H3K4me3="extern_data/SRR5049005.bw",
  sperm5_H3K4me3="extern_data/SRR3225849.bw",
  sperm1_H3K27ac="extern_data/SRR3225845.bw",
  sperm2_H3K27ac="extern_data/SRR5049001.bw",
  sperm1_H3K9ac="extern_data/SRR3225847.bw",
  sperm2_H3K9ac="extern_data/SRR5049003.bw",
  sperm1_H3K4me1="extern_data/SRR3225848.bw",
  sperm2_H3K4me1="extern_data/SRR5049004.bw",
  sperm1_H3.3="extern_data/ChIP_PRJNA480447_Sperm_GSM3262880.bw",  
  sperm2_H3.3="extern_data/ChIP_PRJNA480447_Sperm_GSM3262881.bw",
  sperm1_H2A.Z="extern_data/ChIP_PRJNA480447_Sperm_GSM3262882.bw",
  sperm2_H2A.Z="extern_data/ChIP_PRJNA480447_Sperm_GSM3262880.bw"
)



```

```{r}
peaks <- readRDS("objects/narrow_peaks_with_GRE_sperm.rds")
```

```{r}
peaks2 <- peaks[overlapsAny(peaks,proms)]
seqnames(peaks2)
peaks2 <- keepStandardChromosomes(peaks2,pruning.mode = "coarse") 
peaks2 <- dropSeqlevels(peaks2,value = c("MT","Y","X"),pruning.mode = "coarse") 
```


```{r}
sm_5mC <- signal2Matrix(w = 20L , smooth = T, limit = c(0,100), filepaths = methyl_bed, regions = peaks2)
```

```{r}
plotEnrichedHeatmaps(sm_5mC)
```
```{r}
sm2 <- signal2Matrix(w = 20L, filepaths = tracks[c(1:4,5:9)],regions = peaks2)
plotEnrichedHeatmaps(sm2)
```
```{r}
plotEnrichedHeatmaps(c(epiwraps:::.ese2ml(sm2),epiwraps:::.ese2ml(sm_5mC)),multiScale = T,show_heatmap_legend = F)
```
```{r}
sm_merge <- mergeSignalMatrices(epiwraps:::.ese2ml(sm2)[1:4])
sm_merge_H3K4me3 <- mergeSignalMatrices(epiwraps:::.ese2ml(sm2)[5:9])
sm_merge_5mC <- mergeSignalMatrices(epiwraps:::.ese2ml(sm_5mC)[1:2])
```

```{r,fig.height=10}
plotEnrichedHeatmaps(ml = list(ATAC = sm_merge,H3K4me3 = sm_merge_H3K4me3,"5mC" = sm_merge_5mC),multiScale = T,row_title = "Accessible promoters with GRE")
```
```{r}
p <- plotEnrichedHeatmaps(ml = list(ATAC = sm_merge,H3K4me3 = sm_merge_H3K4me3),multiScale = T,row_title = "Accessible promoters with GRE",color = rocket(2000),scale_title = "density",axis_name = c("-2kb","TSS","+2kb"), use_raster = T,mean_scale_side = "none",  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))+
                       plotEnrichedHeatmaps(ml = list("5mC" = sm_merge_5mC),row_title = "Accessible promoters with GRE",use_raster =TRUE, scale_title = "5mc[%]",axis_name = c("-2kb","TSS","+2kb"),mean_scale_side = "none",  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL));p
# p
 #draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
 
 #draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")

  png(paste0("plots_png/FigS1H.png"), width=750, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/FigS1H.pdf"), width=7.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/FigS1H.svg"), width=7.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```
each heatmap on its own
```{r}
#ATAC
p <- plotEnrichedHeatmaps(ml = list(ATAC = sm_merge),multiScale = F,row_title = "Accessible promoters with GRE",color = rocket(2000),scale_title = "density",axis_name = c("-2kb","TSS","+2kb"), use_raster = T,  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
png(paste0("plots_png/FigS1HATAC.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/FigS1HATAC.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/FigS1HATAC.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
# H*K4m3 
  p <- plotEnrichedHeatmaps(ml = list(H3K4me3 = sm_merge_H3K4me3),multiScale = F,row_title = "Accessible promoters with GRE",color = rocket(2000),scale_title = "density",axis_name = c("-2kb","TSS","+2kb"), use_raster = T,  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
png(paste0("plots_png/FigS1HH3K4me3.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/FigS1HH3K4me3.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/FigS1HH3K4me3.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
# 5mC
  
p <-   plotEnrichedHeatmaps(ml = list("5mC" = sm_merge_5mC),row_title = "Accessible promoters with GRE",use_raster =TRUE, scale_title = "5mc[%]",axis_name = c("-2kb","TSS","+2kb"),  heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.4,"npc"), title=NULL))
png(paste0("plots_png/FigS1H5mc.png"), width=250, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/FigS1H5mc.pdf"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/FigS1H5mc.svg"), width=2.5, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
```

# Fig 2B - Validation of GR-KO

```{r}
methyl_bw2 <- list.files(path = "intern_data/Nanopore/",pattern = "_..bw$",full.names = T)
```



```{r}
ensdb_mm10 <- ah[["AH89211"]]
```


```{r}
plotSignalTracks(files = list( "PGC-GR-KO" = methyl_bw2[3:4], Control = methyl_bw2[1:2]),overlay.alpha = 75,region = "18:39420652-39434571",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb_mm10,extend = c(0,0), fontface.main = 1,cex.main = 1,cex.id=0.9,cex.title = 0.7, ylim = c(0,0.1),) # , rotation.title = 0
grid.text(paste0("Nr3c1"), x = 0.5, y = 0.13, gp = gpar(fontsize = 10, fontface = "bold")) 
```

```{r}
png(paste0("plots_png/Fig2B.png"), width=750, height=512)
plotSignalTracks(files = list( "PGC-GR-KO" = methyl_bw2[3:4], Control = methyl_bw2[1:2]),overlay.alpha = 75,region = "18:39420652-39434571",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb_mm10,extend = c(0,0), fontface.main = 1.4,cex.main = 1,cex.id=0.9,cex.title = 0.9, ylim = c(0,0.1),) # , rotation.title = 0
grid.text(paste0("Nr3c1"), x = 0.5, y = 0.15, gp = gpar(fontsize = 15, fontface = "bold")) 
dev.off()

pdf(paste0("plots_pdf/Fig2B.pdf"), width=8, height=7,colormodel = "cymk")
plotSignalTracks(files = list( "PGC-GR-KO" = methyl_bw2[3:4], Control = methyl_bw2[1:2]),overlay.alpha = 75,region = "18:39420652-39434571",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb_mm10,extend = c(0,0), fontface.main = 1.4,cex.main = 1,cex.id=0.9,cex.title = 0.9, ylim = c(0,0.1),) # , rotation.title = 0
grid.text(paste0("Nr3c1"), x = 0.5, y = 0.15, gp = gpar(fontsize = 15, fontface = "bold")) 
dev.off()


svg(paste0("plots_svg/Fig2B.svg"), width=8, height=7)
plotSignalTracks(files = list( "PGC-GR-KO" = methyl_bw2[3:4], Control = methyl_bw2[1:2]),overlay.alpha = 75,region = "18:39420652-39434571",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb_mm10,extend = c(0,0), fontface.main = 1.4,cex.main = 1,cex.id=0.9,cex.title = 0.9, ylim = c(0,0.1),) # , rotation.title = 0
grid.text(paste0("Nr3c1"), x = 0.5, y = 0.15, gp = gpar(fontsize = 15, fontface = "bold")) 
dev.off()

```


# Fig 2C DMRs PGC-GR-KO sperm heatmap
regions of interest
```{r}
HyperDMR <- readRDS("objects/hyper_mm39.rds") 
HypoDMR <- readRDS("objects/hypo_mm39.rds")

DMR <- unlist(GRangesList(HyperDMR,HypoDMR))


```




```{r}
i <- integer(length(DMR))
i[1:(length(HyperDMR))] <- "HyperDMR"
i[((length(HyperDMR))+1):length(DMR)] <- "HypoDMR"

```

tracks
```{r}
#NanoporeTracks
methyl_bw <- list.files(path = "intern_data/Nanopore//",pattern = "39.bw",full.names = T)
names(methyl_bw) <- c("Control 1","Control 2","PGC-GR-KO 1","PGC-GR-KO 2")
```



```{r}
methyl_bed <- sapply(names(methyl_bw),function(names){
  import.bw(methyl_bw[names])
  
  })
```




initial heatmap of DMRs smooth
```{r}
sm1 <- signal2Matrix(filepaths = methyl_bed,w = 20,regions = DMR,smooth = T,limit = c(0,100))
```


```{r,fig.width=10}
mycolors <- c("HyperDMR"="#ffa500ff", "HypoDMR"= "#c0c0ffff")
p <- plotEnrichedHeatmaps((sm1),row_split = i,scale_rows = F,mean_color = mycolors,mean_scale_side = "left", top_annotation = TRUE,scale_title = "5mC [%]",axis_name = c("-2kb","0","+2kb"),row_title = "DMRs")

p
```

```{r}
mycolors <- c("HyperDMR"="#ffa500ff", "HypoDMR"= "#c0c0ffff")
p <- plotEnrichedHeatmaps((sm1),row_split = i,scale_rows = F,mean_color = mycolors,mean_scale_side = "left", top_annotation = TRUE,scale_title = "5mC [%]",axis_name = c("-2kb","0","+2kb"),row_title = "DMRs", heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL))

p
```


```{r}


  png(paste0("plots_png/Fig2C.png"), width=1000, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/Fig2C.pdf"), width=10, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig2C.svg"), width=10, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```



# Fig 2D DMRs overlapping with GRE in promoter heatmap

```{r}
ensdb <- ah[["AH89457"]] # GRCm39

promoters <- promoters(ensdb)
```
we use fimo to get GRE in all promoters
```{r}
genome <- Rsamtools::FaFile("objects/Mus_musculus.GRCm39.dna_sm.primary_assembly.fa")
#promoters <- keepStandardChromosomes(promoters,pruning.mode = "coarse")
#promoters <- dropSeqlevels(x = promoters,value = "MT",pruning.mode = "coarse")
#promoter_seq <- get_sequence(regions = keepStandardChromosomes(promoters,pruning.mode = "coarse"),genome = genome)
#GRE_promoters <- memes::runFimo(sequences = promoter_seq,motifs = motifs$GCR,meme_path = "/common/meme/bin/")
#GRE_promoters <- reduce(GRE_promoters)
#saveRDS(GRE_promoters,"reduced_GRE_promoters_mm39.rds")
GRE_promoters <- readRDS("objects/reduced_GRE_promoters_mm39.rds")
```



```{r}
DMR_GRE_prom <- DMR[overlapsAny(DMR,GRE_promoters)]

DMR_GRE_up <- HyperDMR[overlapsAny(HyperDMR,GRE_promoters)]

DMR_GRE_down <- HypoDMR[overlapsAny(HypoDMR,GRE_promoters)]


DMR_GRE_both <- unlist(GRangesList(DMR_GRE_up,DMR_GRE_down))
```



initial heatmap of DMRs smooth
```{r}

sm1 <- signal2Matrix(filepaths = methyl_bed,w = 20,regions = DMR_GRE_both,smooth = T,limit = c(0,100))
```



```{r}
i <- integer(length(DMR_GRE_both))
i[1:(length(DMR_GRE_up))] <- "HyperDMR"
i[((length(DMR_GRE_up))+1):length(DMR_GRE_both)] <- "HypoDMR"

```





```{r,fig.width=10}

mycolors <- c("HyperDMR"="#ffa500ff", "HypoDMR"= "#c0c0ffff")
p <- plotEnrichedHeatmaps((sm1),row_split = i,mean_color = mycolors,scale_rows = F, mean_scale_side = "left", top_annotation = TRUE,scale_title = "5mC [%]",row_title = "DMRs overlapping GRE-sites in Promoters",axis_name = c("-2kb","GRE","+2kb"))

p
```

```{r}
mycolors <- c("HyperDMR"="#ffa500ff", "HypoDMR"= "#c0c0ffff")
p <- plotEnrichedHeatmaps((sm1),row_split = i,mean_color = mycolors,scale_rows = F, mean_scale_side = "left", top_annotation = TRUE,scale_title = "5mC [%]",row_title = "DMRs overlapping GRE-sites in Promoters", axis_name = c("-2kb","GRE","+2kb") ,heatmap_legend_param = list(legend_direction = "horizontal", legend_width = unit(0.28, "npc"), title=NULL),use_raster = T,trim = c(0.0001,0.9999))

p
```


```{r}


  png(paste0("plots_png/Fig2D.png"), width=1000, height=768)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  pdf(paste0("plots_pdf/Fig2D.pdf"), width=10, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
  
  svg(paste0("plots_svg/Fig2D.svg"), width=10, height=10)
  draw(p, heatmap_legend_side="bottom", annotation_legend_side="right",  legend_grouping = "original")
  dev.off()
```


# Fig 2E - DEG heatmap PGC-GR-KO embryos

```{r}
# Load the data
se <- readRDS("objects/SEs_for_SEV/VF_2cell_embyro_early_paternal_GR_KO(Alpl).SE.rds")

# Extract assays and row data
assays(se)
rowData(se)
dea <- rowData(se)[[2]]
dea

# Subset differentially expressed genes with FDR < 0.05
deaFDR <- subset(dea, dea$FDR < 0.05)
deaFDR

# Define treatments and colors
treatments <- c("ctrl", "Early_GR_KO") # Specify the order explicitly
colors <- c("#DEDFE4", "darkgrey") # Adjust colors as needed

# Ensure the number of colors matches the number of treatments
if (length(colors) != length(treatments)) {
  stop("Number of colors must match number of treatments")
}

# Create a named vector for color mapping
color_mapping <- setNames(colors, treatments)

# Reorder columns in the se object
se <- se[, order(factor(colData(se)$condition, levels = treatments))]

# Add color_mapping to the heatmap annotation
ha = HeatmapAnnotation(df = data.frame(condition = factor(colData(se)$condition, levels = treatments)),
                       col = list(condition = color_mapping))

# Define color palette for the heatmap
color_palette <- colorRampPalette(c("blue", "black", "orange"))(100)

pdf("plots_pdf/DEGs_VF-Alpl-KO_heatmap.pdf", height=4) #, width=14.5, height=10)
x <- sechm(se, rownames(deaFDR), assayName = "log2FC", do.scale = FALSE, hmcols = color_palette, top_annotation = ha, row_title = "Differentially expressed genes", gaps_at="condition", use_raster = TRUE)
print(x)
dev.off()

png("plots_png/DEGs_VF-Alpl-KO_PROTAC_heatmap.png", height=768)
x <- sechm(se, rownames(deaFDR), assayName = "log2FC", do.scale = FALSE, hmcols = color_palette, top_annotation = ha, row_title = "Differentially expressed genes", gaps_at="condition", use_raster = TRUE)
print(x)
dev.off()

svg("plots_svg/DEGs_VF-Alpl-KO_heatmap.svg") #, width=14.5, height=10)
x <- sechm(se, rownames(deaFDR), assayName = "log2FC", do.scale = FALSE, hmcols = color_palette, top_annotation = ha, row_title = "Differentially expressed genes", gaps_at="condition", use_raster = TRUE)
print(x)
dev.off()

sechm(se, rownames(deaFDR), assayName = "log2FC", do.scale = FALSE, hmcols = color_palette, top_annotation = ha, row_title = "Differentially expressed genes", gaps_at="condition", use_raster = TRUE) 
```


#Fig 2F - volcano DEGs PGC-GR-KO


```{r}
library(scales)
se <- readRDS("objects/VF_2cell_embyro_early_paternal_GR_KO(Alpl).SE.rds")

# Extract assays and row data
assays(se)
rowData(se)
dea <- rowData(se)[[2]]
dea$Gene <- dea$ID

# Subset differentially expressed genes with FDR < 0.05
deaFDR <- subset(dea, dea$FDR < 0.005)
deaFDR

xlim <- c(-7, 7)
ylim <- c(0,6)

# Function to generate volcano plots
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$FDR < 0.01
  df$extreme[df$logFC > xlim[2] | df$logFC < xlim[1] | -log10(df$FDR) > ylim[2] | -log10(df$FDR) < ylim[1]] <- "extreme"
  df$highlight[df$extreme == "extreme"] <- "extreme"
  top_genes <- df[order(df$FDR), ][1:25, ]
  ggplot(df, aes(x=logFC, y=-log10(FDR))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black", "extreme" = "firebrick3")) +
    geom_text_repel(data = subset(top_genes, highlight == TRUE | highlight == "extreme"), aes(label=Gene), size=3.5, min.segment.length = 0, max.overlaps = Inf, nudge_y = 0.1) +
    geom_hline(yintercept=-log10(0.05), linetype="dashed")+
    xlab("logFC") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    scale_x_continuous(limits = xlim, oob = squish) +
    scale_y_continuous(limits = ylim, oob = squish) +
    ggtitle(title)
}

# Apply the function to each dataframe in the list
generate_volcano_plot(dea, "Differentially expressed genes in PGC-GR-KO 2-cell embryos")
ggsave("plots_svg/Fig2F_DEGs_VF-KO_volcano.svg")
ggsave("plots_pdf/Fig2F_DEGs_VF-KO_volcano.pdf")
ggsave("plots_png/Fig2F_DEGs_VF-KO_volcano.png")
```



# Fig 2G 5mC at Mtus1

```{r}
ensdb <- ah[["AH89457"]]
```


```{r}
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "8: 41494253-41495035",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1,cex.main = 1,cex.id=0.9,cex.title = 0.7) # , rotation.title = 0
grid.text(paste0("DMR: Mtus1/Gm16192"), x = 0.5, y = 0.95, gp = gpar(fontsize = 10, fontface = "bold"))
```

```{r}
png(paste0("plots_png/Fig2G.png"), width=750, height=512)
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "8: 41494253-41495035",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Mtus1/Gm16192"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold")) 
dev.off()

pdf(paste0("plots_pdf/Fig2G.pdf"), width=8, height=7,colormodel = "cymk")
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "8: 41494253-41495035",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Mtus1/Gm16192"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold"))
dev.off()


svg(paste0("plots_svg/Fig2G.svg"), width=8, height=7)
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "8: 41494253-41495035",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Mtus1/Gm16192"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold"))
dev.off()

```



#Fig 2H 5mC at Promoter of embryo DEA

```{r}
dea <- readRDS("objects/DEA_e2c.rds")
annGR <- readRDS("objects/annGR.rds")
```



```{r}
set.seed(1234)
annGR <- readRDS("objects/annGR.rds")
annGR2 <-keepStandardChromosomes(annGR[annGR$gene_name %in% dea$ID,],pruning.mode = "coarse"); annGR <- dropSeqlevels(annGR,value = c("chrMT","X","Y"),pruning.mode = "coarse") 
DEA_up <- annGR[annGR$gene_name %in% dea$ID[(dea$FDR < 0.05) & (dea$logFC < 0)] ,]
DEA_up <- keepStandardChromosomes(DEA_up,pruning.mode = "coarse"); DEA_up <- dropSeqlevels(DEA_up,value = c("chrMT","X","Y"),pruning.mode = "coarse")
DEA_up <- reduce(DEA_up)



DEA_down <- annGR[annGR$gene_name %in% dea$ID[(dea$FDR < 0.05) & (dea$logFC > 0)] ,]
DEA_down <- keepStandardChromosomes(DEA_down,pruning.mode = "coarse"); DEA_down <- dropSeqlevels(DEA_down,value = c("chrMT","X","Y"),pruning.mode = "coarse")  
DEA_down <- reduce(DEA_down)
DEA_down <- sample(DEA_down,size = length(DEA_up))


DEA <- unlist(GRangesList(DEA_up,DEA_down))



# we remove random promoters which have now overlap with any methylData
annGR2 <- annGR2[overlapsAny(annGR2,unlist(GRangesList(methyl_bed)))]

randDEA <- sample(reduce(annGR2[!overlapsAny(annGR2,DEA),]),length(DEA_up))

DEA2 <- unlist(GRangesList(DEA_up,DEA_down,randDEA))

DEA <- resize(DEA,width = 2,fix = "start")
DEA2 <- resize(DEA2,width = 2,fix = "start")


# sanity check
#proms <- promoters(ensdb,upstream = 5,downstream = 5)

#overlapsAny(DEA2,proms)
```

```{r}
k <- integer(length(DEA2))
k[1:(length(DEA_up))] <- "PGC-GR-KO\nearlyEmbryo UP"
k[((length(DEA_up))+1):(length(DEA_up)+length(DEA_down))] <- "PGC-GR-KO\nearlyEmbryo DOWN"
k[(((length(DEA_up)+length(DEA_down)))+1):length(DEA2)] <- "randomProms"
```


```{r}
i <- integer(length(DEA))
i[1:(length(DEA_up))] <- "PGC-GR-KO\nearlyEmbryo UP"
i[((length(DEA_up))+1):length(DEA)] <- "PGC-GR-KO\nearlyEmbryo DOWN"

```

```{r}
sm1 <- signal2Matrix(filepaths = methyl_bed,w = 1,regions = DEA2,smooth = T,limit = c(0,100),extend = c(500,1500))

plotEnrichedHeatmaps(sm1)
```
```{r}
melt <- meltSignals(sm1,splitBy = k)
melt$Group <- melt$split
```
```{r}
theme_set(theme_bw())
ggplot(melt, aes(position, mean, colour=Group, group = Group, fill = Group)) +
  geom_vline(xintercept=0, linetype="dashed") +
  labs(x="TSS", y="mean 5mC [%]")+
  stat_summary(geom="ribbon", alpha=0.3, colour = NA ) + stat_summary(fun=median, geom="line", size = 0.6)+
  ggtitle("5mC levels of sperm at\npromoter regions of DEAs in the PGC-GR-KO embryo")+   
  scale_color_manual(values = c("#c0c0ffff","#ffa500ff","forestgreen")) +
  scale_fill_manual(values = c("#c0c0ffff","#ffa500ff","forestgreen")) +
  theme(text = element_text(size = 10), axis.title.x = element_text(hjust = 0.26))





```
```{r}
theme_set(theme_bw())
ggplot(melt, aes(position, mean, colour=Group, group = Group, fill = Group)) +
  geom_vline(xintercept=0, linetype="dashed") +
  labs(x="TSS", y="mean 5mC [%]")+
  stat_summary(geom="ribbon", alpha=0.3, colour = NA ) + stat_summary_bin(fun=mean, geom="line", size = 0.6)+
  ggtitle("5mC levels of sperm at\npromoter regions of DEAs in the PGC-GR-KO embryo")+   
  scale_color_manual(values = c("#c0c0ffff","#ffa500ff","forestgreen")) +
  scale_fill_manual(values = c("#c0c0ffff","#ffa500ff","forestgreen")) +
  theme(text = element_text(size = 10), axis.title.x = element_text(hjust = 0.26))


```



```{r}
melt[melt$split == "randomProms",]
```

## AUC
here we use all genes e2c up and down


```{r}
set.seed(1234)
annGR <- readRDS("objects/annGR.rds")
annGR2 <-keepStandardChromosomes(annGR[annGR$gene_name %in% dea$ID,],pruning.mode = "coarse"); annGR <- dropSeqlevels(annGR,value = c("chrMT","X","Y"),pruning.mode = "coarse") 
DEA_up <- annGR[annGR$gene_name %in% dea$ID[(dea$FDR < 0.05) & (dea$logFC < 0)] ,]
DEA_up <- keepStandardChromosomes(DEA_up,pruning.mode = "coarse"); DEA_up <- dropSeqlevels(DEA_up,value = c("chrMT","X","Y"),pruning.mode = "coarse")
DEA_up <- reduce(DEA_up)



DEA_down <- annGR[annGR$gene_name %in% dea$ID[(dea$FDR < 0.05) & (dea$logFC > 0)] ,]
DEA_down <- keepStandardChromosomes(DEA_down,pruning.mode = "coarse"); DEA_down <- dropSeqlevels(DEA_down,value = c("chrMT","X","Y"),pruning.mode = "coarse")  
DEA_down <- reduce(DEA_down)
#DEA_down <- sample(DEA_down,size = length(DEA_up))


DEA <- unlist(GRangesList(DEA_up,DEA_down))



# we remove random promoters which have now overlap with any methylData
annGR2 <- annGR2[overlapsAny(annGR2,unlist(GRangesList(methyl_bed)))]

randDEA <- sample(reduce(annGR2[!overlapsAny(annGR2,DEA),]),length(DEA_up))

DEA2 <- unlist(GRangesList(DEA_up,DEA_down,randDEA))

DEA <- resize(DEA,width = 2,fix = "start")
DEA2 <- resize(DEA2,width = 2,fix = "start")


# sanity check
#proms <- promoters(ensdb,upstream = 5,downstream = 5)

#overlapsAny(DEA2,proms)
```

```{r}
k <- integer(length(DEA2))
k[1:(length(DEA_up))] <- "PGC-GR-KO\nearlyEmbryo UP"
k[((length(DEA_up))+1):(length(DEA_up)+length(DEA_down))] <- "PGC-GR-KO\nearlyEmbryo DOWN"
k[(((length(DEA_up)+length(DEA_down)))+1):length(DEA2)] <- "randomProms"
```


```{r}
i <- integer(length(DEA))
i[1:(length(DEA_up))] <- "PGC-GR-KO\nearlyEmbryo UP"
i[((length(DEA_up))+1):length(DEA)] <- "PGC-GR-KO\nearlyEmbryo DOWN"

```

```{r}
sm1 <- signal2Matrix(filepaths = methyl_bed,w = 1,regions = DEA,smooth = T,limit = c(0,100),extend = c(500,1500))
# for proper calculations we ensure to remove regions were smoothing failed
sm1[is.na(as.data.frame(assay(sm1)))]
rows_with_na <- apply(as.data.frame(assay(sm1)), 1, function(row) any(is.na(row)))
sm1_clean <- sm1[!rows_with_na,]
i_clean <- i[!rows_with_na]
```


```{r}
plotEnrichedHeatmaps(sm1_clean)
assay(sm1_clean)
sm1_clean_merged <- mergeSignalMatrices(sm1_clean)  
df <- as.data.frame(sm1_clean_merged)
df <- df[,501:ncol(df)]
```
```{r}
AUC_per_region <- sapply(1:nrow(df),function(row){
  sm_auc(x =1:length(unlist(c(df[row,]))) , y =unlist(c(df[row,])))
})


mean(AUC_per_region[i_clean == "PGC-GR-KO\nearlyEmbryo DOWN"])/mean(AUC_per_region[i_clean == "PGC-GR-KO\nearlyEmbryo UP"])
median(AUC_per_region[i_clean == "PGC-GR-KO\nearlyEmbryo DOWN"])/median(AUC_per_region[i_clean == "PGC-GR-KO\nearlyEmbryo UP"])
```
```{r}
library(coin)
coin_wilcox <- coin::wilcox_test(AUC_per_region ~as.factor(i_clean),distribution=approximate(nresample = 10000))
  pvalue_wilcox <- pvalue(coin_wilcox)
  sigma_wilcox <- statistic(coin_wilcox)  
```


```{r}
melt <- meltSignals(sm1_clean,splitBy = i_clean)
melt$Group <- melt$split

melt_UP <- melt[melt$split == "PGC-GR-KO\nearlyEmbryo UP",]
melt_DOWN <- melt[melt$split == "PGC-GR-KO\nearlyEmbryo DOWN",]
```

```{r}
mean_data_up <- melt_UP %>%
  group_by(position) %>%
  summarize(mean_y = mean(mean))


mean_data_down <- melt_DOWN %>%
  group_by(position) %>%
  summarize(mean_y = mean(mean))


x_min <- 0
x_max <- 1500



filtered_data_up <- mean_data_up %>%
  filter(position >= x_min & position <= x_max)
filtered_data_down <- mean_data_down %>%
  filter(position >= x_min & position <= x_max)

```

```{r}
library(smplot2)
up_auc <- sm_auc(filtered_data_up$position,filtered_data_up$mean_y)

down_auc <- sm_auc(filtered_data_down$position,filtered_data_down$mean_y)

down_auc/up_auc
```


# Fig S2C - DMRs with GR track
```{r}
HyperDMR <- readRDS("objects/hyper_mm39.rds") 
HypoDMR <- readRDS("objects/hypo_mm39.rds")

DMR <- unlist(GRangesList(HyperDMR,HypoDMR))


```
```{r}
bw_ChIP <- list.files("intern_data/ChIP/",pattern = "*subtract",full.names = T)
names(bw_ChIP)<- gsub(x = basename(bw_ChIP),pattern = ".subtract.bw",replacement = "")
bw_GR <- bw_ChIP[1]
names(bw_GR) <- "GR"
bed_GR <- import.bw(bw_GR) 
```



```{r}

  sm_GR_5fC <- signal2Matrix(regions = DMR,filepaths = list(GR = bed_GR),type = "scaled")  
```

```{r}
plotEnrichedHeatmaps(sm_GR_5fC,row_title = "DMR",scale_title = "GRdensity", color = mako(2000))
```

# Fig S2E - DEAs PGC-GR-KO early embryo RegionUpset
```{r}
df <- read.csv("objects/DEAFDR005-VF-KO.csv", sep = ";")
# Create a list of vectors, each vector represents a category
sperm_overlaps <- df$Gene[df$Overlap.sperm.ATAC == 1]
gre <- df$Gene[df$GRE.in.gene == 1]
motifs <- df$Gene[df$gene.body.motifs.connected.to.GR == 1]

gene_list <- list(DEGs = df$Gene, "DEGs accesible in Sperm" = sperm_overlaps, "DEGs with GRE motif" = gre, "DEGs with GR binding partner motif" = motifs) 
```

```{r}
png(paste0("plots_png/FigS2E.png"),width=8, height=7,res = 300,units = "in")
regionUpset2(gene_list)
dev.off()

pdf(paste0("plots_pdf/FigS2E.pdf"), width=8, height=7,colormodel = "cymk")
regionUpset2(gene_list)
dev.off()


svg(paste0("plots_svg/FigS2E.svg"), width=8, height=7)
regionUpset2(gene_list)
dev.off()

```

# Fig S2F - GSEA PGC-GR-KO

```{r}
deaGO <- function(dea, threshold=0.05, direction=NULL, ontology="BP"){
  isDE <- dea$adj.P.Val<threshold
  if(!is.null(direction)) isDE <- isDE & sign(dea$logFC)==sign(direction)
  isDE <- setNames(factor(as.integer(isDE)),row.names(dea))
  print(table(isDE))
  god <- new("topGOdata", ontology=ontology, allGenes=isDE, 
             nodeSize=10, annot=annFUN.org, 
             ID="symbol", mapping="org.Mm.eg.db")
  ea <- runTest(god, algorithm="classic", statistic="fisher")
  res <- GenTable(god, classicFisher=ea, topNodes=length(god@graph@nodes))
  res$Term2 <- plgINS::breakStrings(res$Term)
  res$log2Enrichment <- log2(res$Significant/res$Expected)
  res$FDR <- p.adjust(res$classicFisher)
  res
}

deaGSEA <- function(dea, gsets=NULL){
  if(is.null(gsets))
    gsets <- plgINS::getMsigSets("Mus musculus", collections=c("H","C5"))
  signature <- setNames(sign(dea$logFC)*-log10(dea$P.Value), row.names(dea))
  fg <- fgseaMultilevel(gsets, signature, minSize=10, nproc=3)
  fg$genes <- sapply(fg$leadingEdge, collapse=", ", FUN=paste)
  fg$leadingEdge <- NULL
  as.data.frame(fg[order(fg$pval),])
}
```

```{r}
se <- readRDS("objects/VF_2cell_embyro_early_paternal_GR_KO(Alpl).SE.rds")
deas <- getDEA(se)
dea <- deas$glmFit
dea$P.Value <- dea$PValue

dea2 <- dea %>% distinct(ID, .keep_all = TRUE)
rownames(dea2) <- dea2$ID

res2 <- deaGSEA(dea2)

ggplot(head(res2,10), aes(x = reorder(pathway, NES), y = NES, fill = -log10(padj))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "grey", high = "orange") +
  theme_minimal() +
  labs(
    title = "Enriched gene sets in \n PGC-GR-KO derived embryos",
    x = "Gene set",
    y = "Enrichment Score",
    fill = "-log10(FDR)"
  )

ggsave("./plots_svg/GSEA_KO.svg")
ggsave("./plots_pdf/GSEA_KO.pdf")
ggsave("./plots_png/GSEA_KO.png")
```

# Fig S2G - Gene expression of Nr3c1 in early embryo og PGC-GR-KO sperm

```{r}
se <- readRDS("objects/VF_2cell_embyro_early_paternal_GR_KO(Alpl).SE.rds")
se$condition <- as.character(se$condition)
se$condition[se$condition == "Early_GR_KO"] <- "PGC-GR-KO"
se$condition[se$condition == "Ctrl"] <- "Control"
se$condition <- as.factor(se$condition)
```

```{r}
df <- as.data.frame(cbind(t(assays(se)$logcpm),colData(se))) # we extract data from the SE object and combine it with the colData as a dataframe, this makes it more easy to be accessed by ggplot

colnames(df) <- gsub(x = colnames(df),pattern = "^EN(.*)\\.",replacement = "")

df <- df[,!duplicated(colnames(df))]
```

```{r}
p <- ggplot((df), aes(x = condition, y = Nr3c1, color = condition)) +
  geom_violin(trim = TRUE) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "Gene expression levels of Nr3c1",
    y = "logcpm",
    x =""
  )
ggsave("plots_svg/FigS2G_Violing_Nr3c1.svg")
ggsave("plots_pdf/FigS2G_Violing_Nr3c1.pdf")
ggsave("plots_png/FigS2G_Violing_Nr3c1.png")
```



# Fig S2E - Tracks for Rdx and Ptp4a2

```{r}
png(paste0("plots_png/Fig2G_Rdx.png"), width=750, height=512)
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "9:52011472-52012171",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Rdx"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold")) 
dev.off()

pdf(paste0("plots_pdf/Fig2G_Rdx.pdf"), width=8, height=7,colormodel = "cymk")
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "9:52011472-52012171",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Rdx"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold"))
dev.off()


svg(paste0("plots_svg/Fig2G_Rdx.svg"), width=8, height=7)
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "9:52011472-52012171",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Rdx"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold"))
dev.off()

```


```{r}
png(paste0("plots_png/Fig2G_Ptp4a2.png"), width=750, height=512)
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "4: 129711699-129711894",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Ptp4a2"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold")) 
dev.off()

pdf(paste0("plots_pdf/Fig2G_Ptp4a2.pdf"), width=8, height=7,colormodel = "cymk")
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "4: 129711699-129711894",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Ptp4a2"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold"))
dev.off()


svg(paste0("plots_svg/Fig2G_Ptp4a2.svg"), width=8, height=7)
plotSignalTracks(files = list("PGC-GR-KO"=methyl_bw[c(3,4)],Control =methyl_bw[c(1,2)]),overlay.alpha = 75,region = "4: 129711699-129711894",aggregation = "overlay",colors = c("orange","darkgrey"),genomeAxis = T,ensdb = ensdb,extend = c(1000,1000),ylim= c(0,100), fontface.main = 1.4,cex.main = 0.9,cex.id=0.9,cex.title = 0.9) # , rotation.title = 0
grid.text(paste0("DMR: Ptp4a2"), x = 0.5, y = 0.95, gp = gpar(fontsize = 15, fontface = "bold"))
dev.off()

```
# Fig 3A

```{r}
peak <- import("objects/NFpeaks/embryo_consensus.bed")
length(peak)

peak_seqs <- memes::get_sequence(peak, genome)
length(peak_seqs)

ame <- runAme(peak_seqs, database=convert_motifs(motifs), meme_path="/common/meme/bin") #https://meme-suite.org/meme/doc/ame-output-format.html 

ame[1:10]
```

```{r}
ame <- readRDS("objects/ame_embryo_consensus.rds")
ame[1:10]
gr <- ame[82, , drop=TRUE]
```


```{r}
ggplot(ame, aes(log2((1+tp)/(1+fp)), -log10(adj.pvalue))) + 
  geom_point(data=gr, aes(size=tp_percent), color="orange")+
  geom_text_repel(box.padding=0.5, min.segment.length = 0, max.overlaps = 10,  nudge_y = 0.3, direction = "y", hjust = "right", aes(label=motif_id ), data= gr)+
  geom_point(aes(size=tp_percent), alpha=0.3) + 
  geom_text_repel(box.padding=0.5, min.segment.length = 0, max.overlaps = 10,  nudge_y = 0.3, direction = "y", hjust = "right", aes(label=motif_id ), data= head(ame,10))+
  labs(x="log2(fold-enrichment)", size="% of set")+
  ylab("-log10 (adjusted p-value)")+
  xlab("Enrichment")+
  theme_classic() +
  ggtitle("Top enriched motifs in the early embryo")+
  guides(fill = guide_legend(override.aes = aes(color = NA)))

ggsave("./plots_svg/motif_enrichment_embryo_highconfidencepeaks.svg")
ggsave("./plots_pdf/motif_enrichment_embryo_highconfidencepeaks.pdf")
ggsave("./plots_png/motif_enrichment_embryo_highconfidencepeaks.png")
```




# Fig 3B - Clustering GRE peaks

```{r}
tracks <- c(
  sperm1="extern_data/ATAC_PRJNA480448_Sperm_GSM3262855.bw",
  sperm2="extern_data/ATAC_PRJNA480448_Sperm_GSM3262856.bw",
  sperm3="extern_data//ATAC_PRJNA315208_Sperm_GSM2088377.bw",
  sperm4="extern_data//ATAC_PRJNA315208_Sperm_GSM2088378.bw",
  MIIoocyte1="extern_data/ATAC_PRJNA480448_MIIOocyte_GSM3262863.bw",
  embryoearly2cell="extern_data/ATAC_PRJNA277362_Early2cell_merged.bw",
  embryo2cell="extern_data/ATAC_PRJNA277362_2cell_merged.bw",
  embryo4cell="extern_data/ATAC_PRJNA277362_4cell_merged.bw",
  ESC="extern_data/ATAC_PRJNA277362_ESC_GSM1625858.bw"
  )
```




```{r}
mycolors <- c("1"=brewer.pal(n=20,"Set1")[1], "2"=brewer.pal(n=20,"Set1")[2], "3"=brewer.pal(n=20,"Set1")[3], "4"=brewer.pal(n=20,"Set1")[4], "5"=brewer.pal(n=20,"Set1")[5], "6"=brewer.pal(n=20,"Set1")[6], "7"=brewer.pal(n=20,"Set1")[7], "8"=brewer.pal(n=20,"Set1")[8],
              "9"=brewer.pal(n=20,"Paired")[1], "10"=brewer.pal(n=20,"Paired")[2], "11"=brewer.pal(n=20,"Paired")[3], "12"=brewer.pal(n=20,"Paired")[4], "13"=brewer.pal(n=20,"Paired")[5], "14"=brewer.pal(n=20,"Paired")[6], "15"=brewer.pal(n=20,"Paired")[7], "16"=brewer.pal(n=20,"Paired")[8], "17"=brewer.pal(n=20,"Paired")[9], "18"=brewer.pal(n=20,"Paired")[10], "19"=brewer.pal(n=20,"Paired")[11])

set.seed(123)

peaks_emb <- readRDS("./objects/peaks_with_GRE_embryo_inAny.rds") 
peaks_sperm <- readRDS("./objects/narrow_peaks_with_GRE_sperm.rds")
peaks_oo <- readRDS("./objects/narrow_peaks_with_GRE_mIIoocytes.rds")

length(peaks_emb) #2084
length(peaks_sperm) #9311
length(peaks_oo) #792

peaks_emb <- reduce(peaks_emb, with.revmap=TRUE)
length(peaks_emb)
peaks_sperm <- reduce(peaks_sperm, with.revmap=TRUE)
length(peaks_sperm)
peaks_oo <- reduce(peaks_oo, with.revmap=TRUE)
length(peaks_oo)

peaklist <- unlist(GRangesList(peaks_emb, peaks_sperm, peaks_oo))
length(peaklist) #12117

m <- signal2Matrix(tracks, peaklist, w=5, type="center",  BPPARAM = 4)
m <- epiwraps:::.ese2ml(m)

fused_sperm_mat <- mergeSignalMatrices(m[1:4])
fused_m <- c(list(sperm=fused_sperm_mat), m[5:9], m[-1:-9])
fused_n <- renormalizeBorders(lapply(fused_m,log1p))

fused_m <- readRDS("objects/clustering_fused_m.rds")
fused_n <- readRDS("objects/clustering_fused_n.rds")

cl <- clusterSignalMatrices(fused_n, k=9) #% variance explained by clusters
ro <- order(rowMeans(fused_n$embryoearly2cell), decreasing = TRUE)
```

```{r}
pdf("plots_pdf/GREpeaks_clustering.pdf", width=14.5, height=10)
x <- plotEnrichedHeatmaps(fused_n, 
                     row_split = cl,
                     row_order = ro,
                     #trim = 0.99, 
                     raster_resize_mat = FALSE, 
                     mean_color = mycolors, 
                     scale_rows = FALSE,
                     axis_name = c("-2kb","0","+2kb")
                     )
print(x)
dev.off()



png("plots_pdf/GREpeaks_clustering.png", height=768)
x <- plotEnrichedHeatmaps(fused_n, 
                     row_split = cl,
                     row_order = ro,
                     #trim = 0.99, 
                     raster_resize_mat = FALSE, 
                     mean_color = mycolors, 
                     scale_rows = FALSE,
                     axis_name = c("-2kb","0","+2kb")
                     )
print(x)
dev.off()
```



# Fig 3C - 2-cell SmartSeq

```{r}
se <- readRDS("objects/MK_EXP20.1_ProtacEmbryo_merged.SE.rds")

# Generate the dataframes
dex <- rowData(se)[[4]]
protac <- rowData(se)[[3]]
pd <- rowData(se)[[5]]

protac$Gene <- rownames(protac)
dex$Gene <- rownames(dex)
pd$Gene <- rownames(pd)

# List of dataframes
df_list <- list(Dex = dex, "KH-103" = protac, "KH-103+Dex" = pd)

# Function to generate volcano plots
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$FDR < 0.1
  top_genes <- df[order(df$FDR), ][1:10, ]
  ggplot(df, aes(x=logFC, y=-log10(FDR))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black")) +
    geom_text_repel(data = subset(top_genes, highlight == TRUE), aes(label=Gene), size=3.5, min.segment.length = 0, max.overlaps = Inf, nudge_y = 0.1) +
    geom_hline(yintercept=-log10(0.05), linetype="dashed")+
    xlab("logFC") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    ggtitle(title)
}

# Apply the function to each dataframe in the list
plots <- lapply(names(df_list), function(name) {
  generate_volcano_plot(df_list[[name]], paste("Differentially expressed genes in 2-cell embryos upon ", name, " treatment"))
})

print(plots)

print(plots[1])
  ggsave("plots_svg//DEGs_EXP20-1_Dex_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-1_Dex_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-1_Dex_volcano.png")

print(plots[2])
ggsave("plots_svg/DEGs_EXP20-1_PROTAC_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-1_PROTAC_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-1_PROTAC_volcano.png")

print(plots[3])
ggsave("plots_svg/DEGs_EXP20-1_PD_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-1_PD_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-1_PD_volcano.png")
```

Heatmaps
```{r}
se <- readRDS("objects/MK_EXP20.1_ProtacEmbryo_merged.SE.rds")
se <- se[, se$treatment != "Dex"]
se <- se[, se$treatment != "PD"]

# Generate the dataframes
protac <- rowData(se)[[3]]
protac$Gene <- rownames(protac)
protacFDR <- subset(protac, protac$FDR <0.1)

treatments <- unique(se$treatment)

# Define own colors
colors <- c("grey", "orange" )#"800080", "grey", "orange", "black")  

# Make sure the number of colors matches the number of treatments
if (length(colors) != length(treatments)) {
  stop("Number of colors must match number of treatments")
}

# Create a named vector for color mapping
color_mapping <- setNames(colors, treatments)

# Add color_mapping to the heatmap annotation
ha = HeatmapAnnotation(df = data.frame(treatment = factor(colData(se)$treatment, levels = names(color_mapping))),
                       col = list(treatment = color_mapping))

sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes in 2-cell embryos", gaps_at="treatment")

pdf("plots_pdf/DEGs_EXP20-1_PROTAC_heatmap.pdf") #, width=14.5, height=10)
x <- sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes in 2-cell embryos", gaps_at="treatment")
print(x)
dev.off()

png("plots_png/DEGs_EXP20-1_PROTAC_heatmap.png", height=768)
x <- sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes in 2-cell embryos", gaps_at="treatment")
print(x)
dev.off()
#needs title: KH-103 effects on gene regulation in 2-cell embryos
```


# Fig 3D - 8-cell SmartSeq

```{r}
se <- readRDS("objects/MK_EXP20.2_ProtacEmbryo8cells.SE.rds")

# Generate the dataframes
dex <- rowData(se)[[2]]
protac <- rowData(se)[[3]]
pd <- rowData(se)[[4]]

protac$Gene <- rownames(protac)
dex$Gene <- rownames(dex)
pd$Gene <- rownames(pd)

# List of dataframes
df_list <- list(Dex = dex, "KH-103" = protac, "KH-103+Dex" = pd)

# Function for volcano plots
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$adj.P.Val < 0.1
  top_genes <- df[order(df$adj.P.Val), ][1:10, ]
  ggplot(df, aes(x=logFC, y=-log10(adj.P.Val))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black")) +
    geom_text_repel(data = subset(top_genes, highlight == TRUE), aes(label=Gene), size=3.5, min.segment.length = 0, max.overlaps = Inf, nudge_y = 1) +
    geom_hline(yintercept=-log10(0.05), linetype="dashed")+
    xlab("logFC") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    ggtitle(title)
}


# Apply the function to each dataframe in the list
plots <- lapply(names(df_list), function(name) {
  generate_volcano_plot(df_list[[name]], paste("Differentially expressed genes in 8-cell embryos upon ", name, " treatment"))
})
print(plots[1])
ggsave("plots_svg/DEGs_EXP20-2_Dex_volcano.svg")
  ggsave("plots_pdf/DEGs_EXP20-2_Dex_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-2_Dex_volcano.png")

print(plots[2])
ggsave("plots_svg/DEGs_EXP20-2_PROTAC_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-2_PROTAC_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-2_PROTAC_volcano.png")

print(plots[3])
ggsave("plots_svg/DEGs_EXP20-2_PD_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-2_PD_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-2_PD_volcano.png")
```

heatmaps
```{r}
se <- readRDS("objects/MK_EXP20.2_ProtacEmbryo8cells.SE.rds")
se <- se[, se$treatment != "DEXR"]
se <- se[, se$treatment != "DEX"]
se <- se[, se$treatment != "PDR"]
se <- se[, se$treatment != "PD"]

# Generate the dataframes
protac <- rowData(se)[[3]]
protac$Gene <- rownames(protac)
protacFDR <- subset(protac, protac$adj.P.Val <0.1)


treatments <- unique(se$treatment)

# Define own colors
colors <- c("#DEDFE4", "darkgrey") # "#919297", "#393A3F",)# , "#800080", "black", "yellow", "orange")#232429"  

# Make sure the number of colors matches the number of treatments
if (length(colors) != length(treatments)) {
  stop("Number of colors must match number of treatments")
}

# Create a named vector for color mapping
color_mapping <- setNames(colors, treatments)

# Add color_mapping to the heatmap annotation
ha = HeatmapAnnotation(df = data.frame(treatment = factor(colData(se)$treatment, levels = names(color_mapping))),
                       col = list(treatment = color_mapping))

color_palette <- colorRampPalette(c("blue", "black", "orange"))(100)

pdf("plots_pdf/DEGs_EXP20-2_PROTAC_heatmap.pdf") #, width=14.5, height=10)
x <- sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
print(x)
dev.off()

png("plots_png/DEGs_EXP20-2_PROTAC_heatmap.png", height=768)
x <- sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
print(x)
dev.off()
#needs title : KH-103 effects on gene regulation in 8-cell embryos

svg("plots_svg/DEGs_EXP20-2_PROTAC_heatmap.svg")
x <- sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
print(x)
dev.off()

sechm(se, protacFDR, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
```


# Fig 3E - correlation of EXP20.2 target DEGs in 20.1 and Fig S3D - bland-altmann-plot correlation

```{r}
se <- readRDS("objects/MK_EXP20.1_ProtacEmbryo_merged.SE.rds")

# Generate the dataframes
dex <- rowData(se)[[4]]
protac <- rowData(se)[[3]]
pd <- rowData(se)[[5]]
dex$Gene <- rownames(dex)
protac$Gene <- rownames(protac)
pd$Gene <- rownames(pd)

# Generate dataframes with FDR<0.05 used in this analysis
dexFDR <- subset(dex, dex$FDR <0.05)
protacFDR <- subset(protac, protac$FDR <0.05)
pdFDR <- subset(pd, pd$FDR <0.05)
dexFDR$Gene <- rownames(dexFDR)
protacFDR$Gene <- rownames(protacFDR)
pdFDR$Gene <- rownames(pdFDR)



# Create a list of dataframes and add column Gene to it
df_list <- list(dexFDR = dexFDR, protacFDR = protacFDR, pdFDR = pdFDR)

rEXP20_2 <- readRDS("objects/df_list_FDR01.rds")
rEXP20_2 <- lapply(rEXP20_2, function(df) {
  df$Gene <- rownames(df)
  return(df)
})

# Function to automate the process
process_dataframes <- function(df_name, df_list, rEXP20_2) {
  # Get the dataframe from the list
  df <- df_list[[df_name]]
  
  # Construct the variable names
  FDR_var <- paste0(df_name, "FDR")
  
  # Filter the df dataframe to include only the genes that are in rEXP20_2[[FDR_var]]
  df_filtered <- df[df$Gene %in% rEXP20_2[[FDR_var]]$Gene, ]
  
  # Filter rEXP20_2[[FDR_var]] to include only the genes that are in df_filtered
  rEXP20_2_FDR_filtered <- rEXP20_2[[FDR_var]][rEXP20_2[[FDR_var]]$Gene %in% df_filtered$Gene, ]
  
  # Merge the two dataframes on the 'Gene' column
  merged <- merge(df_filtered, rEXP20_2_FDR_filtered, by = "Gene")
  
  # Calculate the correlation of the logFC between the two dataframes
  correlation <- cor.test(merged$logFC.x, merged$logFC.y)
  
  # Print the correlation
  print(correlation)
  
  correlation <- cor(merged$logFC.x, merged$logFC.y)
  
  # Create a scatter plot
  plot <- ggplot(merged, aes(x=logFC.x, y=logFC.y)) +
    geom_point() +
    geom_text_repel(aes(label=Gene)) +  # Add the gene names to the plot
    geom_smooth(method=lm, se=FALSE, color="orange") +
    annotate("text", x = Inf, y = -Inf, label = paste("Correlation:", round(correlation, 4)), vjust = -1.5, hjust = 1.2) +
    xlab(paste("logFC in 2-cell embryos")) +
    ylab(paste("logFC in 8-cell embryos")) +
    theme_classic() +
    ggtitle(paste("Correlation of gene expression changes between 2- and 8- cell embryos"))
  
  print(plot)
  ggsave("plots_svg/correlation_EXP20-2_EXP20-1_PROTAC.svg")
  ggsave("plots_pdf/correlation_EXP20-2_EXP20-1_PROTAC.pdf")
  ggsave("plots_png/correlation_EXP20-2_EXP20-1_PROTAC.png")
  
  
  # Calculate means and differences
  merged$mean <- rowMeans(merged[,c("logFC.x", "logFC.y")])
  merged$difference <- merged$logFC.x - merged$logFC.y
  
  # Create a Bland-Altman plot
  bland_altman_plot <- ggplot(merged, aes(x=mean, y=difference)) +
    geom_point() +
    geom_text_repel(aes(label=Gene)) +
    geom_smooth(method=lm, se=FALSE, color="orange") +
    xlab(paste("Mean of logFC in 2-cell and 8-cell embryos")) +
    ylab(paste("Difference of logFC in 2-cell and 8-cell embryos")) +
    theme_classic() +
    ggtitle("Correlation of gene expression changes between 2- and 8-cell embryos")
  
  print(bland_altman_plot)
  ggsave("plots_svg/correlationBAP_EXP20-2_EXP20-1_PROTAC.svg")
  ggsave("plots_pdf/correlationBAP_EXP20-2_EXP20-1_PROTAC.pdf")
  ggsave("plots_png/correlationBAP_EXP20-2_EXP20-1_PROTAC.png")
  
  return(merged)
}

# Can call this function with the dataframe names and the list of dataframes
df_names <- c("protac") #, "protac", "pd")
df_list <- list(protac = protac) #, protac = protac, pd = pd)

# Apply the function to each dataframe
lapply(df_names, process_dataframes, df_list = df_list, rEXP20_2 = rEXP20_2)
```

# Fig 3F - regionUpset DEGs in 2cell and 8cell


```{r}
library(UpSetR)

regionUpset2 <- function (x = NULL, genelist = NULL, returnList = FALSE, ...) 
{
    if (!is.null(genelist)) {
        x <- genelist
    }
    if (is.character(x)) 
        x <- as.list(x)
    stopifnot(length(x) > 1)
    if (is.list(x)) {
        if (is.null(names(x)) && all(unlist(lapply(x, is.character))) && 
            all(lengths(x) == 1)) 
            names(x) <- unlist(x)
        if (any(unlist(lapply(x, FUN = function(x) {
            if (is.character(x)) {
                if (length(x) != 0) return(TRUE)
                if (!file.exists(x)) return(TRUE)
            }
            FALSE
        })))) {
            stop("`x` should be a named list of character vectors.")
        }
    }
    x <- lapply(x, function(x) as.character(x))
    if (returnList) 
        return(x)
    UpSetR::upset(UpSetR::fromList(x), ...)
}
```

```{r}
df <- read.csv("objects/DEAFDR01-EXP20-1-2-PROTAC.csv", sep = ";")
# Create a list of vectors, each vector represents a category
sperm_overlaps <- df$Gene[df$Overlap.sperm.ATAC == 1]
gre <- df$Gene[df$GRE.in.gene == 1]
motifs <- df$Gene[df$gene.body.motifs.connected.to.GR == 1]

gene_list <- list(DEGs = df$Gene, "DEGs accesible in Sperm" = sperm_overlaps, "DEGs with GRE motif" = gre, "DEGs with GR binding partner motif" = motifs) 
```

```{r}
regionUpset2(gene_list)
```


```{r}
png(paste0("plots_png/Fig3F.png"), width=8, height=7,res = 300,units = "in")
regionUpset2(gene_list)
dev.off()

pdf(paste0("plots_pdf/Fig3F.pdf"), width=8, height=7,colormodel = "cymk")
regionUpset2(gene_list)
dev.off()


svg(paste0("plots_svg/Fig3F.svg"), width=8, height=7)
regionUpset2(gene_list)
dev.off()

```

# Fig S3A - GO term analysis embryo peaks with GRE

```{r}
embryo_consensus_peaks <- readRDS("./objects/peaks_with_GRE_embryo_consensus.rds")
seqlevelsStyle(embryo_consensus_peaks) <- "UCSC"
```

Liftover and pruning as GOterm analysis with submitGreatJob can only use mm10 and standard chromosomes

```{bash, eval = FALSE}
rsync -avzP rsync://hgdownload.soe.ucsc.edu/goldenPath/mm39/liftOver/mm39ToMm10.over.chain.gz -d ../../../tools/liftOver/
gzip -d ../../../tools/liftOver/mm39ToMm10.over.chain.gz
```

```{r}
mm39_mm10_chain <- import.chain("objects/mm39ToMm10.over.chain")

embryo_hc_peaks_mm10 <- liftOver(embryo_consensus_peaks, chain = mm39_mm10_chain)
embryo_hc_peaks_mm10 <- unlist(embryo_hc_peaks_mm10)
embryo_hc_peaks_mm10 <- keepStandardChromosomes(embryo_hc_peaks_mm10, pruning.mode = "coarse")
```

```{r}
length(embryo_consensus_peaks) #110
length(embryo_hc_peaks_mm10) #108
```

```{r, fig.width=10}
job <- submitGreatJob(embryo_hc_peaks_mm10, species="mm10") 
res <- getEnrichmentTables(job)

bp <- res$`GO Biological Process`
head(bp)
write.table(bp, file = "./objects//GO-analysis_biolprocess_hcembryoGRE.csv")

ggplot(head(bp,20), aes(Hyper_Fold_Enrichment, reorder(name, Hyper_Adjp_BH),  color=-log10(Hyper_Adjp_BH))) + geom_point() + scale_color_viridis_c() + ggtitle("Biological processes in embryo peaks with GRE")

ggsave("./plots_png//GO-term_enrichment_BiologicalProcess_embryo_GREhighconfidencepeaks.png")
```

# Fig S3B - sperm GRE peaks across embryo tracks

```{r, eval=FALSE}
tracks_embryos <- c(
  embryoearly2cell="extern_data/ATAC_PRJNA277362_Early2cell_merged.bw",
  embryo2cell="extern_data/tracks/ATAC_PRJNA277362_2cell_merged.bw",
  embryo4cell="extern_data/ATAC_PRJNA277362_4cell_merged.bw",
  ESC="extern_data/ATAC_PRJNA277362_ESC_GSM1625858.bw")

peaks <- readRDS("./rds_saves/peaks/GRE_peaks/peaks_with_GRE_sperm.rds")
m <- signal2Matrix(tracks_embryos, peaks, w=20, BPPARAM = 4)
m <- epiwraps:::.ese2ml(m)
n <- renormalizeBorders(lapply(m, log1p))


pdf("plots_pdf/GREpeaks-sperm-embryotracks.pdf")
x <- plotEnrichedHeatmaps(n, row_title="Peaks from sperm with GRE", trim = 0.99, raster_resize_mat = TRUE)
print(x)
dev.off()


png("plots_png/GREpeaks-sperm-embryotracks.png", width=1024, height=768)
x <- plotEnrichedHeatmaps(n, row_title="Peaks from sperm with GRE", trim = 0.99, raster_resize_mat = TRUE)
print(x)
dev.off()

print(x)
```



# Fig S3C - GSEA of DEGs 

```{r}
deaGO <- function(dea, threshold=0.05, direction=NULL, ontology="BP"){
  isDE <- dea$adj.P.Val<threshold
  if(!is.null(direction)) isDE <- isDE & sign(dea$logFC)==sign(direction)
  isDE <- setNames(factor(as.integer(isDE)),row.names(dea))
  print(table(isDE))
  god <- new("topGOdata", ontology=ontology, allGenes=isDE, 
             nodeSize=10, annot=annFUN.org, 
             ID="symbol", mapping="org.Mm.eg.db")
  ea <- runTest(god, algorithm="classic", statistic="fisher")
  res <- GenTable(god, classicFisher=ea, topNodes=length(god@graph@nodes))
  res$Term2 <- plgINS::breakStrings(res$Term)
  res$log2Enrichment <- log2(res$Significant/res$Expected)
  res$FDR <- p.adjust(res$classicFisher)
  res
}

deaGSEA <- function(dea, gsets=NULL){
  if(is.null(gsets))
    gsets <- plgINS::getMsigSets("Mus musculus", collections=c("H","C5"))
  signature <- setNames(sign(dea$logFC)*-log10(dea$P.Value), row.names(dea))
  fg <- fgseaMultilevel(gsets, signature, minSize=10, nproc=3)
  fg$genes <- sapply(fg$leadingEdge, collapse=", ", FUN=paste)
  fg$leadingEdge <- NULL
  as.data.frame(fg[order(fg$pval),])
}
```

for 2-cell
```{r}
se <- readRDS("objects/MK_EXP20.1_ProtacEmbryo_merged.SE.rds")

deas <- getDEA(se)
dea <- deas$PROTAC
```
```{r, eval=FALSE}
res2 <- deaGSEA(dea)
head(res2)
#write.csv(res2, "./results/GSEA_PROTAC.csv")
colnames(res2)
res2
#saveRDS(res2, "GSEA_2cell.rds")
```
```{r}
res2 <- readRDS("rds_saves/paper/GSEA_2cell.rds")
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$padj < 0.05
  ggplot(df, aes(x=NES, y=-log10(padj))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black")) +
    geom_text_repel(data = subset(df, highlight == TRUE), aes(label=pathway), size=3.5, min.segment.length = 0) +
    xlab("logFC") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    ggtitle(title)
}
generate_volcano_plot(res2, "Enriched gene sets in 2-cell embryos upon KH-103 treatment")
ggsave("plots_svg/GSEA_EXP20-1_PROTAC.svg")
ggsave("plots_pdf/GSEA_EXP20-1_PROTAC.pdf")
ggsave("plots_png/GSEA_EXP20-1_PROTAC.png")
```

alternative plot

```{r}
ggplot(head(res2,10), aes(x = reorder(pathway, NES), y = NES, fill = -log10(padj))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "grey", high = "orange") +
  theme_minimal() +
  labs(
    title = "Gene Ontology Enrichment overall",
    x = "GO Term",
    y = "Enrichment Score",
    fill = "-log10(FDR)"
  )
```

for 8-cell
```{r}
se <- readRDS("objects/MK_EXP20.2_ProtacEmbryo8cells.SE.rds")

deas <- getDEA(se)
dea <- deas$PROTAC
```
```{r, eval = FALSE}
res4 <- deaGSEA(dea)
res4

#saveRDS(res4, "GSEA_8cell.rds")
```
```{r}
res4 <- readRDS("objects/GSEA_8cell.rds")
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$padj < 0.05
  ggplot(df, aes(x=NES, y=-log10(padj))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black")) +
    geom_text_repel(data = subset(df, highlight == TRUE), aes(label=pathway), size=3.5, min.segment.length = 0) +
    xlab("NES") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    ggtitle(title)
}
generate_volcano_plot(res4, "Enriched gene sets in 8-cell embryos upon KH-103 treatment")
ggsave("plots_svg/GSEA_EXP20-2_PROTAC.svg")
ggsave("plots_pdf/GSEA_EXP20-2_PROTAC.pdf")
ggsave("plots_png/GSEA_EXP20-2_PROTAC.png")
```



# Fig 4E - 2cell Smartseq after Dex treatment in sperm and Fig S4F - 2cell Smartseq after KH-103 treatment in sperm

```{r}
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$adj.P.Val < 0.1
  top_genes <- df[order(df$adj.P.Val), ][1:10, ]
  ggplot(df, aes(x=logFC, y=-log10(adj.P.Val))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black")) +
    geom_text_repel(data = subset(top_genes, highlight == TRUE), aes(label=Gene), size=3.5, min.segment.length = 0, max.overlaps = Inf, nudge_y = 0.1) +
    geom_hline(yintercept=-log10(0.05), linetype="dashed")+
    xlab("logFC") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    ggtitle(title)
}

se <- readRDS("objects/MK_EXP16.4_embryoSeq_spermProtac_merged.SE.rds")

# Generate the dataframes 
dex <- rowData(se)[[4]]
protac <- rowData(se)[[3]]
pd <- rowData(se)[[5]]

protac$Gene <- rownames(protac)
dex$Gene <- rownames(dex)
pd$Gene <- rownames(pd)

# List of dataframes
df_list <- list(Dex = dex, "KH-103" = protac, "KH-103+Dex" = pd)

# Apply the function to each dataframe in the list
plots <- lapply(names(df_list), function(name) {
  generate_volcano_plot(df_list[[name]], paste("Differentially expressed genes in 2-cell embryos upon ", name, " treatment in sperm"))
})
print(plots[1])
ggsave("plots_svg/DEGs_EXP16-4_Dex_volcano.svg")
ggsave("plots_pdf/DEGs_EXP16-4_Dex_volcano.pdf")
ggsave("plots_png/DEGs_EXP16-4_Dex_volcano.png")

print(plots[2])
ggsave("plots_svg/DEGs_EXP16-4_PROTAC_volcano.svg")
ggsave("plots_pdf/DEGs_EXP16-4_PROTAC_volcano.pdf")
ggsave("plots_png/DEGs_EXP16-4_PROTAC_volcano.png")

print(plots[3])
ggsave("plots_svg/DEGs_EXP16-4_PD_volcano.svg")
ggsave("plots_pdf/DEGs_EXP16-4_PD_volcano.pdf")
ggsave("plots_png/DEGs_EXP16-4_PD_volcano.png")
```


# Fig 4F - Reversal of Dex by KH-103

Can we see reversal for genes not passing significance?

```{r}
se <- readRDS("objects/MK_EXP16.4_embryoSeq_spermProtac_merged.SE.rds")
se <- se[, se$treatment != "MED"]
rescue <- rowData(se)[[6]]
rescue$Gene <- rownames(rescue)

rescue_sorted <- rescue[order(rescue$adj.P.Val), ]
rescue_sorted <- rescue_sorted[1:20, ]

treatments <- unique(se$treatment)

# Define your own colors
colors <- c(brewer.pal(9,"Set2")[3], "grey", "orange", brewer.pal(9,"Set2")[5] )#"#0000FF", "#800080", "grey", "black")  

# Make sure the number of colors matches the number of treatments
if (length(colors) != length(treatments)) {
  stop("Number of colors must match number of treatments")
}

colors <- c("#919297", "#DEDFE4", "darkgrey", "#393A3F")
# Create a named vector for color mapping
color_mapping <- setNames(colors, treatments)

# Add color_mapping to the heatmap annotation
ha = HeatmapAnnotation(df = data.frame(treatment = factor(colData(se)$treatment, levels = names(color_mapping))),
                       col = list(treatment = color_mapping))

#sechm(se, rescue_sorted, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="KH103-mediated\nrescue of DEX effect", gaps_at="treatment")
#needs title: Effects of KH-103 treatment in sperm on gene regulation in 2-cell embryos


color_palette <- colorRampPalette(c("blue", "black", "orange"))(10000)

pdf("plots_pdf/DEGs_EXP16-4_rescue_heatmap.pdf", width=6, height=4) #, width=14.5, height=10)
x <- sechm(se, rescue_sorted, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="KH103-mediated\nrescue of DEX effect", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
print(x)
dev.off()

png("plots_png/DEGs_EXP16-4_rescue_heatmap.png")#, height=768)
x <- sechm(se, rescue_sorted, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="KH103-mediated\nrescue of DEX effect", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
print(x)
dev.off()

svg("plots_svg/DEGs_EXP16-4_rescue_heatmap.svg")#, height=768)
x <- sechm(se, rescue_sorted, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="KH103-mediated\nrescue of DEX effect", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
print(x)
dev.off()
rescue_sorted

sechm(se, rescue_sorted, assayName = "log2FC", do.scale = FALSE, top_annotation = ha, row_title="Differentially expressed genes", gaps_at="treatment", hmcols = color_palette, use_raster = TRUE)
```


#Fig 4G - correlation with PGC-GR-KO

Calculate correaltion and visualize with scatterplot

```{r}
KO <- readRDS("objects/VF_2cell_embyro_early_paternal_GR_KO(Alpl).SE.rds")
KO <- rowData(KO)[[2]]
koFDR <- subset(KO, KO$FDR <0.05)
koFDR$Gene <- koFDR$ID


se <- readRDS("objects/SEs_for_SEV/MK_EXP16.4_embryoSeq_spermProtac_merged.SE.rds")
protac <- rowData(se)[[3]]
protac$Gene <- rownames(protac)


merged_data <- merge(koFDR, protac, by.x = "ID", by.y = "Gene")

correlation <- cor.test(merged_data$logFC.x, merged_data$logFC.y)
correlation

# Create a scatter plot to visualize the correlation
plot <- ggplot(merged_data, aes(x=logFC.x, y=logFC.y)) +
    geom_point() +
    geom_text_repel(aes(label=Gene)) +  # Add the gene names to the plot
    geom_smooth(method=lm, se=FALSE, color="orange") +
    #annotate("text", x = Inf, y = -Inf, label = paste("Correlation:", correlation)) +#, vjust = -1.5, hjust = 1.2) +
    xlab(paste("logFC in PGC-GR KO embryos")) +
    ylab(paste("logFC in embryos from KH-103 treated sperm")) +
    ggtitle(paste("Correlation of gene expression changes \n between PGC-GR KO embryos and embryos from KH-103 treated sperm"))+
    theme_classic()

print(plot)
ggsave("plots_svg/correlation_VF-KO_EXP16-4_PROTAC.svg")
ggsave("plots_pdf/orrelation_VF-KO_EXP16-4_PROTAC.pdf")
ggsave("plots_png/correlation_VF-KO_EXP16-4_PROTAC.png")
```


# Fig S4G - tracks accesscible in sperm

```{r}
gr1 <- txs[txs$gene_name == "1810019D21Rik"]
gr2 <- txs[txs$gene_name == "Magea14"]
gr3 <- txs[txs$gene_name == "Gm9895"]
gr4 <- txs[txs$gene_name == "Zfp472"]
gr5 <- txs[txs$gene_name == "Cyp4f17"]
gr5 <- txs[txs$gene_name == "Adamts4"]
gr6 <- txs[txs$gene_name == "Tmem68"]
gr7 <- txs[txs$gene_name == "Ptpmt1"]
gr8 <- txs[txs$gene_name == "Zfp873"] #
gr9 <- txs[txs$gene_name == "Trib3"] #

ensdb <- ah[["AH89457"]]

txs <- transcripts(ensdb, columns=c("tx_id","tx_biotype","tx_name","gene_id","gene_name"))

tracks <- c(
  Spermatozoa="extern_data/ATAC_PRJNA315208_Sperm_GSM2088377.bw"
)


plotSignalTracks(tracks, region=gr4[1], ensdb=ensdb, transcripts="full", genomeAxis = TRUE)

plotSignalTracks(tracks, region=gr6[1], ensdb=ensdb, transcripts="full", genomeAxis = TRUE)

plotSignalTracks(tracks, region=gr8[1], ensdb=ensdb, transcripts="full", genomeAxis = TRUE)

gene_symbols <- mapIds(org.Mm.eg.db, keys=tracks$gene_id, column="SYMBOL", keytype="ENSEMBL", multiVals="first")
tracks$gene_symbol <- gene_symbols

```

```{r}
pdf("plots_pdf/rescue_tracks.pdf" , width=10, height=5)
x <- plotSignalTracks(tracks, region=gr8[1], ensdb=ensdb, transcripts="full", genomeAxis = TRUE)
print(x)
dev.off()

png("plots_png/rescue_tracks.png", height=768)
x <- plotSignalTracks(tracks, region=gr8[1], ensdb=ensdb, transcripts="full", genomeAxis = TRUE)
print(x)
dev.off()
```

# Fig S4H - RegionUpset
```{r}
df <- read.csv("objects/DEA_rescueDEGs_Fig4_venn_categories.csv", sep = ";")
# Create a list of vectors, each vector represents a category
sperm_overlaps <- df$Gene[df$Overlap.sperm.ATAC == 1]
gre <- df$Gene[df$GRE.in.gene == 1]
motifs <- df$Gene[df$gene.body.motifs.connected.to.GR == 1]

gene_list <- list(DEGs = df$Gene, "DEGs accesible in Sperm" = sperm_overlaps, "DEGs with GRE motif" = gre, "DEGs with GR binding partner motif" = motifs) 
```





```{r}
library(UpSetR)

regionUpset2 <- function (x = NULL, genelist = NULL, returnList = FALSE, ...) 
{
    if (!is.null(genelist)) {
        x <- genelist
    }
    if (is.character(x)) 
        x <- as.list(x)
    stopifnot(length(x) > 1)
    if (is.list(x)) {
        if (is.null(names(x)) && all(unlist(lapply(x, is.character))) && 
            all(lengths(x) == 1)) 
            names(x) <- unlist(x)
        if (any(unlist(lapply(x, FUN = function(x) {
            if (is.character(x)) {
                if (length(x) != 0) return(TRUE)
                if (!file.exists(x)) return(TRUE)
            }
            FALSE
        })))) {
            stop("`x` should be a named list of character vectors.")
        }
    }
    x <- lapply(x, function(x) as.character(x))
    if (returnList) 
        return(x)
    UpSetR::upset(UpSetR::fromList(x), ...)
}
```


```{r}
png(paste0("plots_png//Fig4rescue.png"),width=8, height=7,res = 300,units = "in")
regionUpset2(gene_list)
dev.off()

pdf(paste0("plots_pdf/Fig4rescue.pdf"), width=8, height=7,colormodel = "cymk")
regionUpset2(gene_list)
dev.off()


svg(paste0("plots_svg/Fig4rescue.svg"), width=8, height=7)
regionUpset2(gene_list)
dev.off()

```

# Fig S4H - 2cell volcano


```{r}
se <- readRDS("objects/MK_EXP20.1_ProtacEmbryo_merged.SE.rds")

# Generate the dataframes
dex <- rowData(se)[[4]]
protac <- rowData(se)[[3]]
pd <- rowData(se)[[5]]

protac$Gene <- rownames(protac)
dex$Gene <- rownames(dex)
pd$Gene <- rownames(pd)

# List of dataframes
df_list <- list(Dex = dex, "KH-103" = protac, "KH-103+Dex" = pd)

# Function to generate volcano plots
generate_volcano_plot <- function(df, title) {
  df$highlight <- df$FDR < 0.1
  top_genes <- df[order(df$FDR), ][1:10, ]
  ggplot(df, aes(x=logFC, y=-log10(FDR))) +
    geom_point(aes(color = highlight)) +
    scale_color_manual(values = c("TRUE" = "orange", "FALSE" = "black")) +
    geom_text_repel(data = subset(top_genes, highlight == TRUE), aes(label=Gene), size=3.5, min.segment.length = 0, max.overlaps = Inf, nudge_y = 0.1) +
    geom_hline(yintercept=-log10(0.05), linetype="dashed")+
    xlab("logFC") +
    ylab("-log10(adjusted p-value)") +
    theme_classic() +
    theme(legend.position="none") +
    ggtitle(title)
}

# Apply the function to each dataframe in the list
plots <- lapply(names(df_list), function(name) {
  generate_volcano_plot(df_list[[name]], paste("Differentially expressed genes in 2-cell embryos upon ", name, " treatment"))
})

print(plots)

print(plots[1])
ggsave("plots_svg/DEGs_EXP20-1_Dex_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-1_Dex_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-1_Dex_volcano.png")

print(plots[2])
ggsave("plots_svg/DEGs_EXP20-1_PROTAC_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-1_PROTAC_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-1_PROTAC_volcano.png")

print(plots[3])
ggsave("plots_svg/DEGs_EXP20-1_PD_volcano.svg")
ggsave("plots_pdf/DEGs_EXP20-1_PD_volcano.pdf")
ggsave("plots_png/DEGs_EXP20-1_PD_volcano.png")
```

# Fig 5B Baseline Cort difference
```{r}
theme_set(theme_classic())
```

```{r}
cort_df <- data.frame(Condition = c(rep("Control",8),rep("Stress",8)),
           Cort = c(34.83,	41.514,	24.323,	31.654,	27.842,	60.724,	52.208,	53.683,	70.372,	40.766,	36.163,	89.217,	61.647,	42.545,	47.178,	76.72))
```

```{r,fig.width=5,fig.height=5}

  e <- ggplot(cort_df, aes(x = Condition, y = Cort, fill = Condition))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+
   ylab("Cort. [ng/ml]")+
   xlab("Condition")+
   ggtitle("Baseline Cort. after Stress")+
   scale_fill_manual(values=c( "seashell2","#ffa500ff"))+
   labs(fill= "Condition")+
   ylim(c(0,100))+

 theme(
  axis.text.x = element_text(size = 13),
  text = element_text(size = 14)
)
 
 
 print(e2)
 
 ggsave("plots_png/Fig5B.StressedAnimalsBaselineCort.png")
  ggsave("plots_pdf/Fig5B.StressedAnimalsBaselineCort.pdf")
   ggsave("plots_svg/Fig5B.StressedAnimalsBaselineCort.svg")

```







# Fig 5C OFT TIP offspring sep
```{r}
report <- read.csv("objects/OFT_Report_offspring.csv")
rownames(report) <- report$file
table(is.na(report))

# we convert NAs to zero
report[is.na(report)] <- 0

meta <- openxlsx::read.xlsx("objects/OFT_metadata_offspring.xlsx")
rownames(meta) <- meta$OFTfile

# we order meta file according to report and check if TRUE
meta <- meta[rownames(report),]
table(rownames(report) == rownames(meta))

# we separate countdata from continues data --> important for downstream analysis

report_count <- report[,grep(x = colnames(report),pattern = "count|trans")]
report_count <- report_count[,-1]
report_rest <-  report[,-grep(x = colnames(report),pattern = "count|trans|percentage|raw")]
report_rest <- report_rest[,c(-1,-2,-which(colnames(report_rest) == "bodycentre.total.time"))]

library(tidyr)

meta <- meta %>%
  separate(condition, into = c("stress", "treatment"), sep = " ", remove = FALSE)

meta$stress <- as.factor(meta$stress)
meta$sex <- as.factor(meta$sex)
meta$treatment <- as.factor(meta$treatment)
meta$condition <- gsub(x = meta$condition,pattern = " ",replacement = ".")
```

```{r}

meta$condition <- gsub(x = meta$condition,pattern = " ",replacement = ".")
# se_count$Group <- relevel(se_count$Group,c("non.stressed.DMSO.f"))
meta$condition <- factor(meta$condition, levels = c("non.stressed.DMSO", "stressed.DMSO", "non.stressed.PROTAC","stressed.PROTAC"))

df <- cbind(report_rest,report_count,meta)

df$condition
```

```{r}
male <- "\u2642"
print(paste("Male:", male))


female <- "\u2640"
print(paste("Female:", female))


e <- ggplot(df, aes(x = df$sex, y = df$bodycentre.periphery.total.time, fill = df$condition))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("Time [s]")+
   xlab("Sex")+
   ggtitle("Time Periphery")+

   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+
   labs(fill= "Condition")+
   scale_x_discrete(labels=c("f"=paste(female), "m"=paste( male)))+

 theme(
  axis.text.x = element_text(size = 20),
  text = element_text(size = 14),
)
 
 
 print(e2)
 
  ggsave("plots_svg/Fig5.Offspring.TIP.svg")
ggsave("plots_pdf/Fig5.Offspring.TIP.pdf")#, width = 6, height = 4)
ggsave("plots_png/Fig5.Offspring.TIP.png")

```

# Fig 5D OFT offspring distance

```{r}

male <- "\u2642"
print(paste("Male:", male))


female <- "\u2640"
print(paste("Female:", female))


y <- 37.5
e <- ggplot(df, aes(x = df$sex, y = df$bodycentre.distance.moving, fill = df$condition))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("Distance [cm]")+
   xlab("Sex")+
   ggtitle("Distance Covered")+

   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+ 
   labs(fill= "Condition")+
   scale_x_discrete(labels=c("f"=paste(female), "m"=paste( male)))+

 theme(
  axis.text.x = element_text(size = 25),
  text = element_text(size = 14), 
)
 
 
 print(e2)
 
ggsave("plots_svg/Fig5.Offspring.Distance.svg")
ggsave("plots_pdf/Fig5.Offspring.Distance.pdf")#, width = 6, height = 4)
ggsave("plots_png/Fig5.Offspring.Distance.png")




```




# Fig 5E UMAP of females offspring of stressed fathers
```{r}
set.seed(123)
i <- 64

#read in US oject
US <- readRDS("objects/US_object_offspring.rds")

#subset to have only females from IVF cohort
US <- SplitUSData(us = US,omit = US$file_names[ US$meta$sex == "m" ])

US_f <- US

#read in US oject

US_f <- TwoGroupAnalysis(us = US_f,group = US_f$meta$Stress)
US_f <- TwoGroupAnalysis_Clusters(us = US_f,group = US_f$meta$Stress)

  # Create the plot
  x <- Plot2DEmbedding(
    US_f,
    reports = US$label_names,
    colorby = "Group",
    color = c("seashell3", "#ffa500ff", "seashell4", "tomato3"),
    plot.sem = TRUE, 
    seed = 64
  ) + ggtitle(paste0("Syllable Count CTRL-vs-IFS offspring females"))

  print(x)  
    
ggsave("plots_svg/Fig5E.UMAPfemales.svg")
ggsave("plots_pdf/Fig5E.UMAPfemales.pdf")#, width = 6, height = 4)
ggsave("plots_png/Fig5E.UMAPfemales.png")


```

# Fig 5F Permutation Test female offspring Stress vs Control in DMSO

```{r}
set.seed(64)

#Histogram for CTRL vs IFS in DMSO group for female offspring

#read in US oject
US <- readRDS("objects/US_object_offspring.rds")

#subset to have only females from IVF cohort
US <- SplitUSData(us = US,omit = US$file_names[ US$meta$sex == "m" ])

US_f <- US

#read in US oject

US_f <- SplitUSData(us = US_f,omit = US$file_names[US$meta$Treatment == "PROTAC"])
US_f <- TwoGroupAnalysis(us = US_f,group = US_f$meta$Stress)
US_f <- TwoGroupAnalysis_Clusters(us = US_f,group = US_f$meta$Stress,n_bootstraps = 10000)

US_f$meta

  x <- PlotClusterStats(US_f, labels = US_f$label_names)
  print(x)

ggsave("plots_svg/Fig5F.PermFemalesIFSvsCtrlDMSO.svg")
ggsave("plots_pdf/Fig5F.PermFemalesIFSvsCtrlDMSO.pdf")#, width = 6, height = 4)
ggsave("plots_png/Fig5F.PermFemalesIFSvsCtrlDMSO.png")



```

```{r}
#read in US oject
US <- readRDS("US_SF_II_OFTs.rds")

#subset to have only females from IVF cohort
US <- SplitUSData(us = US,omit = US$file_names[US$meta$generation == "parent" | US$meta$sex == "m" | US$meta$breeding == "natural"])
#relabel the conditions
US$meta$condition[US$meta$condition == "non stressed PROTAC"] <- "CTRL PROTAC"
US$meta$condition[US$meta$condition == "non stressed DMSO"] <- "CTRL DMSO"
US$meta$condition[US$meta$condition == "stressed PROTAC"] <- "IFS PROTAC"
US$meta$condition[US$meta$condition == "stressed DMSO"] <- "IFS DMSO"

US$meta <- US$meta[,-14]

US$meta <- US$meta %>%
  tidyr::separate(condition, into = c("Stress", "Treatment"), sep = " ")
US$meta$Group <- interaction(US$meta$Stress, US$meta$Treatment)
US_f <- US
US_f <- SplitUSData(us = US_f,omit = US$file_names[US$meta$Treatment == "PROTAC"])

US_f <- TwoGroupAnalysis(us = US_f,group = US_f$meta$Stress)
US_f <- TwoGroupAnalysis_Clusters(us = US_f,group = US_f$meta$Stress)


  x <- PlotClusterStats(US_f, labels = US_f$label_names[4])
  print(x)


```


# Fig 5G Permutation Test female offspring Stress vs Control in KH-103


```{r}
set.seed(64)

#Histogram for CTRL vs IFS in DMSO group for female offspring

#read in US oject
US <- readRDS("objects/US_object_offspring.rds")

#subset to have only females from IVF cohort
US <- SplitUSData(us = US,omit = US$file_names[ US$meta$sex == "m" ])

US_f <- US

#read in US oject

US_f <- SplitUSData(us = US_f,omit = US$file_names[US$meta$Treatment == "DMSO"])
US_f <- TwoGroupAnalysis(us = US_f,group = US_f$meta$Stress)
US_f <- TwoGroupAnalysis_Clusters(us = US_f,group = US_f$meta$Stress)



  x <- PlotClusterStats(US_f, labels = US_f$label_names)
  print(x)

ggsave("plots_svg/Fig5F.PermFemalesIFSvsCtrlPROTAC.svg")
ggsave("plots_pdf/Fig5F.PermFemalesIFSvsCtrlPROTAC.pdf")#, width = 6, height = 4)
ggsave("plots_png/Fig5F.PermFemalesIFSvsCtrlPROTAC.png")



```

# Fig 5H SPT split by sex
```{r}
spt_df <- data.frame(Sex = c(rep("f",13),rep("m",17)),Treatment = c(rep("DMSO",4),rep("PROTAC",2),rep("DMSO",4),rep("PROTAC",3), rep("DMSO",4),rep("PROTAC",4),rep("DMSO",5),rep("PROTAC",4)), Condtion = c(rep("CTRL",6),rep("Stress",7),rep("CTRL",8),rep("Stress",9)),
           SPT = c(83.21917808,	83.60299445,	83.60299445,	78.0042592, 83.56718834,	84.69643753, 74.65865647,	74.45669291,	76.78244973,	78.59216255,85.25290905,	89.54724409,	78.77526754,87.76315789,	88.15100579,	83.04337172,	82.6875, 83.72475341,	85.91406908,	86.40839041,	79.18757711, 85.33941814,	70.62055591,	78.34249953,	80.44692737,	79.90169306,		75.66512082,	82.14158026,	80.68035944,	76.53459821))
```

```{r}
spt_df$Group <- interaction(spt_df$Treatment,spt_df$Condtion)
```

```{r}
spt_df$Group <- factor(spt_df$Group, levels = c("DMSO.CTRL", "DMSO.Stress", "PROTAC.CTRL","PROTAC.Stress"))
```


```{r,fig.width= 5,fig.height=4}

# Male symbol
male <- "\u2642"
print(paste("Male:", male))

# Female symbol
female <- "\u2640"
print(paste("Female:", female))


y <- 37.5
e <- ggplot(spt_df, aes(x = Sex, y = SPT, fill = Group))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("Overall Saccarine consume [%]")+
   xlab("Sex")+
   ggtitle("Saccharine Preference Test")+
   ylim(c(70,100))+

   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+

   labs(fill= "")+
   scale_x_discrete(labels=c("f"=paste(female), "m"=paste( male)))+

 theme(
  axis.text.x = element_text(size = 25),
  text = element_text(size = 14),  
)
 
 
 print(e2)
 
 ggsave("plots_png/Fig5H.SPTbySex.png")
  ggsave("plots_pdf/Fig5H.SPTbySex.pdf")
   ggsave("plots_svg/Fig5H.SPTbySex.svg")



```

# Fig 5I Heatmap of differential expressed genes with a counteraction of KH-103 and Stress


```{r}
se <- readRDS("objects/IFS_smartSeq_embryo_salmon.SE.rds")
meta <- openxlsx::read.xlsx("objects/IFS_smartSeq_embryo_meta.xlsx",colNames = T) 
meta <- meta[order(meta$Sample),]
meta <- meta[meta$Sample %in% colnames(se),]
table(meta$Sample == colnames(se))
se$embryos <- as.factor(meta$embryos)

Group <- factor(paste(se$treatment,se$condition,sep="."))


se$Group <- Group

set.seed(123)
se$Group2 <- sample(Group,size = 17,replace = F)
colData(se)
meta$Sample
```

```{r}
set.seed(123)
se <- SEtools::svacor(se, ~Group+embryos,form0 = ~embryos, n.sv = 3)

```

```{r}
dds <- calcNormFactors(DGEList(assay(se))) # this calculates Normalization factors according to library sequencing depth, this is important since depth can vary greatly between samples

### lets look at the library size
dds$samples



assays(se)$logcpm <- log1p(edgeR::cpm(dds)) # generates the log of Counts-per-million (CPM), this is a very standardized way to normalize data used then for analysis or visualisation




mm <- model.matrix(~0+Group+SV1+SV2+embryos, data=as.data.frame(colData(se))); mm


dds <- dds[filterByExpr(dds, group = Group, min.count=20),] # 20 maybe to stringent for embryo data
dds <- estimateDisp(dds,design = mm)
se <- se[row.names(dds),]

fit <- glmFit(dds, mm, prior.count = 5) # we us prior.count 5 since we are working with low count data
```
Creating contrasts
```{r}
fit$design

my.contrast <- makeContrasts(
  DMSO_IFSvsCTRL = GroupDMSO.IFS - GroupDMSO.CTRL ,
  PROTAC_IFSvsCTRL = GroupPROTAC.IFS - GroupPROTAC.CTRL,
  IFS_PROTACvsDMSO = GroupPROTAC.IFS - GroupDMSO.IFS,
  CTRL_PROTACvsDMSO = GroupPROTAC.CTRL - GroupDMSO.CTRL,
  IFS_PROTAC_int = ((GroupPROTAC.IFS - GroupPROTAC.CTRL) - (GroupDMSO.IFS - GroupDMSO.CTRL)), # interaction term

  Total_IFSvsCTRL = (GroupDMSO.IFS - GroupDMSO.CTRL + GroupPROTAC.IFS - GroupPROTAC.CTRL)/2,
  Total_PROTACvsDMSO =  (GroupPROTAC.IFS - GroupDMSO.IFS + GroupPROTAC.CTRL - GroupDMSO.CTRL)/2,
  levels = mm
)
my.contrast  

```


```{r}
Results <- list()
 
for (con in colnames(my.contrast)){
  Results[[con]] <- as.data.frame(topTags(glmLRT(fit,contrast = my.contrast[,con]),Inf))
  Results[[con]]$ID <- gsub(x = rownames(Results[[con]]),pattern = "ENS(.*)\\.",replacement = "")
  Results[[con]]$DEA <- con
}
Results

```

```{r}
RNA_df_Stress_total <- Results$Total_IFSvsCTRL


RNA_df_Stress_Protac <- Results$IFS_PROTAC_int
RNA_df_Stress_Protac <- RNA_df_Stress_Protac[which(RNA_df_Stress_Protac$FDR < 0.05),]

habituation_list <- RNA_df_Stress_total[, c(1, 6)] #get the logFC and ID
colnames(habituation_list)[1] <- "Stress_coef"

habituation_list$StressProtac_coef <- RNA_df_Stress_Protac[rownames(RNA_df_Stress_total[rownames(RNA_df_Stress_total),]),]$logFC
habituation_list1 <- habituation_list[which(habituation_list$StressProtac_coef * habituation_list$Stress_coef < 0),]
```

```{r}
metadata(se)$anno_colors <- list(condition=c(CTRL="steelblue", IFS ="firebrick3"), treatment =c(PROTAC = "orange", DMSO = "purple4"), Group = c("DMSO.CTRL" = "seashell2","DMSO.IFS" =  "#ffa500ff" , "PROTAC.CTRL" = "seashell4" , "PROTAC.IFS" = "tomato3"))

```
```{r}
color_palette <- colorRampPalette(c("blue", "black", "orange"))(101)
rownames(se) <- gsub(x = rownames(se),pattern = "^ENS(.*)\\.",replacement = "")



sh <- sechm(se,features = habituation_list1$ID,assayName = "corrected",hmcols = color_palette,do.scale =T,gaps_at = c("treatment"),column_order = colnames(se)[order(colData(se)$condition)],show_rownames = T,name = paste0("Counter.Genes","\nz-score"),top_annotation = c("Group"),use_raster = T,column_title = "Differential gene expression in 2-cell embryos\nupon parternal stress and PROTAC counteraction",column_title_gp = gpar(fontsize = 20)) 
print(sh)



pdf("plots_pdf/Fig5I.counteractingProtacGenes.heatmap.pdf", width=14.5, height=10) #, width=14.5, height=10)

print(sh)
dev.off()

png("plots_png/Fig5I.counteractingProtacGenes.heatmap.png", height=638,width = 872,)

print(sh)
dev.off()

svg("plots_svg/Fig5I.counteractingProtacGenes.heatmap.svg") #, width=14.5, height=10)

print(sh)
dev.off()

```

# Fig 5J Differential genes impacted by Stress and KH103 in 2-cell embryo as wel as by Dex in excitatory Neurons in human neural organoids

```{r}
se <- readRDS("objects/SummarizedExperimentObject_EmbryoFromIFS_Sperm_treatedKH103.SE.rds")
```

```{r}

df <- as.data.frame(cbind(colData(se),t(assays(se)$corrected)))
```

"Control" = "#c0c0ffff", "Stress" = "#ffa500ff"
```{r}



# Calculate range and breaks
y_min <- floor(min(df$ENSMUSG00000039114.Nrn1))
y_max <- ceiling(max(df$ENSMUSG00000039114.Nrn1))
y_breaks <- seq(y_min, y_max, length.out = 5)

# Round to nearest 0.5
round_to_half <- function(x) round(x * 2) / 2
y_breaks <- round_to_half(y_breaks)

# Adjust limits to match rounded breaks
y_lim <- range(y_breaks)



e <- ggplot(df, aes(x = Group, y = df$ENSMUSG00000039114.Nrn1, fill = df$Group))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("Corrected Expression Value")+
   xlab("Groups")+
   ggtitle("2Cell Embryo: Nrn1 expression")+

   

scale_y_continuous(breaks = y_breaks, limits = y_lim) +


   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+
   labs(fill= "Condition")+

 theme(
  axis.text.x = element_text(size = 9),
  text = element_text(size = 14), 
)
 
 
 print(e2)


 ggsave("plots_png/Fig5J_2CellNrn1.png")
  ggsave("plots_pdf/Fig5J_2CellNrn1.pdf")
   ggsave("plots_svg/Fig5J_2CellNrn1.svg")
```
```{r}
e <- ggplot(df, aes(x = Group, y = ENSMUSG00000021214.Akr1c18, fill = df$Group))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("Corrected Expression Value")+
   xlab("Groups")+
   ggtitle("2Cell Embryo: Akr1c18 expression")+

   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+
   labs(fill= "Condition")+
   ylim(c(5.5,10.5))+

 theme(
  axis.text.x = element_text(size = 9),
  text = element_text(size = 14), 
)
 
 
 print(e2)

 
  ggsave("plots_png/Fig5J_Akr1c18.png")
  ggsave("plots_pdf/Fig5J_Akr1c18.pdf")
   ggsave("plots_svg/Fig5J_Akr1c18.svg")
```



```{r}
se <- readRDS("objects//Binder_DexOrganoids_ExInh_Neurons.SE.rds")
se <- se[,grep(x = colData(se)$treatment,pattern = "Veh-Veh|Veh-Acu")]
se <- se[,grep(x = colData(se)$celltype,pattern = "Ex.Neurons")]
se <- se[,grep(x = colData(se)$cellline,pattern = "fok4")]


```
```{r}


df <- as.data.frame(cbind(colData(se),t(assays(se)$logcpm)))
```


```{r}
e <- ggplot(df, aes(x = treatment, y = NRN1, fill = treatment))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("logcpm")+
   xlab("Treatment")+
   ggtitle("Binder.Ex.Neurons: NRN1 expression")+

   scale_fill_manual(values=c( "seashell3","orange3"))+
   labs(fill= "Treatment")+

 theme(
  axis.text.x = element_text(size = 11),
  text = element_text(size = 14),
)
 
 
 print(e2)

  ggsave("plots_png/Fig5J_Nrn1Binder_changedcolor.png")
  ggsave("plots_pdf/Fig5J_Nrn1Binder_changedcolor.pdf")
   ggsave("plots_svg/Fig5J_Nrn1Binder_changedcolor.svg")

```
```{r}
e <- ggplot(df, aes(x = treatment, y = AKR1C1, fill = treatment))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("logcpm")+
   xlab("Treatment")+
   ggtitle("Binder.Ex.Neurons: AKR1C1 expression")+

   scale_fill_manual(values=c( "seashell3","orange3"))+
   labs(fill= "Treatment")+

 theme(
  axis.text.x = element_text(size = 11),
  text = element_text(size = 14),  
)
 
 
 print(e2)

  ggsave("plots_png/Fig5J_AKR1C1Binder_recolored.png")
  ggsave("plots_pdf/Fig5J_AKR1C1Binder_recolored.pdf")
   ggsave("plots_svg/Fig5J_AKR1C1Binder_recolored.svg")
 
```

# Fig S5A
```{r}
# load data
report_tracking <- read.csv("objects/OFT_Report.csv",sep = ";")
report_tracking <- data.frame(report_tracking)

meta <- openxlsx::read.xlsx("objects/stressed_males_OFT_metadata.xlsx") # check if everything is right

# check if order is right
report_tracking$file == meta$OFTfile

# if not change order of meta file
report_tracking$file == meta[order(meta$OFTfile),]$OFTfile

report_tracking$condition <- meta[order(meta$OFTfile),]$condition

#Relabel "IFS" as "Stress"

report_tracking$condition <- factor(report_tracking$condition,
                                    levels = c("Control", "IFS"),
                                    labels = c("Control", "Stress"))


```

## TIP
```{r}
e <- ggplot(report_tracking, aes(x = condition, y = bodycentre.periphery.total.time, fill=condition))
cols <- c("seashell2", "#ffa500ff")

e2 <- e +
  geom_boxplot(width = 0.75, outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.1), dodge.width = 0.75) +
  #stat_summary(fun = mean, geom = "point", shape = 4, size = 3, color = "red", fill = "red") +
  geom_signif(comparisons = list(c("Control", "Stress")), map_signif_level = TRUE) +
  scale_fill_manual(values = cols) +
  labs(title = "Time in Periphery") +
  ylab("Time [s]") +
  xlab("Condition") +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 20),
    text = element_text(size = 14)
  )

print(e2)


ggsave("plots_png/FigS5A.TIP.png")
ggsave("plots_pdf/FigS5A.TIP.pdf")
ggsave("plots_svg/FigS5A.TIP.svg")
```

## TtoC
```{r}
e <- ggplot(report_tracking, aes(x = condition, y = bodycentre.center.transitions, fill=condition))
cols <- c("seashell2", "#ffa500ff")

e2 <- e +
  geom_boxplot(width = 0.75, outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.1), dodge.width = 0.75) +
  #stat_summary(fun = mean, geom = "point", shape = 4, size = 3, color = "red", fill = "red") +
  geom_signif(comparisons = list(c("Control", "Stress")), map_signif_level = TRUE) +
  scale_fill_manual(values = cols) +
  labs(title = "Center Transitions") +
  ylab("Counts") +
  xlab("Condition") +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 20),
    text = element_text(size = 14)
  )

print(e2)

ggsave("plots_png/FigS5A.Transitions.png")
ggsave("plots_pdf/FigS5A.Transitions.pdf")
ggsave("plots_svg/FigS5A.Transitions.svg")
```

## Distance Covered
```{r}
e <- ggplot(report_tracking, aes(x = condition, y = bodycentre.distance.moving, fill=condition))
cols <- c("seashell2", "#ffa500ff")

e2 <- e +
  geom_boxplot(width = 0.75, outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.1), dodge.width = 0.75) +
  #stat_summary(fun = mean, geom = "point", shape = 4, size = 3, color = "red", fill = "red") +
  geom_signif(comparisons = list(c("Control", "Stress")), map_signif_level = TRUE) +
  scale_fill_manual(values = cols) +
  labs(title = "Distance covered") +
  ylab("[cm]") +
  xlab("Condition") +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 20),
    text = element_text(size = 14)
  )

print(e2)

ggsave("plots_png/FigS5A.Distance.png")
ggsave("plots_pdf/FigS5A.Distance.pdf")
ggsave("plots_svg/FigS5A.Distance.svg")
```

# Fig S5B Cluster analysis for stressed animals
```{r}
set.seed(123)
US <- readRDS("objects/US_object_IFS.rds")

US <- TwoGroupAnalysis(US, group = US$meta$Stress)
US <- TwoGroupAnalysis_Clusters(US, group = US$meta$Stress)

  x <- Plot2DEmbedding(US, report = US$label_names[2], colorby = "Stress", color = c("seashell3", "#ffa500ff"), plot.sem = T)+ ggtitle(paste0("UMAP of IFS animals"))

  print(x)
  ggsave("plots_pdf/FigS5B_IFS_animals_2D.pdf")
  ggsave("plots_png/FigS5B_IFS_animals_2D.png")
  ggsave("plots_svg/FigS5B_IFS_animals_2D.svg")


set.seed(123)

  x <- PlotClusterStats(US, labels = US$label_names[2])
  print(x)
  ggsave("plots_pdf/FigS5B_IFS_animals_UMAP.pdf")
  ggsave("plots_png/FigS5B_IFS_animals_UMAP.png")
  ggsave("plots_svg/FigS5B_IFS_animals_UMAP.svg")

```



# Fig S5C 

```{r}
report <- read.csv("objects/OFT_Report_offspring.csv")
rownames(report) <- report$file
table(is.na(report))

# we convert NAs to zero
report[is.na(report)] <- 0

meta <- openxlsx::read.xlsx("objects/OFT_metadata_offspring.xlsx")
rownames(meta) <- meta$OFTfile

# we order meta file according to report and check if TRUE
meta <- meta[rownames(report),]
table(rownames(report) == rownames(meta))

# we separate countdata from continues data --> important for downstream analysis

report_count <- report[,grep(x = colnames(report),pattern = "count|trans")]
report_count <- report_count[,-1]
report_rest <-  report[,-grep(x = colnames(report),pattern = "count|trans|percentage|raw")]
report_rest <- report_rest[,c(-1,-2,-which(colnames(report_rest) == "bodycentre.total.time"))]

library(tidyr)

meta <- meta %>%
  separate(condition, into = c("stress", "treatment"), sep = " ", remove = FALSE)

meta$stress <- as.factor(meta$stress)
meta$sex <- as.factor(meta$sex)
meta$treatment <- as.factor(meta$treatment)
```

```{r}

meta$condition <- gsub(x = meta$condition,pattern = " ",replacement = ".")
# se_count$Group <- relevel(se_count$Group,c("non.stressed.DMSO.f"))
meta$condition <- factor(meta$condition, levels = c("non.stressed.DMSO", "stressed.DMSO", "non.stressed.PROTAC","stressed.PROTAC"))

df <- cbind(report_rest,report_count,meta)

df$condition
```
## OFT offspring TIP merge
```{r}


y <- 37.5
e <- ggplot(df, aes(x = df$condition, y = df$bodycentre.periphery.total.time, fill = df$condition))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75),

    )+

   ylab("Time [s]")+
   xlab("Group")+
   ggtitle("Offspring Time Periphery")+

   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+

   labs(fill= "Condition")+

 theme(
  axis.text.x = element_text(size = 8),
  text = element_text(size = 14),  # Change font size of all text

)
 
 
 print(e2)
 
    ggsave("plots_svg/FigS5C.Offspring_merge.TIP.svg")
ggsave("plots_pdf/FigS5C.Offspring_merge.TIP.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5C.Offspring_merge.TIP.png")


```

## OFT offspring Distance merge
```{r}


y <- 37.5
e <- ggplot(df, aes(x = df$condition, y = df$bodycentre.distance.moving, fill = df$condition))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75),

    )+

  # ylim(c(0,40))+
   ylab("Distance [cm]")+
   xlab("Group")+
   ggtitle("Offspring Distance Covered")+

   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+ #c( 
   labs(fill= "Condition")+

 theme(
  axis.text.x = element_text(size = 8),
  text = element_text(size = 14), 
)
 
 
 print(e2)
ggsave("plots_svg/FigS5C.Offspring_merge.Distance.svg")
ggsave("plots_pdf/FigS5C.Offspring_merge.Distance.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5C.Offspring_merge.Distance.png")
  

```


# Fig S5D Cluster analysis for male offspring of stressed fathers


## UMAP

```{r}
set.seed(123)
i <- 64

#read in US oject
US <- readRDS("objects/US_object_offspring.rds")

#subset to have only females from IVF cohort
US <- SplitUSData(us = US,omit = US$file_names[ US$meta$sex == "f" ])

US_f <- US

#read in US oject

US_f <- TwoGroupAnalysis(us = US_f,group = US_f$meta$Stress)
US_f <- TwoGroupAnalysis_Clusters(us = US_f,group = US_f$meta$Stress)

  # Create the plot
  x <- Plot2DEmbedding(
    US_f,
    reports = US$label_names,
    colorby = "Group",
    color = c("seashell3", "#ffa500ff", "seashell4", "tomato3"),
    plot.sem = TRUE, 
    seed = 64
  ) + ggtitle(paste0("Syllable Count CTRL-vs-IFS offspring males"))

  print(x)  
    
ggsave("plots_svg/FigS5D.UMAPmales.svg")
ggsave("plots_pdf/FigS5D.UMAPmales.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5D.UMAPmales.png")


```
## Permutation Test Control vs Stress DMSO




```{r}
set.seed(64)

#Histogram for CTRL vs IFS in DMSO group for male offspring

#read in US oject
US <- readRDS("objects/US_object_offspring.rds")

#subset to have only females from IVF cohort
US <- SplitUSData(us = US,omit = US$file_names[ US$meta$sex == "f" ])

US_f <- US

#read in US oject

US_f <- SplitUSData(us = US_f,omit = US$file_names[US$meta$Treatment == "PROTAC"])
US_f <- TwoGroupAnalysis(us = US_f,group = US_f$meta$Stress)
US_f <- TwoGroupAnalysis_Clusters(us = US_f,group = US_f$meta$Stress)



  x <- PlotClusterStats(US_f, labels = US_f$label_names)
  print(x)

ggsave("plots_svg/FigS5D.PermMalesIFSvsCtrlDMSO.svg")
ggsave("plots_pdf/FigS5D.PermMalesIFSvsCtrlDMSO.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5D.PermMalesIFSvsCtrlDMSO.png")



```

## Permutation Test Control vs Stress KH-103



```{r}
set.seed(64)

#Histogram for CTRL vs IFS in KH-103 group for male offspring

#read in US oject
US <- readRDS("objects/US_object_offspring.rds")

#subset to have only females from IVF cohort
US <- SplitUSData(us = US,omit = US$file_names[ US$meta$sex == "f" ])

US_f <- US

#read in US oject

US_f <- SplitUSData(us = US_f,omit = US$file_names[US$meta$Treatment == "DMSO"])
US_f <- TwoGroupAnalysis(us = US_f,group = US_f$meta$Stress)
US_f <- TwoGroupAnalysis_Clusters(us = US_f,group = US_f$meta$Stress)



  x <- PlotClusterStats(US_f, labels = US_f$label_names)
  print(x)

ggsave("plots_svg/FigS5D.PermMalesIFSvsCtrlPROTAC.svg")
ggsave("plots_pdf/FigS5D.PermMalesIFSvsCtrlPROTAC.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5D.PermMalesIFSvsCtrlPROTAC.png")



```



# Fig S5E Cluster analysis for male and female offspring of stressed fathers


## UMAP
```{r}
set.seed(123)
i <- 64

#read in US oject
US <- readRDS("objects/US_object_offspring.rds")

US_f <- US

#read in US oject

US_f <- TwoGroupAnalysis(us = US_f,group = US_f$meta$Stress)
US_f <- TwoGroupAnalysis_Clusters(us = US_f,group = US_f$meta$Stress)

  # Create the plot
  x <- Plot2DEmbedding(
    US_f,
    reports = US$label_names,
    colorby = "Group",
    color = c("seashell3", "#ffa500ff", "seashell4", "tomato3"),
    plot.sem = TRUE, 
    seed = 64
  ) + ggtitle(paste0("Syllable Count CTRL-vs-IFS offspring"))

  print(x)  
    
ggsave("plots_svg/FigS5E.UMAPfemales+males.svg")
ggsave("plots_pdf/FigS5E.UMAPfemales+males.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5E.UMAPfemales+males.png")


```

## Permutation Test Control vs Stress DMSO
```{r}
set.seed(64)

#Histogram for CTRL vs IFS in DMSO group for  offspring

#read in US oject
US <- readRDS("objects/US_object_offspring.rds")



US_f <- US

#read in US oject

US_f <- SplitUSData(us = US_f,omit = US$file_names[US$meta$Treatment == "PROTAC"])
US_f <- TwoGroupAnalysis(us = US_f,group = US_f$meta$Stress)
US_f <- TwoGroupAnalysis_Clusters(us = US_f,group = US_f$meta$Stress)



  x <- PlotClusterStats(US_f, labels = US_f$label_names)
  print(x)

ggsave("plots_svg/FigS5E.PermIFSvsCtrlDMSO.svg")
ggsave("plots_pdf/FigS5E.PermIFSvsCtrlDMSO.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5E.PermIFSvsCtrlDMSO.png")



```

## Permutation Test Control vs Stress KH-103



```{r}
set.seed(64)

#Histogram for CTRL vs IFS in KH-103 group for male offspring

#read in US oject
US <- readRDS("objects/US_object_offspring.rds")


US_f <- US

#read in US oject

US_f <- SplitUSData(us = US_f,omit = US$file_names[US$meta$Treatment == "DMSO"])
US_f <- TwoGroupAnalysis(us = US_f,group = US_f$meta$Stress)
US_f <- TwoGroupAnalysis_Clusters(us = US_f,group = US_f$meta$Stress)



  x <- PlotClusterStats(US_f, labels = US_f$label_names)
  print(x)

ggsave("plots_svg/FigS5E.PermIFSvsCtrlPROTAC.svg")
ggsave("plots_pdf/FigS5E.PermIFSvsCtrlPROTAC.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5E.PermIFSvsCtrlPROTAC.png")



```
# Fig S5F SPT no-split
```{r}
spt_df <- data.frame(Sex = c(rep("f",13),rep("m",17)),Treatment = c(rep("DMSO",4),rep("PROTAC",2),rep("DMSO",4),rep("PROTAC",3), rep("DMSO",4),rep("PROTAC",4),rep("DMSO",5),rep("PROTAC",4)), Condtion = c(rep("CTRL",6),rep("Stress",7),rep("CTRL",8),rep("Stress",9)),
           SPT = c(83.21917808,	83.60299445,	83.60299445,	78.0042592, 83.56718834,	84.69643753, 74.65865647,	74.45669291,	76.78244973,	78.59216255,85.25290905,	89.54724409,	78.77526754,87.76315789,	88.15100579,	83.04337172,	82.6875, 83.72475341,	85.91406908,	86.40839041,	79.18757711, 85.33941814,	70.62055591,	78.34249953,	80.44692737,	79.90169306,		75.66512082,	82.14158026,	80.68035944,	76.53459821))
```

```{r}
spt_df$Group <- interaction(spt_df$Treatment,spt_df$Condtion)
```

```{r}
spt_df$Group <- factor(spt_df$Group, levels = c("DMSO.CTRL", "DMSO.Stress", "PROTAC.CTRL","PROTAC.Stress"))
```
```{r}
## done with wilcox test
 ## supported rears
# Male symbol
male <- "\u2642"
print(paste("Male:", male))

# Female symbol
female <- "\u2640"
print(paste("Female:", female))


y <- 37.5
e <- ggplot(spt_df, aes(x = Group, y = SPT, fill = Group))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("Overall Saccarine consume [%]")+
   xlab("Group")+
   ggtitle("Saccharine Preference Test")+
   ylim(c(70,100))+

   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+

   labs(fill= "")+
   scale_x_discrete(labels=c("f"=paste(female), "m"=paste( male)))+

 theme(
  axis.text.x = element_text(size = 10),
  text = element_text(size = 14),  
)
 print(e2)
 

 
 ggsave("plots_png/FigS5F.SPT.png")
  ggsave("plots_pdf/FigS5F.SPT.pdf")
   ggsave("plots_svg/FigS5F.SPT.svg")


```


# Fig S5G Behavior of offspring mice male + female LDB
```{r}
report <- read.csv("objects/LDB_Report_offspring.csv",sep = ",")
rownames(report) <- report$file
table(is.na(report))

# we convert NAs to zero
report[is.na(report)] <- 0

meta <- as.data.frame(readxl::read_excel("objects/LDB_metadata_offspring.xlsx"))
meta$OFTfile <- gsub(x = meta$OFTfile,pattern = ".csv(.*)$",replacement = ".csv")
row.names(meta) <- meta$OFTfile

# we order meta file according to report and check if TRUE
meta <- meta[rownames(report),]
table(rownames(report) == rownames(meta))

# we separate countdata from continues data --> important for downstream analysis



# we separate countdata from continues data --> important for downstream analysis

report_count <- report[,grep(x = colnames(report),pattern = "count|trans")]
# we remove non interesting or redundant behaviors
report_count <- report_count[,c(-2,-4,-5)]

report_rest <-  report[,-grep(x = colnames(report),pattern = "file|count|trans|percentage|raw")]

# we remove non interesting or redundant behaviors
report_rest <- report_rest[,c(-2,-which(colnames(report_rest) == "bodycentre.total.time"))]
```

```{r}

meta$condition <- gsub(x = meta$condition,pattern = " ",replacement = ".")
# se_count$Group <- relevel(se_count$Group,c("non.stressed.DMSO.f"))
meta$condition <- factor(meta$condition, levels = c("non.stressed.DMSO", "stressed.DMSO", "non.stressed.PROTAC","stressed.PROTAC"))

df <- cbind(report_rest,report_count,meta)

df$condition
```

## Time in Dark


```{r}

e <- ggplot(df, aes(x = df$condition, y = df$bodycentre.dark.total.time, fill = df$condition))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("Time [s]")+
   xlab("Group")+
   ggtitle("Offspring Time in Dark")+

   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+
   labs(fill= "Condition")+

 theme(
  axis.text.x = element_text(size = 8),
  text = element_text(size = 14), 
)
 
 
 print(e2)
ggsave("plots_svg/FigS5G.Offspring_merge.TimeInDark.svg")
ggsave("plots_pdf/FigS5G.Offspring_merge.TimeInDark.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5G.Offspring_merge.TimeInDark.png")
  

```

## Light to Dark transitions
```{r}

e <- ggplot(df, aes(x = df$condition, y = df$bodycentre..transitions.light.to.dark, fill = df$condition))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("Counts")+
   xlab("Group")+
   ggtitle("Offspring Light to Dark transitions")+

   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+
   labs(fill= "Condition")+

 theme(
  axis.text.x = element_text(size = 8),
  text = element_text(size = 14),  
)
 
 
 print(e2)
 
ggsave("plots_svg/FigS5G.Offspring_merge.LightToDark.svg")
ggsave("plots_pdf/FigS5G.Offspring_merge.LightToDark.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5G.Offspring_merge.LightToDark.png")

```
## Unsupported Rears

```{r}


e <- ggplot(df, aes(x = df$condition, y = df$classifications.UnsupportedRear.count, fill = df$condition))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("Count")+
   xlab("Group")+
   ggtitle("Offspring Unsupported rears")+

   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+
   labs(fill= "Condition")+

 theme(
  axis.text.x = element_text(size = 8),
  text = element_text(size = 14),
)
 
 
 print(e2)
ggsave("plots_svg/FigS5G.Offspring_merge.LDBunrear.svg")
ggsave("plots_pdf/FigS5G.Offspring_merge.LDBunrear.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5G.Offspring_merge.LDBunrear.png")

  

```

# Fig S5H Behavior of offspring mice male + female LDB separated by sex
## Time in Dark
```{r}

# Male symbol
male <- "\u2642"
print(paste("Male:", male))

# Female symbol
female <- "\u2640"
print(paste("Female:", female))


y <- 37.5
e <- ggplot(df, aes(x = df$sex, y = df$bodycentre.dark.total.time, fill = df$condition))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("Time [s]")+
   xlab("Sex")+
   ggtitle("Offspring Time in Dark")+

   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+ # c( "seashell2","#ffa500ff","seashell4","tomato3")
   labs(fill= "Condition")+
   scale_x_discrete(labels=c("f"=paste(female), "m"=paste( male)))+

 theme(
  axis.text.x = element_text(size = 25),
  text = element_text(size = 14),  
)
 
 
 print(e2)
ggsave("plots_svg/FigS5H.Offspring.TimeInDark.svg")
ggsave("plots_pdf/FigS5H.Offspring.TimeInDark.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5H.Offspring.TimeInDark.png")
  

```
## Light to Dark transitions

```{r}

# Male symbol
male <- "\u2642"
print(paste("Male:", male))

# Female symbol
female <- "\u2640"
print(paste("Female:", female))


y <- 37.5
e <- ggplot(df, aes(x = df$sex, y = df$bodycentre..transitions.light.to.dark, fill = df$condition))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("Count")+
   xlab("Sex")+
   ggtitle("Offspring Light to Dark transitions")+

   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+ # c( "seashell2","#ffa500ff","seashell4","tomato3")
   labs(fill= "Condition")+
   scale_x_discrete(labels=c("f"=paste(female), "m"=paste( male)))+

 theme(
  axis.text.x = element_text(size = 25),
  text = element_text(size = 14), 
)
 
 
 print(e2)
 
ggsave("plots_svg/FigS5H.Offspring.LightToDark.svg")
ggsave("plots_pdf/FigS5H.Offspring.LightToDark.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5H.Offspring.LightToDark.png")

```

## Unsupported Rears

```{r}
# Male symbol
male <- "\u2642"
print(paste("Male:", male))

# Female symbol
female <- "\u2640"
print(paste("Female:", female))


y <- 37.5
e <- ggplot(df, aes(x = df$sex, y = df$classifications.UnsupportedRear.count, fill = df$condition))


 e2 <- e + geom_boxplot(width = 0.75,outlier.shape = NA ) + 
   geom_jitter(
   position=position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)
    )+

   ylab("Count")+
   xlab("Sex")+
   ggtitle("Offspring Unsupported rears")+

   scale_fill_manual(values=c( "seashell2","#ffa500ff","seashell4","tomato3"))+ # c( "seashell2","#ffa500ff","seashell4","tomato3")
   labs(fill= "Condition")+
   scale_x_discrete(labels=c("f"=paste(female), "m"=paste( male)))+

 theme(
  axis.text.x = element_text(size = 25),
  text = element_text(size = 14), 
)
 
 
 print(e2)
ggsave("plots_svg/FigS5H.Offspring_sep.LDBunrear.svg")
ggsave("plots_pdf/FigS5H.Offspring_sep.LDBunrear.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5H.Offspring_sep.LDBunrear.png")
  

```

# Fig S5I Heatmap of differential expressed genes in 2-cell embryos by IFS fathers
```{r}
se <- log2FC(se, fromAssay = "corrected",controls = se$condition == "CTRL",isLog = F,by = se$treatment)#
```
```{r}
color_palette <- colorRampPalette(c("blue", "black", "orange"))(101)
rownames(se) <- gsub(x = rownames(se),pattern = "^ENS(.*)\\.",replacement = "")



sh <- sechm(se,features = Results$Total_IFSvsCTRL$ID[Results$Total_IFSvsCTRL$FDR <0.05],assayName = "scaledLFC",hmcols = color_palette,do.scale =F,gaps_at = c("treatment"),column_order = colnames(se)[order(colData(se)$condition)],show_rownames = F,name = paste0("StressGenes","\nlog2FC"),top_annotation = c("Group"),use_raster = T,column_title = "Stress effect in 2-cell embryos\nupon parternal stress and sperm treatment",column_title_gp = gpar(fontsize = 20)) 
print(sh)



pdf("plots_pdf/FigS5I.StressGenes.heatmap.pdf", width=14.5, height=10) #, width=14.5, height=10)

print(sh)
dev.off()

png("plots_png/FigS5I.StressGenes.heatmap.png", height=516,width = 758,)

print(sh)
dev.off()

svg("plots_svg/FigS5I.StressGenes.heatmap.svg") #, width=14.5, height=10)

print(sh)
dev.off()

```

# Fig S5J GO:BP 2-cell embryos of paternal stress



```{r}
se <- readRDS("objects/IFS_smartSeq_embryo_salmon.SE.rds")
meta <- openxlsx::read.xlsx("objects/IFS_smartSeq_embryo_meta.xlsx",colNames = T) 
meta <- meta[order(meta$Sample),]
meta <- meta[meta$Sample %in% colnames(se),]
table(meta$Sample == colnames(se))
se$embryos <- as.factor(meta$embryos)

Group <- factor(paste(se$treatment,se$condition,sep="."))


se$Group <- Group

set.seed(123)
se$Group2 <- sample(Group,size = 17,replace = F)
colData(se)
meta$Sample
```

```{r}
set.seed(123)
se <- SEtools::svacor(se, ~Group+embryos,form0 = ~embryos, n.sv = 3)

```

```{r}
dds <- calcNormFactors(DGEList(assay(se))) # this calculates Normalization factors according to library sequencing depth, this is important since depth can vary greatly between samples

### lets look at the library size
dds$samples



assays(se)$logcpm <- log1p(edgeR::cpm(dds)) # generates the log of Counts-per-million (CPM), this is a very standardized way to normalize data used then for analysis or visualisation




mm <- model.matrix(~0+Group+SV1+SV2+embryos, data=as.data.frame(colData(se))); mm


dds <- dds[filterByExpr(dds, group = Group, min.count=20),] # 20 maybe to stringent for embryo data
dds <- estimateDisp(dds,design = mm)
se <- se[row.names(dds),]

fit <- glmFit(dds, mm, prior.count = 5) # we us prior.count 5 since we are working with low count data
```
Creating contrasts
```{r}
fit$design

my.contrast <- makeContrasts(
  DMSO_IFSvsCTRL = GroupDMSO.IFS - GroupDMSO.CTRL ,
  PROTAC_IFSvsCTRL = GroupPROTAC.IFS - GroupPROTAC.CTRL,
  IFS_PROTACvsDMSO = GroupPROTAC.IFS - GroupDMSO.IFS,
  CTRL_PROTACvsDMSO = GroupPROTAC.CTRL - GroupDMSO.CTRL,
  IFS_PROTAC_int = ((GroupPROTAC.IFS - GroupPROTAC.CTRL) - (GroupDMSO.IFS - GroupDMSO.CTRL)), # interaction term

  Total_IFSvsCTRL = (GroupDMSO.IFS - GroupDMSO.CTRL + GroupPROTAC.IFS - GroupPROTAC.CTRL)/2,
  Total_PROTACvsDMSO =  (GroupPROTAC.IFS - GroupDMSO.IFS + GroupPROTAC.CTRL - GroupDMSO.CTRL)/2,
  levels = mm
)
my.contrast  

```


```{r}
Results <- list()
 
for (con in colnames(my.contrast)){
  Results[[con]] <- as.data.frame(topTags(glmLRT(fit,contrast = my.contrast[,con]),Inf))
  Results[[con]]$ID <- gsub(x = rownames(Results[[con]]),pattern = "ENS(.*)\\.",replacement = "")
  Results[[con]]$DEA <- con
}
Results

```


```{r}
set.seed(312)
Results_geo_uni <- list()
 
for (i in 1:length(colnames(my.contrast)[1])){
  dea <- Results[[i]]  
  gene_list <- dea[dea$FDR <= 0.05,]$ID # this get us all significantly differential expressed genes
kegg <- enrichGO(gene          = gene_list,
                OrgDb         = org.Mm.eg.db,
                keyType       = "SYMBOL",
                ont           = "BP",  # BP: Biological Process, MF: Molecular Function, CC: Cellular Component
                pAdjustMethod = "BH",
                pvalueCutoff  = 1,
                qvalueCutoff  = 1,
                 universe = dea$ID
                    )

  
  
  
  d <- dotplot(kegg, showCategory=10)

}
print(d)
ggsave("plots_svg/FigS5J.2cellGOBPDMSOvsIFS.svg")
ggsave("plots_pdf/FigS5J.2cellGOBPDMSOvsIFS.pdf")#, width = 6, height = 4)
ggsave("plots_png/FigS5J.2cellGOBPDMSOvsIFS.png")

```

# Fig S5G GO:BP in excitatory neurons (of a human neural organoid) acutely treated with Dexamethason
```{r}
se <- readRDS("objects/Binder_DexOrganoids_ExInh_Neurons.SE.rds")
gene_lists <- metadata(se)$feature.lists
```

```{r}

deas <- as.list(rowData(se)[,grepl(x = names(as.list(rowData(se))),pattern = "FOK4_Ex")])
deas <- lapply(deas,function(x){
  na.omit(x)
})
```


```{r}
library(clusterProfiler)


 

  dea <- deas[[1]]  
  gene_list <- dea[dea$FDR <= 0.05,]$gene # this get us all significantly differential expressed genes
kegg <- enrichGO(gene          = gene_list,
                OrgDb         = org.Hs.eg.db,
                keyType       = "SYMBOL",
                ont           = "BP",  # BP: Biological Process, MF: Molecular Function, CC: Cellular Component
                pAdjustMethod = "BH",
                pvalueCutoff  = 1,
                qvalueCutoff  = 1,
                universe = dea$gene 
                    )


  

d <- dotplot(kegg, showCategory=10)+ggtitle("GO:BP Ex.Neurons Acute Dex Effect")
print(d)


ggsave(paste0("plots_png//FOK4.Ex.Neurons.GOBP.Acute90d.png"))

ggsave(paste0("plots_pdf/FOK4.Ex.Neurons.GOBP.Acute90d.pdf"))

ggsave(paste0("plots_svg/FOK4.Ex.Neurons.GOBP.Acute90d.svg"))




  
```

```{r}
sessionInfo()
```



